---
title: "Differentials"
author: "Daniel Kaplan"
runtime: shiny_prerendered
---

```{r include=FALSE}
library(etude2)
```

The notation we have been using in this course for "the derivative of $f()$ with respect to $x$ is $$\partial_x f$$ The subscript on the $\partial$ indicates the "with respect to" variable, the curly-d $\partial$ itself signals "the derivative." The identification of the "with respect to" variable is particularly important when the function $f()$ has more than one input, for example with $f(x, t) \equiv e^{-k x} \sin(\omega t)$.

With anti-differentiation (also called "integration"), we write things differently, for example $$\int f(x, t) dx$$ where the $dx$ signals the "with respect to" variable. And just as $\partial_x f(x, t)$ and $\partial_t f(x,t)$ are different functions, so are $\int f(x, t) dx$ and $\int f(x, t) dt$. Writing anti-differentiation/integration this way follows a convention of long standing.

In interpreting a sequence of symbols like $\int f(x,t) dx$, it's natural to think of the sequence as three parts:

$$\underbrace{\int}_{\mbox{integral sign}} \overbrace{f(x, t)}^{\mbox{function to be anti-differentiated}} \underbrace{dx}_{\mbox{identifies 'with respect to' variable}}$$

By analogy, the sentence 

$$\mbox{We loaded up on snacks.}$$ 

consists of five parts: the five words in the sentence.

But you can also see "We loaded up on snacks" as having *three* parts:

$$\underbrace{\mbox{We}}_{\mbox{subject}}\  
\overbrace{\mbox{loaded up on}}^{\mbox{verb}}\ \ \ 
\underbrace{\mbox{snacks}}_{\mbox{object}}$$

Likewise, the integrate sentence can be seen as consisting of just two parts:

$$\underbrace{\int}_{\mbox{integral sign}} \overbrace{f(x, t) dx}^{\mbox{differential}}$$

A differential corresponds to the little rectangles that we add up when calculuting a definite integral numerically. That is
$$\underbrace{\int}_{\mbox{Sum}} \underbrace{\overbrace{f(x,t)}^\mbox{height} \overbrace{dx}^\mbox{width}}_\mbox{area of box}$$

A differential is a genuine mathematical object and is used, for example, in analyzing the geometry of curved spaces, as in the Theory of General Relativity. But this is well beyond the scope of this introductory calculus course.

There is one place in introductory calculus where you need to manipulate differentials: carrying out anti-differentiation of some functions that are not basic modeling functions with a linear interior function but which nonetheless have an anti-derivative that can be expressed in algebra. 

You should be thinking in terms of differentials when you see a sentence like the following:

> "In $\int \sin(x) \cos(x) dx$, make the substitution $u = \sin(x)$, implying that $du = \cos(x) dx$ and getting $\int u du$, which is simple to integrate."

The table gives some examples of functions and their differentials. "w.r.t" means "with respect to."

Function | derivative | w.r.t. |differential
---------|------------|--------|-------------
$v(x) \equiv x$      |    1       | x |   $dv = dx$
$u(x) \equiv x^2$    |   $2x$     | x | $du = 2x dx$
$f(x) \equiv \sin(x)$ | $\cos(x)$ | x | $df = \cos(x)dx$
$u(x) \equiv e^{3 x}$ | $3 e^{3 x}$ | x | $du = 3 e^{3 x} dx$ 
$g(x) \equiv t^3$     | $3 t^2$     | t | $dg = 3 t^2 dt$

As you can see, the *differential* of a function is simply the derivative of that function followed by the little $dx$ or $dt$ or whatever is appropriate for the "with respect to" variable.

Note: Tradition is to write something like $u \equiv e^{3 x}$ rather than $u(x) \equiv e^{3 x}$. The traditional notation is what we'll use in the following exercises.

```{r diff1, echo=FALSE}
etude2::etudeQ(
  "What is the differential of $u = x + 5$?",
  "+$du = dx$+" = "Since $\\partial_x (x+5) = 1$.",
  "$du = (x+5)dx$" = "You should have used $\\partial_x (x+5)$ instead of $(x+5)$ on the right side.",
  "$du = 5 dx$" = "Is $\\partial_x (x+5) = 5$? No!",
  "$du = x dx$" = "You should have used $\\partial_x (x+5) = 1$ instead of $x$ on the right side."
)
```

```{r diff2, echo=FALSE}
etude2::etudeQ(
  "What is the differential of $u = \\sin(2x + 5)$?",
  "+$du = 2 \\cos(2x + 5) dx$+" = "Since $\\partial_x \\sin(2x + 5) = 2 \\cos(2x + 5)$.",
  "$du = (2x+5)dx$" = "You should have used $\\partial_x \\sin(2x+5)$ instead of $(2x+5)$ on the right side.",
  "$du = 5 dx$" = "Is $\\partial_x \\sin(2x+5) = 5$? No!",
  "$du = 2x dx$" = "You should have used $\\partial_x \\sin(2x+5) = 2 \\cos(2x + 5)$ instead of $2x$ on the right side."
)
```

```{r diff3, echo=FALSE}
etude2::etudeQ(
  "What is the differential of $v = e^x$?",
  "$du = e^x dx$" = "The name of the function here is $v()$, not $u$. So the differential should be written $dv$.",
  "+$dv = e^x dx$+" = "Since $\\partial_x e^x = 1$.",
  "$du = x dx$" = "The name of the function here is $v()$, not $u$. So the differential should be written $dv$.",
  "$dv = x dx$" = "But $\\partial e^x = e^x$, not $x$."
)
```

```{r diff4, echo=FALSE}
etude2::etudeQ(
  "What is the differential of $f = \\cos(\\ln(t))$?",
  "+$df = -\\sin(\\ln(t))/t dt$+" = "Since the chain rule tells us $\\partial_t\\cos(\\ln(t)) = -\\sin(\\ln(t))/x$.",
    "$du = -\\sin(\\ln(t))/t dt$" = "There's no $u$ in sight in this problem. It's $f$.",
    "$dv = -\\sin(\\ln(t))/t dt$" = "There's no $v$ being used in this problem. It's $f$ that we're asking about.",
    "$df = -\\sin(\\ln(x))/x dx$" = "$f$ is a function of $t$, not $x$."
)
```
