---
title: "U substitution"
author: "Daniel Kaplan"
---

Differentiation and anti-differentiation are operations that undo one another. For instance, $$F(t) = \int \left[\partial_t F(t)\right] dt + C$$ and $$f(t) = \partial_t \int f(t) dt$$
We use differentiation and integration when we have information in one form (for instance a measurement of current in a circuit) and we find it more usefult to translate into another form (for instance the voltage across an element of the circuit). 

There is little reason to use $\partial_t$ and $\int \left[\right]dt$ to cancel each other out, but it is the basis of a successful strategy for finding some anti-derivatives algebraically. Here's the differentiate/integrate algorithm behind u-substitution. 

1. Pick a function $f()$ and another function $g()$. Typically $f()$ and $g()$ belong to the family of basic modeling functions, e.g. $e^x$, $\sin(t)$, $x^n$, $\ln(x)$, and so on. For the purpose of illustration, we'll use $f(x) = \ln(x)$ and $g(t) = \cos(t)$.
2. Compose $f()$ with $g()$ to produce a new function $f(g())$ which, in our case, will be $\ln(\cos(t))$.
3. Use the chain rule to find $\partial_t f(g(t))$. In the example, the derivative of $\ln(x)$ is $1/x$, the derivative of $g(t)$ is $-\sin(t)$. By the chain rule, $$\partial_t f(g(t)) = - \frac{1}{g(t)} \sin(t)= -  \frac{\sin(t)}{\cos(t)} = - \tan(t)$$ 

In a sense, we have just watched a function give birth to another through the straightforward process of differentiation. Having witnessed the birth, we know who is the integration mother of $\tan(t)$, namely $\int \tan(t) dt = \ln(\cos(t)$. For future reference, we might write this down in our diary of integrals:
$$\int \tan(t) dt = - \ln(\cos(t)) + C$$
Saving this fact in your diary is helpful. The next time you need to find $\int \tan(x) dx$, you can look up the answer ($-\ln(\cos(x)) + C$) from your diary. If you use $\int \tan(x) dx$ a lot, you will probably come to memorize the answer, just as you have already memorized that $\int \cos(t) dt = \sin(t)$ (a fact that you actually will use a lot in the rest of this course).

Now for the u-substitution game. The trick is to take a problem of the form $\int h(t) dt$ and extract from $h(t)$ two functions, an $f()$ and a $g()$. You're going to do this so that $h(t) =  \partial_t F(g(t))$, where $\partial_x F(x) = f(x)$ Once you've done this, you have an answer to the original integration question: $\int h(t) dt = F(g(t)) + C$.

Here's Example 7 from the MMAC book:

> Evaluate the definite integral $\int \frac{\sin(\ln(x))}{x} dx$.

You don't know ahead of time that this is an integral amenable to solution by u-substitution. For all you know, it's not. So you look at the instruction manual for calculating integrals that you don't yet know:

1. Is the integrand (that is $h(x)$ in $\int h(x) dx$) one of the basic modeling functions (with, perhaps, a linear interior function)? 

> If so, you've already memorized the answer and you are done. If not ...

2. Assume for a moment---without any guarantee that this will work, mind you---that the answer can be built using u-substitution. You will therefore look hard at $h()$ and try to see in it a plausible form that looks like the derivative of some $f(g(x))$. 

In the problem at hand, we can readily see something of the form $f(g(x))$ in the $\sin(\ln(x))$. This immediately gives you a candidate for $g(x)$, namely, $g(x)\equiv \ln(x)$. We don't know $f()$ yet, but if $g()$ is the right guess, and if u-substitution is going to work, we know that $f()$ has to be something that produces $\sin()$ when you differentiate it. That's $-\cos()$. So now we have a guess $$h_\mbox{guess}(x) = -\cos(\ln(x)) \partial_x \ln(x) = - \cos(\ln(x)) / x$$

> If this guess matches the actual $h()$ then you win. The answer to $\int h(x) dx$ will be $f(g(x)) = -\cos(\ln(x))$. If not, see if there is any other plausible guess for $g(x)$ to try. If you can't find one that works, go on to 

3. Integration by parts. We'll talk about this later. 

> If integration by parts doesn't work ... there is a variety of possibilities such as asking a math professor (who has a much larger set of functions at hand than you), looking through a table of integrals (which is to say, the collective calculus diary of generations of math professors), using a computer algebra system, or using numerical integration. One of these will work.

Back to u-substitution.

```{r usub1, echo=FALSE}
askMC(
  "In $h(x) = 2x/(x^2 + 2)$ which of the following is a plausible candidate for an interior function $g(x)$?",
  "$\\sin(x)$",
  "$\\ln(x)$",
  "$2x$" = "In order for this to be an interior function $g(x)$ there has to be some other function embracing it. $2x$ is standing proudly on its own, so it's not a good choice for $g(x)$",
  "+$x^2 + 2$+"
)
```

```{r usub2, echo=FALSE}
askMC(
  "Continuing with the integral of $h(x) = 2x/(x^2 + 2)$ and the working guess that $g(x) = x^2 + 2$, do you see any part of $h()$ which is a match to $\\partial_x g()$?",
  "$1/x$" = "But that's not anything like $\\partial_x g(x)$.",
  "$\\ln(x)$",
  "+$2x$+"
)
```

```{r usub3, echo=FALSE}
askMC(
  "Taking seriously the progress we made in the previous two questions, we now need to write  $h(x)$ as $f(x^2 + 2) 2x$? What should $f()$ be to make this match $h(x)$?",
  "$f(x) = \\sin(x)$",
  "$f(x) = \\ln(x)$",
  "+$f(x) = 1/x$+",
  "$x^2 + 2$"
)
```

Now that you have found both $g()$ and $f()$, you simply need to find a function $F(x)$ such that $\partial_x F(x) = f(x)$.  Since $\partial_x \ln(x) = 1/x$, we know that $F(x) = \ln(x)$. Thus,  $\int h(x) dx = F(g(x)) = F(x^2 + 2) = \ln(x^2 + 2)$.

```{r usub4, echo=FALSE}
askMC(
  "$$\\mbox{Find a plausible interior g(x) in} \\ x \\exp(x^2 + 3)$$",
  "$\\exp(x)$",
  "$x$", 
  "+$x^2 + 3$+",
  "$x^2$" = "The interior function has to be *everything* embraced by the parentheses of the exterior function $f()$."
)
```

```{r usub5, echo=FALSE}
askMC(
  "Using your candidate for $g()$ from the previous question, which of these is a *exterior* f(x) in $x \\exp(x^2 + 3)$",
  "+$f(x) = \\exp(x)$+",
  "$f(x) = x$", 
  "$f(x) = x \\exp(x)$" = "If this were so, the resulting $h()$ would have something like $(x^2 + 3) \\exp(x^2 + 3)$ in it.",
  "$f(x) = \\ln(x)$"
)
```

Confirm that $h(x) = f(g(x)) \partial_x g(x)$ and you win. The answer will be $F(g(x)) + C$.

```{r usb6, echo=FALSE}
askMC(
  "Which of these candidates for $f()$ and $g()$ will produce  $$f(g(x)) \\partial_x g(x) = x^3 \\cos(x^4)\\ ?$$",
  "+$f(x) = \\cos(x)/4$ and $g(x) = x^4$+" = "The 1/4 cancels out the 4 produced by $\\partial_x g(x)$.",
  "$f(x) = \\cos(x)$ and $g(x) = x^4$",
  "$f(x) = x^4$ and $g(x) = \\cos(x)$" = "But $h(x)$ involves $\\cos(x^4)$ not $cos^4(x)$."
)
```

Once again, $\int h(x) dx = F(g(x))$, where $\partial_x F(x) = f(x)$.

```{r usb7, echo=FALSE}
askMC(
  "What is $$\\int \\frac{\\sin(x)}{\\cos^5(x)}dx\\ ?$$",
  "$\\ln(cos(x))$",
  "+$- \\frac{1}{4} \\cos^{-4}(x)$+",
  "$\\frac{1}{6} \\cos^{-6}(x)$"
)
```

```{r usb8, echo=FALSE}
askMC(
  "Use u-substitution to find $$\\int \\frac{4 e^{4x} + 4}{e^{4x}}dx$$",
  "+$\\ln(e^{4x} + 4)$+",
  "$1/(e^{4x} + 4)$",
  "$\\frac{1}{4} e^{4x} + 4$",
  "$\\frac{1}{2} 1/(e^{4x} + 4)^2$"
)
```

