# Iteration and accumulation

Let's review the concept of an ***Euler step***.


**Problem statement**:

1. You seek to find a function, e.g. $f(t)$ which you do not yet know.
2. You already know $\partial_t f(t)$. (In the following chapters we'll show you some of the many instances of this unlikely-sounding situation, where you don't know $f(t)$ but you do know $\partial_t f(t)$.) This relationship, $\partial_t f(t) \longrightarrow f(t)$ is, of course, revealed by ***anti-differentiation***.

**Setup**:

3. You know or make up a value for $f(t)$ at a particular time of interest $t=t_0$. "Making up" is a perfectly legitimate practice here to get started in creating an an anti-derivative. The reason is hard to explain until you see the whole process of anti-differentiation by the ***Euler method***, so hang on. In the population modeling problem in Chapter \@ref(change-accumulation), we were interested in forecasting the population into the future, so we chose $t_0$ to be the most recent year for which we have information: $t_0 = 2020$. (In polite mathematical conversation, "making up" is pronounced "assuming.")
4. Consider a quantity $h$ which is to be regarded as a little bit of $t$: they have the same dimension. $h$ should be "small," but "small" compared to what? Often, the modeling problem at hand involves a domain of interest. For instance, the population modeling problem in Chapter \@ref(change-accumulation) involved a forecast over the time interval from year 2020 to 2100. Provisionally, take small to mean 1% or 0.1% of the extent of the domain of interest. (In the population modeling problem, we chose $h=1$ year.)

**The Step**

1. Compute the value of the (still unknown) function $f(t)$ at input $t_0 + h$ as
$$f(t_0+h) = f(t_0) + h \times \partial_t f(t)$$

Now you know a little bit more about $f(t)$, namely, its value at $t_0 + h$.

Since the original problem was to figure out $f(t)$ over some domain of interest (starting at $t=t_0$), we are not finished. We have taken but one step of our journey and have many more to go.

To continue on the journey, we take what we just found out, namely $f(t_0 + h)$. We use this as the starting point in a new step. We don't have to make anything up because we already know $f(t_0 + h)$.

The second step brings us to $f(t_0 + 2h)$

$$f(t_0+2h) = f(t_0+h) + h \times \partial_t f(t+h)$$
And the third:
$$f(t_0+3h) = f(t_0+2h) + h \times \partial_t f(t+2h)$$

How long to continue this step-by-step process? The original problem involved some domain of interest, and we selected $h$ as 1% or 0.1% of the extent of this domain. For $h$ as 1%, taking 100 steps will get us to the far end of the domain. For $h$ as 0.1%, we would need to take 1000 steps. 

```{exercise, name="euler-step"}
Exercises like this:
```
<details>
  
Using the Euler method find $\int f(t) dt$ over the interval $t_0=0$ to $t_{end}=1$. The $t$ quantity is in steps of $h=0.01$.

$t$  | $\partial_t f(t)$ | $\int f(t) dt$
-----|-------------------|-----------------
0    | `r round(dnorm(0)   , 3)`   |    0.5
0.01 | `r round(dnorm(0,01), 3)`   |  
0.02 | `r round(dnorm(0,02), 3)`   |  
0.03 | `r round(dnorm(0.03), 3)`   |  

</details>


## As a spreadsheet

## Using `cumsum()`

## Using `iterate()`


## Using a loop




## The accumulated quantity

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Int-2e", "Determine the units of an anti-derivative given the base function and the variable of integration.")
```
:::


## Graphical construction

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Int-5a", "Determine dimensions and units utilizing a graph.")
state_objective("Int-5b", "Graph the anti-derivative of a function given the graph of the base function and a single point on the anti-derivative.")
```
:::

## Step by step with Euler

::: {.objectives}
```{r echo=FALSE, results="markup"}
state_objective("Int-2a", "Utilize Euler’s Method to approximate anti-derivatives when the value of is small.")
```
:::




