--- 
title: "CalcZ Student Notes"
author: "Daniel Kaplan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Textbook for Math 141Z/142Z for 2021-2022"
---
--- 
title: "CalcZ Student Notes"
author: "Daniel Kaplan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Textbook for Math 141Z/142Z for 2021-2022"
---
```{r include=FALSE, cache=FALSE}
library(mosaic)
library(mosaicCalc)
library(glue)
library(math141Z) # REPLACE THIS WHEN PACKAGES ARE RE-ALIGNED
library(here) # for file locations



exercise_file <- function(day,  file_name, course="141Z") {
    path=glue::glue("Exercises/DD-{course}-{day}/{file_name}")
    here(path)
}

if (!exists("objective_list"))
  objective_list <- list()

add_objective <- function(ID, text) {
  objective_list[[length(objective_list) + 1]] <<-
                      list(ID = ID, text = text)
}

state_objective <- function(ID, text) {
  add_objective(ID, text)
  format_objective(list(ID=ID, text=text))
}

format_objective <- function(obj) {
  knitr::asis_output(glue::glue("#. **[{obj$ID}]** *{obj$text}*\n\n"))
}

show_objectives <- function() {
  Tmp <- lapply(objective_list, tibble::as_tibble) %>%
    bind_rows()
  readr::write_csv(Tmp, file="objective-list.csv")

  lapply(objective_list, format_objective) %>%
    unlist() %>%
    paste(collapse="\n") %>%
    knitr::asis_output()
}

# A blank image for spacing
BlankImage <- gf_blank(hp ~ wt, data=mtcars) %>% gf_theme(theme_void())
```

# Status {.unnumbered}

This is the working version for Summer 2021. 




<!--chapter:end:index.Rmd-->


# (PART) Orientation {.unnumbered}

Placeholder



<!--chapter:end:Orientation.Rmd-->


# (PART) Block 1: Functions and quantity {.unnumbered}

Placeholder



<!--chapter:end:Fun-what-is-a-function.Rmd-->


# Calculus: Functions and change

Placeholder


## Naked modeling functions
## Change and slope
## Continuous change
## Functions, graphs, and slopes
## Functions, functions, functions

<!--chapter:end:Fun-slopes.Rmd-->


# Notation for functions

Placeholder


## Mathematical formulas
## Computer notation

<!--chapter:end:Fun-notation.Rmd-->


# Unassigned exercises

Placeholder


## Dimension and magnitude
## Combining functions
## Hump & sigmoid functions
## Sinusoids
## Shapes of graphs
## Exponential functions
## Log scales
## Power-law functions

<!--chapter:end:Fun-exercises.Rmd-->

```{r include=FALSE, cache=FALSE}
library(mosaic)
library(mosaicCalc)
library(glue)
library(math141Z) # REPLACE THIS WHEN PACKAGES ARE RE-ALIGNED
library(here) # for file locations



exercise_file <- function(day,  file_name, course="141Z") {
    path=glue::glue("Exercises/DD-{course}-{day}/{file_name}")
    here(path)
}

if (!exists("objective_list"))
  objective_list <- list()

add_objective <- function(ID, text) {
  objective_list[[length(objective_list) + 1]] <<-
                      list(ID = ID, text = text)
}

state_objective <- function(ID, text) {
  add_objective(ID, text)
  format_objective(list(ID=ID, text=text))
}

format_objective <- function(obj) {
  knitr::asis_output(glue::glue("#. **[{obj$ID}]** *{obj$text}*\n\n"))
}

show_objectives <- function() {
  Tmp <- lapply(objective_list, tibble::as_tibble) %>%
    bind_rows()
  readr::write_csv(Tmp, file="objective-list.csv")

  lapply(objective_list, format_objective) %>%
    unlist() %>%
    paste(collapse="\n") %>%
    knitr::asis_output()
}

# A blank image for spacing
BlankImage <- gf_blank(hp ~ wt, data=mtcars) %>% gf_theme(theme_void())
```
# Unassigned functions about derivatives

```{exercise, name="SIR-dimensions"}
```
<details>`r knitr::knit_child("Exercises/Diff/SIR-dimensions.Rmd")`</details>


```{exercise, name="crow-mean-dress"}
```
<details>`r knitr::knit_child("Exercises/Diff/crow-mean-dress.Rmd")`</details>


<!--chapter:end:Diff-exercises.Rmd-->

