# (PART) Block 5: Dynamical systems {.unnumbered}

# States, Dynamics, Trajectory
    
::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-1a", "Understand the difference between state, dynamics, trajectory")
state_objective("Dyn-1b", "Understand the difference between continuous and discrete time")
state_objective("Dyn-1c", "Recognize the form of a differential or finite-difference equation")
```
:::

# Difference equations in 1 dimension

    Objectives:")
::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-2a", "Understand how discrete time operates and what a step is ")
state_objective("Dyn-2b", "Define a fixed point in discrete time and determine whether the fixed point is stable or unstable")
state_objective("Dyn-2c", "Understand the relationship between oscillations and stability as we step through discrete time")
```
:::

# Difference equations in 2 dimensions

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-3a", "Learn how to construct future states of a dynamical system in two dimensions given the dynamical rule and the two-dimensional initial condition.")
state_objective("Dyn-3b", "Appreciate that global dynamics can be complicated, potentially with many fixed points, learn how to identify the locations of the fixed points.")
state_objective("Dyn-3c", "Understand how local dynamics near a fixed point are comparatively simple and can be meaningfully approximated by a matrix.")
state_objective("Dyn-3d", "Review the mechanics of matrix arithmetic and iterate a dynamical system written in terms of a matrix.")
```
:::

# Wrap-up Discrete Time difference equations

  

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-4a", "Understand how eigenvalues can describe the properties of difference equations in discrete time")
state_objective("Dyn-4b", "Understand sequences as steps through discrete time")
```
:::

# Continuous-Time differential equation in 1 dimension


::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-5a", "Using slope fields, understand that a differential equation has a family of solutions")
state_objective("Dyn-5b", "Determine a particular solution to a differential equation in one dimension")
state_objective("Dyn-5c", "Determine end behavior of a differential equation graphically")
state_objective("Dyn-5d", "Describe a scenario using proportional or inversely proportional growth with a differential equation")
```
:::

# Continuous-Time differential equation in 2 dimensions


::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn6-lost", "Determine a particular solution to a differential equation in two dimensions")
state_objective("Dyn-6a", "Describe a scenario using logistics growth with a differential equation")
state_objective("Dyn-6b", "Understand the three types of dynamics stable, instable, and saddle points and how lambda values can clearly show which one is which")
```
:::

# Catch-up: Review Day

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-7a", "Understand the benefits of using the exponential function as an educated guess for a solution to a differential equation")
state_objective("Dyn-7b", "Comprehend how solutions to differential equations can have complex eigenvalues")
state_objective("Dyn-7c", "Understand how complex eigenvalues are related to oscillations")
```
:::

 Force-balance equation

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-8a", "Damped harmonic oscillator example")
state_objective("Dyn-8b", "Translate eigenvalues into the description of qualitative dynamics")
state_objective("Dyn-8c", "Understand eigenvectors and their connection to flow fields")
```
:::

Nonlinear phase plane and systems of nonlinear differential equations

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-9a", "Interpret phase plane diagrams, flow field diagrams, and time series plots of nonlinear systems of differential equations")
state_objective("Dyn-9b", "Be able to interpret individual terms in a dynamical function that describes specific relationships among real-world objects")
state_objective("Dyn-9c", "Predator-prey model")
state_objective("Dyn-9d", "Attrition model")
```
:::

Complex Eigenvalues

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Dyn-10a", "Understand the benefits of using the exponential function as an educated guess for a solution to a differential equation")
state_objective("Dyn-10b", "Comprehend how solutions to differential equations can have complex eigenvalues")
state_objective("Dyn-10c", "Understand how complex eigenvalues are related to oscillations")
```
:::
