# Slope function {#fun-slopes}

For our purposes, the ***definition of calculus*** is

> *The use of functions to model and explore continuous change*

The agenda of this chapter is to give specific mathematical meaning to the word "change."

## Change and slope

You have an solid, intuitive sense of what "change" means. In mathematics, and especially the mathematics of functions, change has a very simple meaning that you have already touched on in your previous math education.

The word that encapsulates "change" in high-school math is ***slope***. For instance, you've undoubtedly had to calculate the slope of a straight line in a graph. You learned about "rise" and "run" and how to read them from a graph or from a formula. The slope is the ratio: rise over run. 

Slope is a lovely ***metaphor*** for change, since everyone has a intuitive sense of the slope of a road or of a hillside. You learned to apply this intuition to reading graphs and the slope of a line. We'll exploit the intuitive ability to read a landscape in order to introduce abstract mathematical ideas in a down-to-earth setting. It's a very effective pedagogical strategy.

But not everything that changes has a "slope." For instance, the population of a country can change, as can the number of new cases of an epidemic disease, the temperature of a cup of coffee, or the distance from Earth of a spacecraft. A major part of learning calculus is generalizing and abstracting the mathematical concept of which "slope" is an example and becoming proficient with mathematical procedures for working with change. 

## Continuous change

Most people are comfortable with the ideas of daily changes in temperature or monthly changes in credit-card debt or quarterly changes in the unemployment rate or annual changes in the height of a child. Such things are easy to record in, say, a spreadsheet. For example, as I write, the weather forecast for the next several days (in southeastern Colorado in mid-May) is

Day | High | Low | Description
----|------|-----|------------
Thursday | 73 | 43 | sunny
Friday   | 72 | 48 | windy
Saturday | 66 | 48 | thunderstorms
Sunday   | 68 | 43 | windy
Monday   | 70 | 39 | sunny
Tuesday  | 70 | 43 | sunny
Wednesday| 66 | 45 | partly cloudy

Such data is said to be ***discrete***. The day is listed, but not the time of day. The high temperature is forecast, but not the time of that high. The "description" is also discrete, one of the several words that are used to summarize the quality of the weather, as opposed to the quantity of rain. 

Calculus is about ***continuous change***. For instance, if the weather bureau provide a web interface that let me enter the date and time to the nearest fraction of a second, they would be giving a way to track the change ***continuously***. Many physical processes are intrinsically continuous, for instance the motion (change in position) of a spacecraft or the height of the tide or the stress on a tree as a function of wind velocity.

Finding a language to describe continuous change---famously, the position of the moon or planets in their orbit, or the speed of a ball rolling down a ramp---was central to the emergence of what historians call the "Age of Enlightenment" or "modern scientific method." The first complete presentation of that language was published by Isaac Newton based on his work in the 1660s. As you might guess, the name of the language is "calculus."

## Slope

You already know pretty much everything there is to know about the straight-line function,

- **Formula**: $f(x) \equiv a x + b$. The parameters $a$ and $b$ are the "slope" and "intercept" respectively. (More precisely, $b$ is the "y-intercept." But in statistics and modeling, it's just the "intercept.")
- **Reading parameters from a graph**: You learned several ways to do this which are all equivalent. Maybe the easiest is the read the y-intercept off the graph. That's $b$. Then choose some non-zero $x_1$ and read off from the graph the value of $f(x_1)$. The slope is simply $$\frac{f(x_1) - b}{x_0}$$

    The ***y-intercept method*** is a special case of a more general method, the ***two-point method***, that you can use even if the y-intercept isn't shown on the graph. Pick two specific values of $x$, which we'll call $x_0$ and $x_1$. Evalate the function at these input values and compute the rise over run: $$\mbox{rise over run} \equiv \frac{f(x_1) - f(x_0)}{x_1 - x_0}$$
The rise over run is the slope of the straight line.

    The y-intercept method is exactly the same as the two-point method with $x_0 = 0$.
    
- **Matching a straight-line function to data**: You might not have been taught this formally, but the basic process is easy to imitate. The process is called ***line fitting*** or, in statistics and other fields, ***linear regression***.

## The fitted line

::: {.objectives}
```{r echo=FALSE, results="asis"}
state_objective("Fun-4-a-1", "Construct a straight-line model that fits data.")
```
:::



To illustrate line fitting, let's return to the cooling mug of water. Figure \@ref(fig:Fun-4-a) shows the data along with a dozen candidate straight line functions, each one drawn in its own color.

```{r Fun-4-a, echo=FALSE, warning=FALSE, fig.cap="Some candidate straight-line function models plotted on top of the cooling water data. Which one(s) would you pick as good matches to the data?"}
set.seed(101)
gf_point(temp ~ time, data=CoolingWater %>% sample_n(20)) %>%
  gf_labs(x = "Time (minutes)", y="Temperature (deg. C)") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20), color="blue") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20), color="aquamarine") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20), color="cadetblue") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20), color="green") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20), color="cadetblue") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20), color="royalblue") %>%
  gf_lm(temp ~ time, data = CoolingWater %>% sample_n(20),color="seagreen") %>%
  gf_abline(intercept = 100, slope=-0.21, color = "plum") %>%
  gf_abline(intercept = 90, slope=-0.19, color = "plum3") %>%
  gf_abline(intercept = 63, slope=-0.05, color = "purple") %>%
  gf_abline(intercept = 120, slope=-0.51, color = "purple3") %>%
  gf_abline(intercept = 7, slope=.21, color = "orchid")
```

Some of the straight-line models are a much better match to the data than others. The blue-shaded functions are pretty good fits, the greenish functions are maybe OK but a little sketchy, and the purple-shaded functions are just horrible.

Now that you know what a reasonable straight-line model looks like, you will find it pretty easy to draw one on data graphics that even remotely show a straight-line pattern. 

Step 1: Draw a reasonable straight-line through the data points. 

Step 2: Find the parameters that correspond to the line you drew. 

## Average rate of change

Since the slope is our standard way of representing a relationship of change, we will often use it as a way of summarizing a function. To illustrate, consider the exponential model we constructed to match the cooling-water data:

```{r}
water <- makeFun(60.7*exp(-0.019*t) + 25.93 ~ t)
```



```{r water-average, echo=FALSE, fig.cap="The exponential function that was previously matched to the cooling-water data. The slope of the straight line connecting two points on the function graph is the average rate of change during the interval."}

Pts <- tibble::tibble(
  x = c(10,125),
  y = water(x),
  baselabel = c("t0", "t1"),
  flabel = c("f(t0)", "f(t1)")
)
slice_plot(water(t) ~ t, domain(t = c(0,200))) %>%
  gf_point(y ~ x, data = Pts, color="red", alpha=0.25, size=4) %>%
  gf_line(y ~ x, data = Pts, color="red") %>%
  gf_point(0 ~ x, data = Pts, color="blue", alpha=0.25, size=4) %>%
  gf_lims(y=c(0,90)) %>%
  gf_segment(0 + y ~ x + x, data=Pts, color="blue", linetype=2) %>%
  gf_text(0 ~ x, data = Pts, label = ~baselabel, color="blue", nudge_x=7) %>%
  gf_text(y ~ x, data = Pts, label = ~flabel, color="red", nudge_y=7, nudge_x=2)
```

During the interval $[t_0, t_1]$ the rate at which the water cools is higher at first and lower at the end. The ***average*** rate of change is a single number that summarizes the whole interval.

For all except straight-line models, the average rate of change depends on the interval chosen.

EXERCISE: From the graph, compute the average rate of change over the interval $10 \leq t \leq 200$. How does it compare to the average rate of change over the interval $10 \leq t \leq 125$?

EXERCISE: Same, but for a sinusoid. Give a couple of intervals, including one where there is no net change.



## Instantaneous rate of change

The water is cooling continuously. Sometimes we'll be interested in the rate of change at a given instant $t_0$. If we choose a very small interval, say $[t_0, t_0 + 0.1]$, we'll get a good approximation to this ***instantaneous rate of change***.
$$\text{instantaneous rate of change} = \frac{f(t_0 + 0.1) - f(t_0)}{0.1}$$

On a graph, the instantaneous rate of change is the slope of the tangent line to a point, as shown in Figure \@ref(fig:tangent-line-exp).

```{r tangent-line-exp, echo=FALSE, fig.cap="Instantaneous rate of change is the slope of the line tangent to a the curve at a single point."}
Pt <- tibble::tibble(
  x = 25,
  y = water(x),
  baselabel="x0",
  flabel = "f(x0)"
)
tangent <- makeFun(-0.71722*(t-25) + 63.7 ~ t)
slice_plot(water(t) ~ t, domain(t=c(0, 200))) %>%
  gf_point(y ~ x, data=Pt, 
           color="sienna", alpha = 0.25, size=4) %>%
  gf_point(25 ~ x, data=Pt, 
           color="tomato", alpha = 0.25, size=4) %>%
  slice_plot(tangent(t) ~ t, color = "sienna", 
             domain(t=c(0,80))) %>%
  gf_text(y ~ x, data = Pt, label = ~ flabel, 
          nudge_x = 10, color="sienna") %>%
  gf_text(25 ~ x, data = Pt, label = ~ baselabel, 
          nudge_x = 10, color="tomato") %>%
  gf_segment(25 + y ~ x + x, data=Pt, color="tomato", linetype=2)
```

In Block 2, we'll see that a good way to define an instantaneous rate of change at $t_0$ is as the average rate of change over the interval $t_0 \leq t \leq t_1$ with the proviso that the interval length $t_1 - t_0$ goes as closely as it can to zero.

Figure \@ref(fig:three-rates) shows the fitted rate of change and average rate of change in the interval $10 \leq t \leq 125$ as well as the instantaneous rate of change at $t_0 = 50$. As you can see, they are different from one another.

```{r three-rates, echo=FALSE, warning=FALSE, fig.cap="Comparing the average rate of change (red) over the interval $10 \\leq t \\leq 125$, the instantaneous rate of change at $x_0 =25$, and the fitted linear model (orange). The exponential model is shown in gray."}
gf_point(temp ~ time, data = Stans_data, color="orange") %>%
  gf_lm(color="orange") %>%
  slice_plot(water(time) ~ time, domain(time=c(0, 200)), color="gray", size=2, alpha = 0.5) %>%
  slice_plot(tangent(time) ~ time, color = "sienna", 
             domain(time=c(0,80))) %>%
  gf_point(y ~ x, data = Pts, color="red", alpha=0.25, size=4) %>%
  gf_line(y ~ x, data = Pts, color="red") %>%
  gf_point(25 ~ x, data = Pts, color="blue", alpha=0.25, size=4) %>%
  gf_point(y ~ x, data=Pt, 
           color="sienna", alpha = 0.25, size=4) %>%
  gf_point(25 ~ x, data = Pt, color="tomato", alpha=0.25, size=4) %>%
  gf_segment(25 + y ~ x + x, data=Pt, color="tomato", linetype=2) %>%
  gf_segment(25 + y ~ x + x, data=Pts, color="blue", linetype=2) %>%
  gf_lims(y=c(25, NA))
```

## The slope function

We have a way to estimate the instantaneous rate of change of a function at a given input $t_0$:
$$\text{instantaneous_rate_of_change} \equiv \frac{f(t_0 + 0.1) - f(t_0)}{0.1}$$
Now we'll let notation do some work for us and re-write the above as the instantaneous-rate-of-change function, which we'll call the ***slope*** function.
$$\text{slope}(t) \equiv \frac{f(t + 0.1) - f(t)}{0.1}$$
We've merely made it clear that $\text{slope}(t)$ is a function and written the formula in terms of the input $t$ rather than a fixed number $t_0$. Of course, the "instantaneous rate of change at $t_0$" is simply $\text{slope}(t)$.

Let's look at the slope function that corresponds to some of our naked modeling functions: $e^x$, $\sin(x)$, $x^{-1}$ and $\ln(x)$. Recall that the input to the naked functions is always a number. The output is a number as well. Thus we can display the function and the corresponding slope function on the same coordinate frame. (This will not be true when the input is a quantity and the output is another quantity.)


```{r compare-naked-slopes, echo=FALSE, out.width="50%", fig.show="hold", fig.cap="Comparing the naked modeling function (blue) to it's slope function (red)"}
slice_plot(sin(x) ~ x, domain(x=c(-10,10)), color="blue") %>%
  slice_plot(cos(x) ~ x, color="red", alpha=0.25, size=2) %>%
  gf_labs(title="Sinusoid")
slice_plot(exp(x) ~ x, domain(x=c(-2,2)), color="blue") %>%
  slice_plot(exp(x) ~ x, color="red", alpha=0.25, size=2) %>%
  gf_labs(title="Exponential function")
slice_plot(log(x) ~ x, domain(x=c(0.1,5)), color="blue") %>%
  slice_plot(1/x ~ x, color="red", alpha=0.25, size=2) %>%
  gf_labs(title="Logarithm function")
slice_plot(1/x ~ x, domain(x=c(0.1,5)), color="blue") %>%
  slice_plot(-1/x^2 ~ x, color="red", alpha=0.25, size=2) %>%
  gf_labs(title="Power law function (with p=-1)")
```

EXERCISE: There is a web of connections between the naked modeling functions and their slopes. 

```{r slope-sin, echo=FALSE, results="markup"}
etude2::etudeQ(
  "1. Which naked modeling function has a slope function that is simply a input-shifted version of itself?",
  "exponential",
  "+sinusoid+",
  "logarithm",
  "power-law $x^{-1}$",
  random_answer_order = FALSE
)
```

```{r slope-exp, echo=FALSE, results="markup"}
etude2::etudeQ(
  "2. Which naked modeling function has a slope function that is identical to itself?",
  "+exponential+",
  "sinusoid",
  "logarithm",
  "power-law $x^{-1}$",
  random_answer_order = FALSE
)
```

```{r slope-log, echo=FALSE, results="markup"}
etude2::etudeQ(
  "3. Which naked modeling function has a slope function that is another naked modeling function? (Hint: The other function is also shown in the set of graphs.)",
  "exponential",
  "sinusoid",
  "+logarithm+",
  "power-law $x^{-1}$+",
  random_answer_order = FALSE
)
```


::: {.why}
We defined the "slope function" as
$$\text{slope}(t) \equiv \frac{f(t + 0.1) - f(t)}{0.1}$$

The 0.1 in the definition of the slope function is definitely a kluge.  0.1 minutes may be small when it comes to cooling water, but it might not be small in other contexts. We'll fix that in Block 2 when we define the ***derivative operator*** which replaces 0.1 with something that small in every context.
:::






