# (PART) Orientation {.unnumbered}

Welcome to CalcZ. Calculus is a mathematical topic of long standing. Some parts date to antiquity, some parts to the 17th century, with important topics added in the 19th and 20th centuries. You will be using calculus in the 21st century: applying it to current problems with current technology and data.

1. Exercises often have feedback for the right or wrong answer.
2. But not always. These are marked with an icon.
3. Sandboxes -- to be opened in another window. We often give "scaffolding" to get you started. Just copy and paste this into the sandbox.

# Themes {.unnumbered}

- functions
- computation
    - notation
    - universality: calculus is not just for the mathematical functions that can be written with a formula. Separates algebraic work from ability to use calculus.
    - iteration
        
        
- formula reading
- units and dimension


# Short response Exercises {.unnumbered}

```{r child=exercise_file("01", "exercise-intro.Rmd")}
```

# Computing {.unnumbered}

```{r child=exercise_file("01", "sandbox-intro.Rmd")}
```

````{exercise, NOAA, name="Tables are functions"}
```
`r knitr::knit_child(exercise_file("01", "NOAA.Rmd"))`


```{exercise, MMAC-01, name="Title of the exercise"}
```
`r knitr::knit_child(exercise_file("01", "MMAC-01.Rmd"))`

```{r child=exercise_file("01", "MMAC-01.Rmd")}
```

```{exercise}
```
`r knitr::knit_child(exercise_file("01", "MMAC-01-26.Rmd"))`


