---
id: "gradient-field-orange"
author: "Daniel Kaplan"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```


It's relatively easy to assess partial derivatives when you know the gradient. After all, the gradient is the vector of $(\partial_x\,f(x,y), \partial_y f(x,y))$. To train your eye, here's a contour plot and a corresponding gradient plot.

```{r echo=FALSE, fig.show="hold", warning=FALSE, out.width="50%"}
g <- rfun(~ x + y, seed = 93923)
dom <- domain(x=c(-4,4), y=c(-4,4))
dom2 <- domain(x=c(-3.5, 3.5), y=c(-3.5, 3.5))
contour_plot(g(x,y) ~ x + y, dom)
gradient_plot(g(x,y) ~ x + y, dom2, n = 10, alpha=1) %>%
  gf_lims(x=c(-4, 4), y=c(-4, 4))
```


```{r gradient-field-orange-QA4, echo=FALSE}
askMC(
  prompt = "What is the rule for determining $\\partial_x f(x,y)$ from the direction of the gradient vector?",
  choices = list(
    "+If the vector has a component pointing right, $\\partial_x f$ is positive.+" = "",
    "If the vector has a component pointing left, $\\partial_x f$ is positive" = "If the gradient points left, then uphill is to the left. So the function is increasing from right to left. That's a negative partial derivative.",
    "If the vector has a vertical component pointing up, $\\partial_x f$ is positive." = "This would be true for the partial derivative with respect to $y$, but that has to relevance to the partial with respect to $x$.",
    "If the vector has a component pointing downward, the partial derivative $\\partial_x f$ is positive." = "No, but the partial with respect to $y$ would be negative."
  ),
  inline=FALSE
)
```

Using the gradient field depicted below, figure out the sign of the partial derivatives at the labeled points. We'll use "neg" to refer to negative partial derivatives, "pos" to refer to positive partial derivatives, and "zero" to refer to partials that are so small that you can't visually distinguish them from zero.

```{r echo=FALSE}
set.seed(1114)
Points <- tibble(
  x = 2*round(runif(15, -2, 2)),
  y = 2*round(runif(15, -2, 2))
) %>% 
  group_by(x, y) %>%
  filter(row_number() == 1) %>% 
  head(6) 

Points$label = LETTERS[1:6]
g <- rfun(~ x + y, seed = 9823252)
gradient_plot(g(x, y) ~ x + y, domain(x=c(-4.5, 4.5), y=c(-4.5, 4.5)), n=20) %>%
  gf_text(y ~ x, label=~label, data = Points, color="red", inherit=FALSE)
```


```{r gradient-field-orange-QA11, echo=FALSE}
askMC(
  prompt = "Which is $\\partial_y f$ at point A?",
  "+neg+", "zero", "pos",
  inline=FALSE,
  random_answer_order = FALSE
)
```


```{r gradient-field-orange-QA12, echo=FALSE}
askMC(
  prompt = "Which is $\\partial_x f$ at point A?",
  "neg", "zero", "+pos+",
  inline=FALSE,
  random_answer_order = FALSE
)
```

```{r gradient-field-orange-QA13, echo=FALSE}
askMC(
  prompt = "Which is $\\partial_x f$ at point B?",
  "neg", "zero", "+pos+",
  inline=FALSE,
  random_answer_order = FALSE
)
```

```{r gradient-field-orange-QA14, echo=FALSE}
askMC(
  prompt = "Which is $\\partial_x f$ at point C?",
  "neg", "+zero+", "pos",
  inline=FALSE,
  random_answer_order = FALSE
)
```

```{r gradient-field-orange-QA15, echo=FALSE}
askMC(
  prompt = "Which is $\\partial_y f$ at point E?",
  "neg", "zero", "+pos+",
  inline=FALSE,
  random_answer_order = FALSE
)
```

```{r gradient-field-orange-QA16, echo=FALSE}
askMC(
  prompt = "Which is $\\partial_x f$ at point E?",
  "+neg+", "zero", "pos",
  inline=FALSE,
  random_answer_order = FALSE
)
```

```{r gradient-field-orange-QA17, echo=FALSE}
askMC(
  prompt = "At which letter are both the partial with respect to $x$ and the partial with respect to $y$ negative.?",
  "A", "B", "C", "D", "E", "F", "+none of them+",
  inline=FALSE,
  random_answer_order = FALSE
)
```

