---
chapter: "derivatives"
author: "Daniel Kaplan"
difficulty: "XX"
output: learnr::tutorial
tutorial:
  id: "pine-lead-car"
  version: 0.1
date: 2020-10-12 
tags: [derivatives, monotonicity]
id: pine-lead-car
---

```{r pine-lead-car-setup, include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

Recall that a function is ***monotonically increasing*** on a given domain when the function's slope is positive everywhere in that domain. A ***monotonically decreasing*** function, similarly, as a negative slope everywhere in the domain. When the slope is zero, or positive in some places and negative in others, the function is neither monotonically increasing or decreasing.

Each of the following graphs shows the derivative of some function $f(x)$. (Note: the graph doesn't show $f()$ but rather the function $\partial_x f()$ which is the derivative of $f()$.) For each graph, say whether the function $f()$ is monotonically increasing, monotonically decreasing, or neither. (Note that the horizontal scale is the same in every graph, but the vertical scale can be different from one scale to another.)

```{r echo=FALSE}
r1 <- rfun(~ x, seed = 101)
r2 <- rfun(~ x, seed = 102)
r3 <- rfun(~ x, seed = 103)
r4 <- rfun(~ x, seed = 104)

df1 <- makeFun(r1(x) + 1.5  ~ x)
df2 <- makeFun(-r2(x)  ~ x)
df3 <- makeFun(r3(x)  ~ x)
df4 <- makeFun(r4(x)/2 + 7.5  ~ x)
df5 <- makeFun(r1(x)  ~ x)
df6 <- makeFun(abs(r2(x)) ~ x)
df7 <- r1
df8 <- r2
df9 <- r3
df10 <- r4

ddf1 <- D(df1(x) ~ x)
ddf2 <- D(df2(x) ~ x)
ddf3 <- D(df3(x) ~ x)
ddf4 <- D(df4(x) ~ x)
ddf5 <- D(df5(x) ~ x)
ddf6 <- D(df6(x) ~ x)
ddf7 <- D(df7(x) ~ x)
ddf8 <- D(df8(x) ~ x)
ddf9 <- D(df9(x) ~ x)
ddf10 <- D(df10(x) ~ x)


f1 <- antiD(df1(x) ~ x)
f2 <- antiD(df2(x) ~ x)
f3 <- antiD(df3(x) ~ x)
f4 <- antiD(df4(x) ~ x)
f5 <- antiD(df5(x) ~ x)
f6 <- antiD(df6(x) ~ x)
f7 <- antiD(df7(x) ~ x)
f8 <- antiD(df8(x) ~ x)
f9 <- antiD(df9(x) ~ x)
f10 <- antiD(df10(x) ~ x)

F1 <- antiD(f1(x) ~ x)
F2 <- antiD(f2(x) ~ x)
F3 <- antiD(f3(x) ~ x)
F4 <- antiD(f4(x) ~ x)
F5 <- antiD(f5(x) ~ x)
F6 <- antiD(f6(x) ~ x)
F7 <- antiD(f7(x) ~ x)
F8 <- antiD(f8(x) ~ x)
F9 <- antiD(f9(x) ~ x)
F10 <- antiD(f10(x) ~ x)
```


```{r echo=FALSE, out.width = "50%", fig.show = "hold"}
slice_plot(df1(x) ~ x, domain(x = c(-5,5))) %>% gf_labs(title="Derivative of function A", y="df(x)")
slice_plot(df2(x) ~ x, domain(x = c(-5,5))) %>% gf_labs(title="Derivative of function B", y="df(x)")
slice_plot(df3(x) ~ x, domain(x = c(-5,5))) %>% gf_labs(title="Derivative of function C", y="df(x)")
slice_plot(df4(x) ~ x, domain(x = c(-5,5))) %>% gf_labs(title="Derivative of function D", y="df(x)")
slice_plot(df5(x) ~ x, domain(x = c(-5,5))) %>% gf_labs(title="Derivative of function E", y="df(x)") %>%
  gf_theme(axis.text.y=element_blank())
slice_plot(df6(x) ~ x, domain(x = c(-5,5))) %>% gf_labs(title="Derivative of function F", y="df(x)")
```

```{r pine-lead-car-QA1, echo=FALSE}
etude2::etudeQ(
  prompt = "Function A is ...",
    "+monotonically increasing+" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "monotonically decreasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "constant" = "A constant function has a derivative that is everywhere 0.",
    "non-monotonic" = "A non-monotonic function goes up and down, hence the derivative is positive in some places and negative in others.",
    "Can't tell from the info provided" = "This is the case if you cannot tell if the derivative is positive or negative.", 
  random_answer_order = FALSE
)
```


```{r pine-lead-car-QA2, echo=FALSE}
etude2::etudeQ(
  prompt = "Function B is ...",
    "monotonically increasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "+monotonically decreasing+" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "constant" = "A constant function has a derivative that is everywhere 0.",
    "non-monotonic" = "A non-monotonic function goes up and down, hence the derivative is positive in some places and negative in others.",
    "Can't tell from the info provided" = "This is the case if you cannot tell if the derivative is positive or negative.",
  random_answer_order = FALSE
)
```
```{r pine-lead-car-QA3, echo=FALSE}
etude2::etudeQ(
  prompt = "Function C is ...",
    "monotonically increasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "monotonically decreasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "constant" = "A constant function has a derivative that is everywhere 0.",
    "+non-monotonic+" = "A non-monotonic function goes up and down, hence the derivative is positive in some places and negative in others.",
    "Can't tell from the info provided" = "This is the case if you cannot tell if the derivative is positive or negative.", 
  random_answer_order = FALSE
)
```
```{r pine-lead-car-QA4, echo=FALSE}
etude2::etudeQ(
  prompt = "Function D is ...",
    "+monotonically increasing+" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "monotonically decreasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "constant" = "A constant function has a derivative that is everywhere 0.",
    "non-monotonic" = "A non-monotonic function goes up and down, hence the derivative is positive in some places and negative in others.",
    "Can't tell from the info provided" = "This is the case if you cannot tell if the derivative is positive or negative.", 
  random_answer_order = FALSE
)
```
```{r pine-lead-car-QA5, echo=FALSE}
etude2::etudeQ(
  prompt = "Function E is ...",
    "monotonically increasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "monotonically decreasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "constant" = "A constant function has a derivative that is everywhere 0.",
    "non-monotonic" = "A non-monotonic function goes up and down, hence the derivative is positive in some places and negative in others.",
    "+Can't tell from the info provided+" = "This is the case if you cannot tell if the derivative is positive or negative.", 
  random_answer_order = FALSE
)
```
```{r pine-lead-car-QA6, echo=FALSE}
etude2::etudeQ(
  prompt = "Function F is ...",
    "+monotonically increasing+" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "monotonically decreasing" = "A monotonically increasing function has a function that is everywhere $> 0$",
    "constant" = "A constant function has a derivative that is everywhere 0.",
    "non-monotonic" = "A non-monotonic function goes up and down, hence the derivative is positive in some places and negative in others.",
    "Can't tell from the info provided" = "This is the case if you cannot tell if the derivative is positive or negative.", 
  random_answer_order = FALSE
)
```
