The standard model of epidemics used in public health planning is called the ***SIR model***. (SIR stands for "Susceptible (S), Infective (I), Recovered (R)",  the sequence that a  person starts in,  moves to,  and ends up in  (hopefully!) in an epidemic.)

One of the equations in the SIR model is
$$\frac{dS}{dt} = -a S I$$

The notation $dS/dt$  means  "the rate of change of number of susceptibles, S,  with  respect  to time." This has dimension "people/T". The dimensions $[S]$ and $[I]$  are each simply "people."


```{r daily-digital-10-QA9, echo=FALSE}
etude2::etudeQ(
  prompt = "What is $[a]$?",
    "T" = "",
    "T$^{-1}$" = "",
    "people/T" = "Then $[a S I]$ would be  people$^3$/T, but that's not the same as $[dS/dt]$.",
    "+people$^{-1}$ T$^{-1}$+" = "This correctly  gives $[a S I]$ as people/T, which  is the same as $[dS/dt]$.",
    "people $\\times$ T" = "",
    "None of the above." = ""
)
```


Exercise 57 refers to a second equation of the SIR model

$$\frac{dI}{dt}  = - a S I - b I$$
where  $[\frac{dI}{dt}] =$ people/T.

```{r daily-digital-10-QA10, echo=FALSE}
etude2::etudeQ(
  prompt = "What is $[b]$?",
    "T" = "",
    "+T$^{-1}$+" = "",
    "people/T" = "Then $[a S I]$ would be  people$^3$/T, but that's not the same as $[dS/dt]$.",
    "people$^{-1}$ T$^{-1}$" = "If this were true, $[bI]$ would be T$^{-1}$. But $[bI]$ has to be the same as $[dI/dt]$, which is people $T^{-1}$.",
    "people $\\times$ T" = "",
    "None of the above." = ""
)
```
