---
title: "Flies in a mangrove swamp"
author: "UND 1.5"
runtime: shiny_prerendered
---

```{r include=FALSE}
etude2::load_learnr_safely()
library(etude2)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```

```{r fly_dyn_def, context="setup", echo=FALSE}
fly_dyn <- function(x, K=1e3, r=1.75e-4) {
  ifelse(x < K,
         0.01*x^2,
         0.01*K^2*exp(-r*(x - K)))
}
```

The year-to-year growth/decline in the number of flies in a mangrove swamp is modeled by the function `fly_dyn()`:
$$x_{t+1} = \mbox{fly_dyn}(x_t)$$

The sandbox shows the function and enables you to perform other calculations using it.

```{r mangrove1, exercise=TRUE, exercise.cap="Yearly fly population.", exercise.nlines=8, exercise.setup="fly-dyn-def"}
slice_plot(fly_dyn(x) ~ x, domain(x=c(0,6000)))
```

```{r mangrove2, echo=FALSE}
askMC(
  "At which of these values of $x_t$ do the dynamics have a fixed point? (Hint: How would you confirm that a given value is a fixed point?)",
  "+0, 100, 4981+",
  "0, 100, 121" = "When you give 121 as the input, do you get 121 as the output?",
  "0, 121, 5683",
  "0, 100, 5683" = "When you give 5683 as the input, do you get 5683 as the output?"
)
```

```{r mangrove3, echo=FALSE}
askMC(
  "Which one of these values for $x$ is an **unstable** fixed point?",
  0, "+100+", 121, 4981, 5683,
  random_answer_order = FALSE
)
```

An important vocabulary word in dynamics is *transient*. In everyday speech, this means something like "just passing through." It's the same in dynamics: that part of the trajectory which precedes stable, fixed behavior such as at a fixed point. Transients occur whenever a dynamical system has an initial state not on a stable, fixed state. They are also common in systems that are disrupted by some external force, for example in the recovery of an electrical power distribution grid after a disturbance such as an ice storm. After a sharp bang, the ringing in your ears is a transient. When you stand up too suddenly, the "stars" you see are a transient due to diminished blood flow. Turn on an oven? The transient is the warming up until the oven reaches the temperature setting. 

Although transients are ... well, transient, they can be very important. Key to the Wright Bros. success was their recognition that air turbulence elicits transients in attitude and that aircraft need control systems that can work fast enough for the craft to survive the transient. If you have driven a car with a broken suspension, you'll know that it can be hard to control after the transient caused by hitting a bump in the road.

Small disturbances often elicit transients that decay away exponentially. Such transients, like all exponentially decaying processes, can be characterized by a half life: the time it takes the transient to shrink to half its original value. (Not all transients decay exponentially, but that's a story for another course.)

Using the sandbox, iterate `fly_dyn()` for 50 years starting at $x_0 = 4500$. Here's the command to do this, which you'll have to copy to the sandbox.

```{r results="hide", eval=FALSE, echo=TRUE}
Trajectory <- Iterate(fly_dyn, x0=4500, n=50)
gf_point(x ~ n, data = Trajectory) %>%
  gf_line(alpha = 0.2)
```

```{r mangrove4, echo=FALSE}
askMC(
  "There is a transient that oscillates while exponentially approaching the fixed point. What is the half-life of the exponential transient?",
  "1 year",
  "3 years",
  "+6 years+",
  "12 years",
  "24 years",
  random_answer_order = FALSE
)
```
