---
title: "Stability 2"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(shiny)
library(mosaic)
library(mosaicCalc)
```

Previously, we examined the stability of fixed points by using iteration: start at an initial condition arbitrarily close (but not exactly on!) a fixed point and see if the trajectory tends toward or away from the fixed point.

Now we'll look at stability another way, by considering the shape of of the dynamical function near a fixed point. At this point in CalcZ, you're aware that "near" suggests local, and that we routinely model (continuous) functions in terms of the value at a point, the value of the derivative at the point, and the value of the 2nd derivative at the point.

The stability of a fixed point is determined in all but very special situations by the slope of the function at the fixed point. Here's the rule: 

> In the dynamical system $x_{n+1} = g(x_n)$ with a fixed point at $x^\star$, the stability of that fixed point depends only on the magnitude of $\left| \partial_x g(x^\star) \right|$. The rule is: 

> If $\left| \partial_x g(x^\star) \right| > 1$, the fixed point at $x^\star$ is **unstable**.

> If $\left| \partial_x g(x^\star) \right| < 1$, the fixed point at $x^\star$ is **stable**.

Notice that it doesn't matter what is the **sign** of $\partial_x g(x^\star)$; the fixed point will be stable or not depending just on the magnitude of the derivative.

```{r stb2-1, exercise=TRUE, exercise.cap="Stability sandbox", exercise.nlines = 6, eval=FALSE}
g <- makeFun(sin(x)^2 + 1 ~ x)
fixed_points <- findZeros(g(x) - x ~ x)
dg <- D(g(x) ~ x)

## Now, apply dg() to the fixed point(s)
dg(fixed_points)
```

```{r stb2-2, echo=FALSE, results="markup"}
askMC(
  "Calculate the derivative with respect to $x$ of $g(x) \\equiv  \\sin(x)^2 + 1$ at the fixed point at $x^\\star = 1.8972$. Is that fixed point stable?",
  "+Yes+",
  "No",
  random_answer_order = FALSE
)
```



