---
title: "An IV Drip"
author: "Daniel Kaplan"
runtime: shiny_prerendered
---

```{r include=FALSE}
library(etude2)
library(mosaic)
etude2::load_learnr_safely()
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```

Morphine is administered to a patient intravenously at a rate of 2.5 mg/hr. About 35% of the morphine is metabolized and leaves the body each hour.

```{r iv1, echo=FALSE}
explain1 <- "It says, \"About 35% of the morphine ... leaves the body each hour.\" Equal proportions in equal times implies an exponential function for the solution."
etude2::etudeQ(
"Which differential equation best describes the amount of morphine in the body as a function of time ($M$ is in mg; $t$, in hr)?",
    "$dM/dt = 2.5 - 0.35$" = explain1,
    "$dM/dt = 2.5 - 0.35M$" = "The differential equation $dM/dt = -0.35M$ gives a solution of the form $M=M_0 \\exp(-0.35 t)$. Evaluate this solution and see if you get $(1-0.35) M_0$. What would $\\alpha$ have to be so that $\\exp(-\\alpha t) = 0.65. (Why 0.65? If 35% leaves the body, then 65% remains. The state is the amount of morphine in the body, so all accounting must be done using this.)",
    "+$dM/dt = 2.5 - 0.43M$+" = "Right. The 2.5 is the drip rate in mg/hr, and the 0.43 is the fraction of the drug that would be left after an hour (without any new drug being administered).",
    "$dM/dt = 2.5 - 0.43$"= explain1
)
```

```{r iv2, echo=FALSE} 
explain <- "Plug in $\\partial_t dM = 0$ into the dynamics and solve for $M$."
etude2::etudeQ(
"After a day on the IV drip, the blood level of morphine is at a fixed point. (More precisely, the level is extremely close to the fixed point.) What is the level $M$ (in mg) at this fixed point?",
"+2.5 / 0.43+" = "You solved $0 = 2.5 - 0.43M$ for $M$. Good thinking.",
"0.43 / 2.5" =  explain1,
"2.5 $\\times$ 0.43" = explain, 
"Not enough information was given",
"Enough information is given but the answer is none of the above." = "Then the problem author made a mistake!"
)
```

Write an R function that can work as a reference to the doctors and nursing staff so they can translate the desired drug level (in mg) to sustain, along with the metabolization rate (35% in the previous example), to get the IV drip rate needed . Your formula should be written in terms of the desired `drug_level` and the `metabolization` rate as a fraction (e.g. 0.35).

You will want to refer to the equations you used in the previous two questions. In those questions, the drip rate was set at 2.5 mg/hr, but now the drip rate is what you want calculate as the output of your function. And remember the logic that leads to transforming the metabolizatioin rate into the appropriate $\\alpha$ in the differential equation for the amount of drug in the body. 

```{r iv3, exercise=TRUE, exercise.cap="The IV drip rate should be ...", exercise.nlines=5}
makeFun( ...your_formula_here... 
         ~ drug_level + metabolization)
```

```{r iv3-check}
grade_result(
  pass_if( ~ all(abs(.result(1:3, 0.35) - (1:3) / abs(log(1-0.35))) < 0.1)),
  fail_if( ~ all(abs(.result(1:3, 0.35) - (1:3) / abs((1-0.35))) < 0.1), 
           message="You forgot about the logarithm."),
  fail_if( ~ all(abs(.result(1:3, 0.35) - (1:3) / abs((0.35))) < 0.0001), 
           message="You forgot about the logarithm."),
  fail_if( ~ all(abs(.result(1:3, 0.35) - (1:3) / abs(log(0.35))) < 0.0001), 
           message="You forgot about `1-metabolism`."),
  fail_if( ~ TRUE)
)
```
