---
title: "Daily Digital 18"
subtitle: "Math 142Z"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
library(math141Z)
```


```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD-18") # Add the DD number
etude2::etude_document_name("DD-142Z-18") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
```

```{r eval=FALSE}
# Leave this safety check here
if (etude2::etude_document_name() == "DD-") 
  stop("You must set the master_id() and the document name specifically and uniquely for this document")
```

## Objectives

1. Learn the mechanics of iterating a two-dimensional finite-difference equation by hand and on the computer for linear systems.
2. Meet eigenvalues for the first time and see how their values correspond to the characteristic patterns of motion in linear systems. 
3. Associate the shape of a trajectory in state-space to the graph of a time series of a component of the state.


## `Iterate()` with 2-dim systems 


```{r child="iterate-two-D.Rmd"}
```


## Stepping through space

```{r child="stepping-through-space.Rmd"}
```

## Characterizing matrix patterns

```{r child="matrix-patterns.Rmd"}
```

## Eigenvectors

```{r child="eigenvectors.Rmd"}
```

<!--
## Time lag format

```{rr child="time-lag.Rmd"}
```
-->

<!-- 
## Word meanings (DRAFT)


```{rr child="definitions.Rmd"}
```
-->

## ~~Hand in your work~~

<div style="color: red;">USAFA CalcZ students will hand in their work via GradeScope</div>


<!--
REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>


```{rr echo=FALSE, messages=FALSE}
etude2::submissions_ui()
```

```{rr context="server", echo=FALSE, messages=FALSE}
etude2::submissions_logic()
```

-->


-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name()`

