---
title: "Eigenfunctions"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library("etude2")
```

Throughout CalcZ, we have emphasized a small set of modeling function families:

i. low-order polynomials, e.g. straight-line functions, quadratics 
#. exponentials $e^{kt}$
#. sinusoids e.g. $\sin(2 \pi t/P)$
#. logarithms
#. power-law functions
#. sigmoidal functions
#. hump functions

In studying dynamics, the first three of these will be particularly prominent, so much so that you will only rarely have reason to consider the others. So much of the drama will take place in the context of the three leading families that it's worth reviewing their algebra and properties and keeping them on the top of your mind and the tip of your tongue. 

Recall that there are three fundamental entities in dynamics:

a. the **state**. In our work, the state will almost always be a single *instantaneous* quantity or a pair of *instantaneous* quantities. All the phenomena of differential dynamics---except one---can be illustrated quantitatively with systems with states that are one- or two-dimensional vectors. (The phenomenon we can't reach with only one- or two-dimensions in continuous time is called "chaos.") For consistency, we'll generally use $x$ and $y$ to refer to the instantaneous quantities, sometimes using their Greek equivalents $\xi$ and $\eta$ when we need to tell the "backstory" of where the dynamics came from.
    
    Note that when we refer to $x$ and $y$ as *instantaneous* quantities we mean that they are themselves **functions of time** or an analogous quantity. (Recall that a quantity like velocity can vary; it's a function of time. The quantity at a given, single instant of time is *instantaneous*: the quantity at that *instant*.) This goes back to the basics of calculus, where the derivative of a function of time (such as position) gives the change in that quantity at each instant of time).
    
    
b. the **dynamics**, that is, the function(s) that relate the change in state to the value of the state. We will, with few exceptions model the dynamics using low-order polynomials. For systems with a one-dimensional state, the fundamental form will be the straight-line function $$f(\xi) = a_0 + a_1 \xi .$$  
    
    For systems with a two-dimensional state, the fundamental form will be $$g(\xi, \eta) = a_0 + a_x \xi + a_y \eta + a_{xy} \xi\, \eta$$ The most important algebraic processes will be finding zeros of these forms and re-writing the functions on the scales $$x = \xi - \xi_0\\y = \eta - \eta\,_0$$ Don't worry too much about the use of Greek names, since almost always we'll use the re-scaled variables that we're giving Roman names.


c. **trajectories/"solutions"**. Here the important functions are $e^{kt}$ and $\sin(2 \pi t/P)$ and $\cos(2 \pi t/P)$. These will be used so often in tandem, that it will be convenient to write the parameters for the sinusoids in a simpler way: $\sin(\omega t)$ and $\cos(\omega t)$. There's no fundamental change here. It's just that we'll write the verbose $2 \pi/P$ with a single character, $\omega$, just as we might refer to  a friend name "William" with the shorter "Bill."  

The exponential and sinusoids are fundamental to differential dynamics because they have these familiar properties:

- $\partial_t\, e^{kt} = k\, e^{kt}$: the first derivative of an exponential is proportional to itself.
- $\partial_{tt}\, \sin(\omega t) = - \omega^2 \sin(\omega t)$: the second derivative of a sinusoid is proportional to itself.

Your life among differential equations will be easier if you remember these three things: 

- functions of the state $x$ and $y$ are generally the *dynamics*
- functions of time (we'll almost always use time) are generally the *solutions* to the dynamics.
- **differential equations** relate the derivatives of the *solutions* to the *dynamics*, e.g. $\partial_t\, x(t) = f(x(t))$. Although almost always we will streamline the notation by writing $\partial_t\, x = f(x)$, do remember that $x$ is a function of $t$, that is, $x(t)$.

Speaking of streamlining notation, physicists and often engineers use an even more concise notation, replacing $\partial_t$ (which is our computer-ready abbreviation for the $\frac{d}{dt}$ in convential textbook notation) with the ultimate in compactness: $\dot{\ }$ That single little dot was Newton's notation for a derivative. Two dots, $\ddot{\ }$ stands for a second derivative: $\partial_{tt}$ (or $\frac{d^2}{dt^2}$ in the conventional notation). 

```{r eig1-1, echo=FALSE, results="hide"}
askMC(
  "The most common differential equation we will encounter can be written $$\\dot{x} = \\lambda x$$ Which of these is the equivalent expression when written using the more verbose $\\partial_t$?",
  "+$\\partial_t\\, x(t) = \\lambda x(t)$+",
  "$\\partial_{tt}\\, x(t) = \\lambda x(t)$" = "The single dot is a first derivative, not a second derivative like $\\partial_{tt}$.",
  "$\\partial_t\\, x(t) = \\lambda \\partial_t\\, x(t)$" = "A differential equation relates the derivative of a state (e.g. $\\partial_t\\, x$) to the state itself (e.g. $x$). Besides, this equation could be true only when $\\lambda = 1$, which makes it a tautology.",
  "$\\partial_{tt}\\, x(t) = \\lambda^2 \\partial_t\\, x(t)$" = "This equation is in fact mathematically equivalent to $\\partial_{t}\\, x = \\lambda x$ as is $\\partial_{ttt}\\, x = \\lambda \\partial_{tt}\\, x$ or $$\\partial_{tttt}\\, x = \\lambda \\partial_{ttt}\\, x$ (and so on), but nobody would write things this way."
)
```

```{r eig1-2, echo=FALSE, results="markup"}
askMC(
  "What's the solution to $\\partial_t\\, x = \\lambda x$?",
  "+$x_0\\, e^{\\lambda t}$+" = "Notice that when $t=0$ this becomes $x_0$, the initial condition.",
  "$e^{\\lambda x_0}$" = "Solutions are functions of $t$",
  "$t^{\\lambda x_0}$" = "When $t=0$, this does not reduce to the initial condition $x_0$",
  "$x_0^{\\lambda t}$" = "When $t=0$, this reduces to $1$. But the point is the solution at $t=0$ should be the initial condition $x_0$."
)
```

```{r eig1-3, echo=FALSE, results="markup"}
askMC(
  "What's the solution to $\\partial_t\\, x = \\lambda x$?",
  "+$x_0 e^{\\lambda t}$+" = "You will want to memorize this!",
  "Something else" = "No!!!!!!"
)
```

```{r eig1-4, echo=FALSE, results="markup"}
askMC(
  "What's the solution to $\\partial_t\\, x = \\lambda x$?",
  "+$x_0 e^{\\lambda t}$+" = "No joke. This should become second nature!",
  "Anything else" = "Stop torturing me! Straighten up and fly right."
)
```

