---
title: "Daily Digital 35"
subtitle: "Math 142Z"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
library(mosaic)
library(mosaicCalc)
library(math141Z)
gradescope=TRUE
```


```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD-24") # Add the DD number
etude2::etude_document_name("DD-142Z-24") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
```

```{r eval=FALSE}
# Leave this safety check here
if (etude2::etude_document_name() == "DD-") 
  stop("You must set the master_id() and the document name specifically and uniquely for this document")
```

## Objectives

* Be able to recognize situations where there is more than one objective to be achieved simultaneously.
* Distinguish between **commensurate** and **incommensurate** objectives.
* Understand that in general there is no purely mathematical solution to optimization with multiple incommensurate objectives, but be able to distinguish between dominating and non-dominating choices.

<!--
## What to say about the "best"?

* https://en.wikipedia.org/wiki/Perfect_is_the_enemy_of_good
* The best is the enemy of the good. 
* Better an approximate answer to the right question ...
* Radar: 'Robert Watson-Watt, who developed early warning radar in Britain to counter the rapid growth of the Luftwaffe, propounded a "cult of the imperfect", which he stated as "Give them the third best to go on with; the second best comes too late, the best never comes."' https://en.wikipedia.org/wiki/Robert_Watson-Watt

-->

## Commensurate and incommensurate

```{r child="incommensurate.Rmd"}
```


## Multiple objectives

```{r child="multiple-objectives.Rmd"}
```

## Yours and mine

```{r child="yours-and-mine.Rmd"}
```

## QALY and life years

```{r child="three-policies.Rmd"}
```

<!--

## Optimize on a constraint

Computer exercise where we parameterize the constrain with one variable, then translate the objective function into a function of that one variable.

-->

## Hand in your work

<div style="color: red;">USAFA CalcZ students should hand in their work via GradeScope</div>

<!--

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>


```{rr echo=FALSE, messages=FALSE}
etude2::submissions_ui()
```

```{rr context="server", echo=FALSE, messages=FALSE}
etude2::submissions_logic()
```

-->

-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name()`

