---
title: "QALY constraint substitution"
author: "Daniel Kaplan"
---

```{r include=FALSE, contect="setup"}
library(mosaic)
library(mosaicCalc)
library(etude2)
```

In a previous exercise, a graph of the hypothetical impact of three forms of health-care expenditures was presented.  Here is a similar exercise, which will explore a constrained optimization using slightly  hypothetical functions, $f_1$, $f_2$, and $f_3$:
```{r fundef}
f1 <- makeFun(500*pnorm(x, mean=m, sd=s) + .05*x - 11.375 ~ x,  m = 300, s =  150)
f2 <- makeFun(350*pnorm(x, mean=m, sd=s) + 0.5*x - 175 ~ x,
              m =  0, s  = 100)
f3 <- makeFun( x/1.6 ~ x )
```
 

```{r qalyfuns,fig = TRUE,include = TRUE,echo = FALSE}
Labels <-  tribble(
  ~ x, ~ y, ~  label, ~ color,
  1000, f1(1000) - 50, "f1", "black",
  900, f2(1000), "f2", "orange3",
  1050, f3(1000), "f3", "dodgerblue",
)
slice_plot( f1(money)~money, domain(money=c(0,1000)),lwd=3)  %>%
  gf_lims(y=c(0,700)) %>%
  gf_labs(x="Expenditure ($1000)", y="QALYs added") %>%
  slice_plot( f2(money)~money, color = "orange3") %>%
  slice_plot( f3(money)~money, color = "dodgerblue") %>%
  gf_text(y  ~ x,  label = ~ label, data = Labels)
```
 

The problem is how to allocate the available funds, $1,000K, among the different health-care interventions.  

One strategy for solving constrained optimization functions is to use the constraint to reduce the ``dimension" of the problem.  In the unconstrained problem, there are three quantities to be set: $x_1$, $x_2$, and $x_3$, the amount of money spent on each intervention.  The constraint imposes a relationship among the three quantities, namely that $x_1 + x_2 + x_3 = 1000$.  This means that there are really only two independent quantities, since setting two of the quantities determines the value of the third: a reduction in dimension from three to two.

To see this, construct the overall output from any pattern of expenditure meeting the constraint and plot it out:
```{r overall}
contour_plot( f1(x1) + f2(x2) + f3(1000-(x1+x2)) ~ x1 + x2, 
        domain(x1=range(0,1000), x2=range(0,1000)) )
```
 

Make sure that you understand why the overall function has been constructed this way: $x_1$ is the amount spent on intervention one, $x_2$ is the amount spent on intervention two, leaving $1000-x_1-x_2$ to be spent on intervention three.

The constraint, rather than being graphed, has been used to turn a function of three variables $x_1$, $x_2$, and $x_3$, into a function of just two variables. 

From the graph you can read off the values of $x_1$ and $x_2$ that maximize the overall function subject to the constraint.  Then you can use the constraint to find the value of $x_3$.  To get a more precise estimate of the location of the maximum, remake the plot in a narrow region around the peak.

1. What are the values of $x_1$ and $x_2$ at the optimum?
    i. $x_1$: {494.25}{475.25,483.25,494.25,498.25}
    #. $x_2$: {219.70}{219.70,221.70,223.70,225.70}

2. The derivative of the functions tells how much the output changes for an additional dollar of expenditure.  Calculate the derivatives at the optimum point:
    i. For function 1: $\frac{d f_1}{d x_1}$ {0.625}{0.000,0.500,0.523,0.625,1.20,1.37}
    #. For function 2: $\frac{d f_2}{d x_2}$ {0.625}{0.000,0.500,0.523,0.625,1.20,1.37}
    #. For function 3:$\frac{d f_3}{d x_3}$  {0.625}{0.000,0.500,0.523,0.625,1.20,1.37}

NEED TO FIX BUG IN stats::D() which doesn't capture parameters to pnorm() when  differentiating.

<!--begin-answer-->
    ```{r }
df1 <- numD( f1(x) ~ x) 
df2 <- numD( f2(x) ~ x)
df3 <- D( f3(x) ~ x )
df1(x = 494.25)
df2(x = 219.70)
df3(x =  1000 - (494.25 + 219.70))
```
<!--end-answer-->

#. Explain what aspect of your answer to the above questions about the derivative tells you that your values of $x_1$, $x_2$ and $x_3$ are optimum?

<!--begin-answer-->
    All the derivatives are the same at the optimal point.  This reflects the optimum in that any small reallocation of funds from one intervention to the next will result in no net change in output.  If the derivatives of the different interventions were not the same, it would be possible to improve the overall outcome by a reallocation.
<!--end-answer-->
