---
title: "MMAC ยง2.1"
---

```{r include=FALSE, context="setup"}
library(etude2)
etude2::load_learnr_safely()
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
```

```{r context="setup"}
Prescription_sales <- tibble::tribble(
  ~ year,  ~ sales,
  2006, 262.7,
  2007, 261.6,
  2009, 266.8,
  2010, 266.4)
```


### Exercises 53, 56, & 58

These three exercises are based  on those on page 135 of the MMAC textbook. We have altered them somewhat to use appropriate standards for data storage and updated graphics software,

```{r}
Prescription_sales
```

The data frame  `Prescription_sales` is available to you for use in the code box below.  Sales are in billions of US dollars.

The  following computer commands will be useful in answering the questions:

```r
gf_point(sales ~ year, data = Prescription_sales) %>%
  gf_lims(x  = c(2005, 2015))
  
model_1 <- fitModel(sales ~ m*year + b, data = Prescription_sales)

model_1 # shows the model as implmented in software

model_1(year = 2004)
```

Lines ending  in the "pipe" symbol  (`%>%`) are commands that are continued on the next line. To form a complete command, you need to include all the lines that are connected with the pipe. 

```{r MM2-1-S1, exercise=TRUE, exercise.cap="Prescription drug sales ", exercise.setup="daily-digital-05-E1-setup", exercise.eval=FALSE, exercise.lines=5, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE}


```

In MMAC exercise  2.1.53, you are asked to plot the data and verify that the straight-line model is reasonable. This is a bit silly, since there are only 4 data points. Besides, whether a given model is reasonable can only be determined relative to the purpose for which the model is being constructed.  The textbook  doesn't explain what the purpose is.

The `fitModel()` command will generate what the textbook authors call the "best possible" straight-line model. Keep in mind that "best" does not necessarily mean good!

The output of `fitModel()` is itself a function, as you can see from the `function(year, ...)` that starts the output. The algorithm in the function is in software jargon most of which will not make any sense to you. Still, you should be able to figure  out what the parameters of the model are from the output.

```{r MM2-1-1, echo=FALSE}
askMC(
  prompt = "**A)** According to the output of `fitModel()`, what is the *slope* of the \"best possible\"  line?",
  "+1.26+", -2306, "0x7fa2718423a8",
  random_answer_order = FALSE
)
```


```{r MM2-1-2, echo=FALSE}
askMC(
  prompt = "What is the value (rounded to one decimal place) of the fitted model at an input of `year = 2008` ?",
    "259.3"  = "That's the output for `year = 2004`.",
    "260.6"  = "That's the output for `year = 2005`.",
    "+264.4+" = "",
    "None of the above" = "",
  random_answer_order = FALSE
)
```

The textbook asks you  to evalute the fitted model for `year = 2015`, treating this as a purely mathematical exercise. Do so and use the value to answer the following question. 

```{r MM-2-1-3, echo=FALSE}
etude::true_or_false(
  prompt = "True or False? The model output for  `year = 2015` falls inside the range of the previous year's sales data?",
  FALSE,
  allow_retry = FALSE)
```

Add a line to the graphics command that will draw the "best possible" line on top of the data. Do this by inserting the following line between the `gf_point()` and the `gf_lims()` lines:

`gf_lm() %>%`

Note the pipe symbol at the end of the above line, which says, more or less, "after you do me, go on to do the next thing."

Let's presume that the objective behind evaluating the model at `year = 2015` is to make a  prediction of what drug sales will be in  that year.

Regretably, the textbook leaves out a fundamental matter when using a "best possible" model to make a prediction. This is the matter of how precise the prediction is. A standard format for indicating such precision is called a "95-percent prediction interval." You can add a 95-percent  prediction interval to the graph by placing the following as *an argument to* `gf_lm()`, that is

`gf_lm(interval = "prediction")`

```{r M2-1-4, echo=FALSE}
askMC(
  prompt = "True or false? The 95-percent prediction interval is consistent with a possibility that sales in 2015 will be *less* than in 2010?",
  "+TRUE+",
  allow_retry = FALSE)
```
