---
title: "Daily Digital 24"
subtitle: "Math 142Z"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```


```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD-24") # Add the DD number
etude2::etude_document_name("DD-142Z-24") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
```

```{r eval=FALSE}
# Leave this safety check here
if (etude2::etude_document_name() == "DD-") 
  stop("You must set the master_id() and the document name specifically and uniquely for this document")
```

## Objectives

* Use low-dimensional polynomials to construct global, nonlinear differential-equation models.
* Become familiar with "classic" nonlinear differential-equation models.
* Be able to interpret individual terms in the dynamical function as describing specific relationships among real-world objects.

## Attrition

```{r child="Lanchester.Rmd"}
```

## Predation

```{r child="predation.Rmd"}
```

<!--
## Predation near the fixed point


child="predation.Rmd"
-->

## Numerics 

```{r child="predation-numerics.Rmd"}
```

<!-- 
We'll also do van der Pol oscillator, to show a system with a limit cycle.

## Conserved rabbits

You've heard about the "conservation of energy." There's an analogous situation in the predator-prey dynamics: a function 

In which we'll propose a few candidate Lyapunov functions and ask students to show that the quantity is preserved and to plot out the resulting orbits as a contour plot

$$L(R, B) = \gamma \ln{R} - \mu R + r \ln{B} - \beta B$$
and some others that aren't right.

-->

<!-- 
## Maybe

Level curves of energy function $v^2 + \int \frac{1}{r} dr$ to look for elliptical orbits.

And/or, maybe work this up as an example to be used in the first semester for contour plots.


## THE FOLLOWING WILL BE REMOVED

The project will be an in-depth examination of SIR

## Epi(demi)c trajectory!

child="SIR-I.Rmd"

-->

## For practice

```{r child="arms-race.Rmd"}
```

<!--

## Guerilla fighting

child="guerilla-fighting.Rmd"

## Lighting a match

An example with a quadratic polynomial in one variable

child="match-lighting.Rmd"

-->

## Hand in your work

<div style="color: red;">USAFA CalcZ students should hand in their work via GradeScope</div>

<!--

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>


```{rr echo=FALSE, messages=FALSE}
etude2::submissions_ui()
```

```{rr context="server", echo=FALSE, messages=FALSE}
etude2::submissions_logic()
```

-->

-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name()`

