---
title: "A match lighting"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(etude2)
```


Consider this video showing a match lighting.

<iframe width="560" height="315" src="https://www.youtube.com/embed/bFi_UkNSRoc?start=9" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

Without careful quantitative analysis of the video, it's hard to say whether the growth of the flame is exponential or not. But let's explore a possibility.

Imagine that the sulfur in the match head vaporizes when heated and the heat generated by striking the match is enough to ignite the airborne sulfur in a small region of radius $r$. The heat added by the igniting sulfur brings more volume up to a sufficient temperature to ignite the sulfur vapor, the reaction being limited by the amount of oxygen that can cross into the sufficiently hot region.

Modeling the flame by a ball of radius $r$ at or above ignition temperature, the ball will grow in time. If the growth is proportional to the surface area of the ball, a differential equation modeling the system is $$\frac{dr}{dt} = k r^2$$

This is an example of a "separable" differential equation. It involves two variables $r$ and $t$, and the equation can be re-arranged to put all the $r$ elements on the left side and all the $t$ elements on the right side $$\frac{dr}{r^2} = k dt$$

Integrate both sides of this equation to get $$-1/r = kt + C$$

Solve this equation for $r$ as a function of $t$.

```{r match1, echo=FALSE, results="markup"}
askMC(
  "As always, the constant of integration $C$ reflects the initial conditions of the function $r(t)$. Suppose the ignition takes off at $t=0$, at which point the ball of hot, sulfurous gas as a radius $r(0) = 1$ mm. What's the numerical value of $C$?",
  "Not enough information. I need to know $k$.",
  "+$C= -1 mm^{-1}+" = "Looking at the solution, $$r(t) = 1/(-kt -C)$$ you can see that at time $t=0$ the radius is $-1/C$.",
  "$C=0$",
  "$C= -1 mm^2$",
  "$C=- 1 mm$"
)
```

```{r match2, echo=FALSE, results="markup"}
askMC(
  "Unlike exponential growth, which proceeds steadily, our differential equation has a solution that *explodes*, going to $\\infty$ at a finite time $t$. What's that time? (Use the value of $C$ from the previous question.)",
  "$t=1$",
  "+$t = 1/k$+",
  "$t = k$",
  "$t = \\sqrt{2k}$"
)
```  

Thankfully for the existence of the universe, the radius of an actual matches flame ball does not reach infinity! 

```{r match3, echo=FALSE, results="markup"}
askMC(
  "Using your physical intuition, what feature is missing from our simple model of match ignition that might account for why the flame ball doesn't grow to the size of a beach ball in an instant?",
  "+There's a limited amount of sulfur in the match head.+",
  "The oxygen can't approach the flame from the stick-side.",
  "The flame ball isn't a ball; it's flame-shaped!" 
)
```

We developed the differential equation based on an assumption of how matches work. This assumption is not necessarily true. Nonetheless, models like this point out that simple mechanisms can produce growth that is faster than exponential.
