---
title: "Following SIR"
author: "Daniel Kaplan"
runtime: shiny_prerendered
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(scales)
library(mosaicCalc)
library(math141Z)
```

```{r echo=FALSE}
SIR_flow <- 
  function(alpha = 0.5, beta = 0.001, new = 5, 
           dom = domain(I=c(0,300), S=c(0,1000)), nsteps=5) {
    # I HATE THIS. streamlines() isn't looking in the right environment
    Irule <<- makeFun(beta*S*I - alpha*I ~ I & S, 
                     beta=beta, alpha = alpha)
    Srule <<- makeFun(new - beta*S*I ~ I & S, 
                     beta=beta, new=new)
    math141Z::streamlines(list(~ Irule(I, S), ~ Srule(I, S)), #list(~ Irule(I, S), ~ Srule(I, S)), 
                          dom, n=15, nsteps=nsteps) 
  }
```


With the onset of the COVID-19 pandemic, many concepts from calculus have been front-and-center in the news. Last semester we looked at the exponential nature of epidemic growth and decay. In this section, we'll look in more detail at the dynamics of epidemics and see how mathematical concepts such as stability and bifurcation underlie the way we communicate about them, such as with the now-famous $R_0$, the "basic reproduction number."

There is a standard mathematical framework for describing the spread of communicable diseases. This is feasible because the *local dynamics* of spread are can be described simply, even if the global patterns produced are complicated. The calculus strategy---zoom in on a small part of the whole to construct a model, work out the consequences of that model, then assemble the many small parts together into the model of the whole.

All differential equation models involve "zooming in" with respect to time. Perhaps better to say, "involve working with snapshots." Even though we're interested in how systems move in time, we freeze time. The model doesn't describe the trajectory of the system. Instead, it gives the instantaneous rate of change with respect to time, as if you took a photo of a car's speedometer in order to figure out where the car might be going. We leave it to the more-or-less automatic mechanisms of calculus---anti-differentiation, Euler steps, our algebraic fluency with differentiation---to figure out the trajectory over time from the information.

We're going to explore the standard model of infectious disease spread: the so-called SIR model. The state in this model has three components:

i. $S$ the number of people susceptible to infection. Typically, news reports about the COVID pandemic don't give this number, but you may see reports of the **proportion** of the population who are susceptible.
ii. $I$ the number of people who are currently infectious. Again, typically news reports don't give this number, preferring instead to give the number of new cases each day or the total number of people who have gotten COVID-19 since the start of the pandemic.
iii. $R$ the number of people "recovered" after having gotten COVID-19. In terms of the SIR model, this means the number of people who were infectious at some point in the past but aren't currently infectious and aren't susceptible. (It's unclear with COVID-19 whether everyone who has had COVID-19 remains unsusceptible, but we'll start with that assumption.)

The snapshot we take is based on conditional reasoning: If the state were here, how would it move. We do this for many different possible states and graph the resulting flow. 



```{r echo=FALSE, out.width="400pt", out.height="400pt"}
Pts <- tibble::tribble(
  ~ I, ~ S, ~ label,
  150, 750, "1",
  150, 250, "2",
  100, 500, "3"
)
SIR_flow(nsteps=10, dom=domain(I=c(0,200), S=c(0,1000))) %>%
  gf_lims(x = c(0,200)) %>%
  gf_label(S ~ I, label=~label, data =Pts, alpha = 1, group = NA, size=NA)
```


```{r sir1-1, echo=FALSE, results="markup"}
askMC(
  "At the state labeled [1], how many susceptible and how many infective people are there?",
  "$S=500, I = 150$",
  "+$S=750, I=150$+",
  "$S=1000, I = 50$", 
  random_answer_order=FALSE
)
```
  
```{r sir1-2, echo=FALSE, results="markup"}
askMC(
  "At the state labeled [1], how will $S$ and $I$ be changing?",
  "$S$ going up, $I$ going up",
  "$S$ going up, $I$ going down",
  "+$S$ going down, $I$ going up+",
  "$S$ going down, $I$ going down",
  random_answer_order=FALSE
)
```

```{r sir1-3, echo=FALSE, results="markup"}
askMC(
  "Is there any place in the state space where both $S$ and $I$ are increasing?",
  "Yes",
  "+No+" = "The reason is that the model imagines that susceptible people are being converted to infectious people (that is, when susceptible people get sick, they become infective.",
  random_answer_order=FALSE
)
```


Consider the flow near label [2]. The number of susceptibles is going down and the number of infectives are also going down. Why? Infective people naturally recover. (Regretably, "recover" includes "die.") If there are lots of susceptible people and lots of infective people, then the recovering infectives are more than balanced by the number of new infectives. That's the story near label [1].

Near label [2], the number of susceptible isn't so big, so there are few new infectious people.

Make sure you understand the two preceding paragraphs before moving on. If you can't make sense of them, ask for help.

```{r sir1-4, echo=FALSE, results="markup"}
askMC(
 "Near label [3], the flows are vertical and downward. This means that the number of susceptibles is decreasing, but the number of infectives is staying the same. What's happening?",
 "+There are just enough susceptibles to replace those infectives who recover.+",
 "Some other illness is removing people from the population." = "Possible, but ... The nice thing about a model is that *we make the rules*. This model focuses on the pandemic.",
 "The infectives are becoming susceptible again at just the rate that susceptibles become infective." = "If this were true, the number of susceptibles would stay level, not decrease."
)
```

The model isn't intended to account for where the original infectives come from. Remember, each flow point is a statement about what *would happen* if the susceptibles and the infectives were at the given level. 

```{r sir1-5, echo=FALSE, results="markup"}
askMC(
  "Suppose the epidemic starts with about 800 susceptibles and 10 infectives. Follow the flow to work out the long-term result. Which of these is the best description?",
  "The infection will dissipate with only a handful of people getting sick.",
  "About 250 people will get sick." = "Do you mean that at the end, only 250 people will still be susceptible? If so, then 800-250 people got sick.",
  "About 600 people will get sick.",
  "Everybody will get sick.",
  random_answer_order = FALSE
)
```
  
Let's focus on a small area of the state space:

```{r echo=FALSE, out.width="400pt", out.height="400pt"}
SIR_flow(dom = domain(I=c(0,20), S=c(450,550)), nsteps=25)
```

```{r sir1-6, echo=FALSE, results="markup"}
askMC(
  "Is there a fixed point in the dynamics?",
  "None is visible in this region.",
  "+Yes, at about $I=10, $S=500$+",
  "Yes, at $I = 0$"
)
```

```{r sir1-7, echo=FALSE, results="markup"}
askMC(
  "Look carefully at the flow near $I=0$ and $S=500$. Which of these is the best description?",
  "+The number of susceptibles increases over time, but the number of infectives stays near 0.+",
  "Steady state. The population doesn't change at all.",
  "The number of susceptible eventually grows to the point where the epidemic takes off again." = "It looks like that's right, but you are looking at the big picture. The trick in making differential equation models is to make your focus local. Let the global picture emerge from the mechanism (say, Euler steps) for solving the dynamics."
)
```

The fixed point near $I=10$, $S=500$ has a biological description: the disease has become **endemic**, meaning that it persists in the population. People who get sick recover, but just enough new people get sick to balance out that recovery. In endemic disease, it may be that just about everyone gets the disease at some point, or it may be that a substantial fraction of the population never gets it. To answer that, we'll have to look more closely at the rate of new infections, that is, $\dot{I}$.
  
---

Building a model

We want to translate these "biological" facts into a mathematical model.

- If there are very few infectives, there will be (at least initially) very few susceptibles becoming infective.
- Infectives "recover" or, more precisely, are removed from the infective pool at a given rate.
- If there are more susceptibles, at any given number of existing infectives, there will be more new infectives created .
- If there are more infectives, at any given number of susceptibles, there will be more new infectives

Sdot = new arrivals - getting sick
Idot = +getting sick - recovering

- new arrivals: This is a model of infectious disease spread, so keep things very simple: a constant rate of new arrivals, n

- recovering: People are sick for a few days, then get better. Why we use exponential dynamics for this and how the parameter is related to the average time people are sick. We're not thinking that infectiousness depends on how long you've been infective except that eventually you recover. We just need something that works in bulk, not much changes over the duration of one person's infectiousness.  So -a I

- getting sick. We want this to increase with both S and I. Should they add? No, because if I is small, it doesn't matter how big S is, and vice versa. So b SI.



