---
title: "Following SIR"
author: "Daniel Kaplan"
runtime: shiny_prerendered
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

The SIR model of epidemics has these dynamics, where the state is the number $S$ of susceptibles and the number $I$ of infectives at any given time:
$$ \frac{d}{dt} S = - a S I \ \ \ \ \frac{d}{dt} I = a S I - b I  .$$

```{r echo=FALSE}
epi <- integrateODE(
    dS ~ -a*S*I, dI ~ a*S*I - b*I,
    a=0.001, b=0.5, S=999, I=1, tdur=40 )
```


Here is a trajectory of an epidemic. Note that neither axis is time. Instead, the axes are the state variables $I$ and $S$.
```{r echo=FALSE}
Traj <- tibble::tibble(
  time = seq(0,40,length=1000),
  I = epi$I(time),
  S = epi$S(time)
)

gf_line( I ~ S, data = Traj, size=2) %>% 
  gf_lims(x=range(0,1100), y=range(0,200)) %>%
  gf_labs(x="Susceptibles (# of people)",
          y="Infectives (# of people)")
```


And here are the solutions that correspond to the trajectory:
```{r echo=FALSE, fig.keep="hold"}
gf_line( S ~ time, data=Traj, size=2) %>%
  gf_lims(x=range(0,40), y=c(0,1050)) %>%
  gf_labs(x="Time (days)", y="S and I") %>%
  gf_line( I ~ time, color="orange3", size=3)
```



a. Draw an arrowhead on the trajectory to indicate which way the state is moving as time increases.
#. Label each of the graphs in the solution plot.
#. Estimate the location of the state at time 10 days and mark this on the trajectory plot.
#. Estimate the size of this population at the start of the epidemic.
#. How many people were never infected during this epidemic?
#. (DIFFICULT) Given the curves in the solution plot, explain how to use calculus operations and arithmetic to estimate the values of parameters $a$ and $b$ in the dynamical equations.  You don't have to make the estimate, just describe how you would make it.
#. (EXTRA CREDIT) Estimate the parameter $a$ from the given data.
