---
title: "Exponentials"
author: "Daniel Kaplan"
---

```{r include=FALSE, context="setup"}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

Some years ago, you learned to calculate the slope of a straight-line function. If you had a graph, you would mark out a **run** and the corresponding **rise**, then compute rise/run. If you had a formula definition, e.g. $f(x) \equiv 3 - 4 x$, you would simply read the slope off as the **coefficient** on the function input. Here, the input is named $x$ and its coefficient in the formula is $-4$. 

In this question, we ask you to estimate the slope from a graph of the function. But the function is **exponential**, so not a straight line. 

A fundamental idea in calculus is that even a function with a curved graph, if you zoom in closely around a given point, will look like a straight line. And you know how to calculate the slope of a straight line.

When the graph is curved, the slope will be different at different points along the graph. So there's not a single slope for the function. Still, we can talk about the "slope at a point."

One way to specify a point on a function's graph is to give the horizontal coordinate: the **input** to the function. But here we're going to give you the **output** of the function.  

Estimate the slope of the exponential function $g(x) \equiv e^x$ at several inputs, which we'll call $x_1$, $x_2$, $x_3$ and $x_4$. We won't give you numerical values for the $x_i$ points, but we will tell you the output of the function at each of those inputs. the values of $x$ where:

a. $g(x_1) = 1$
b. $g(x_2) = 5$
c. $g(x_3) = 10$
d. $g(x_4) = 0.1$

The code sandbox below already contains an R command to plot out the mathematical exponential function over the domain $-3 \leq x \leq 3$. For each of (a)-(d), modify the command to zoom in on the domain around around the appropriate value of $x_i$.  Then calculate the slope  of  the curve at that $x_i$.

```{r daily-digital-02-sandbox1, exercise=TRUE, exercise.cap="Zooming in on exp(x)", exercise.eval=FALSE, exercise.lines=4, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE}
slice_plot(exp(x) ~ x, domain(x = c(-3, 3)))
```

```{r daily-digital-02-QA1, echo=FALSE}
# indicate correct choices with +_+ in the name of the list item.
etude2::etudeQ(
  prompt = "Using your answers for the slope at the points given in (a)-(d), choose the best answer to this question: What is the pattern in the slope as $x$ varies?",
    "+The slope at each value $x_i$ is the same as $e^{x_i}$.+" = "This property of the exponential function becomes important when describing a wide range of phenomena, from nuclear isotope decay to population growth.",
    "The slope at each value $x_i$ is the same as $x_i$." = "That would being saying the slope at $x_3$ is $\\approx 2.30$. Is that what you got?",
    "The slope at each value of $x_i$ is the same as $x_i^2$." = "That would being saying the slope at the $x_3$ is $\\approx 5.30$. Is that what you got?",
    "The slope at each value of $x$ is the same as $\\sqrt{x}.$" = "That would being saying the slope at the $x_i$ is $\\approx 1.52$. Is that what you got?"
)
```

Now that you've found the answer and seen how to specify the domain on a slice plot, here's a technique that can make your work a bit easier. Add this to the command that plots $e^x$ on the domain $-3 \leq x \leq 3$

    %>% plotly::ggplotly()

Make sure the `%>%` is on the *same line* as the `slice_plot()` expression itself. Run the amended command then hover the cursor over the graph.

Reflection: One goal of this problem is to introduce you to a way of writing about functions that is somewhat different than you may have seen before. 

1. Instead of using the letter $y$---as in $y = mx + b$, we give each function a name and, in the parentheses after the name, list the inputs to the function. For instance, we use $f(x)$ and $g(x)$ in this problem. 
2. We are not using the equal sign ($=$) to give a name to the function; we use $\equiv$. As you know, $=$ means "equality." But $h(x) \equiv x^2$ describes a **definition**: "we define $h$  to be the name of a function of $x$ with the formula $x^2$." Definitions are different from **equations**. Equations have an important role in mathematics, so it's worth not confusing them with definitions. 
3. We will often use letters from the end of the alphabet---$x$, $y$, $z$, $t$, $u$, $v$, $w$---as the **names** of inputs to functions. It will be important for us to give names to inputs because we will often deal with functions with **more than one input** and we'll need to keep straight which one stands for what.
4. A function is a kind of machine that takes inputs and produces outputs. Similarly, a printer is a machine that takes several inputs---paper, ink, electricity, PDF file---and produces the familiar kind of output. Printers are designed to have different places for the different inputs, so you can't mistake one for the other. With mathematical functions, all we have for this purpose is the different **names** for the inputs.
5. The input slot on a printer is often called the "paper tray." You would never confuse the paper tray with paper itself. Inputs to function are like the paper tray. When we put a particular value into the input slot $x$ of a function, we will often call the particular value something like $x_0$ or $x^\star$, or $x_3$. When you see something like $x_0$, remember that it is a particular piece of paper, not the input tray for paper. We'll even sometimes give you the output from the function---as we did here with the exponential function---and ask you to figure out the particular input that corresponds to that output. This is like asking, "Find a piece of paper that, once it's printed on, the ink won't show through to the other side."

There are good reasons why to start to make simple distinctions that might have been ignored or glossed over in high-school math. 
