---
title: "MMAC 1-3"
---

```{r include=FALSE, context="setup"}
library(etude2)
```


**Exercises 1-7 on paper**

Do textbook exercises 1-7 in ยง1.3 on paper. Hold on to the paper so you can hand it in at the appropriate time.


**Exercises 55-58**

Consider total United States health expenditures as a percentage of gross domestic product (GDP) each year.

Year     |   Percent_GDP
---------|---------------
2009     | 17.71
2010     | 17.66
2011     | 17.68
2012     | 17.91


The linear function

$$f(x)  \equiv 0.062 x - 106.911$$
is a reasonable straight-line model of this data set. Imagine that the domain of $f()$ is the whole real line. State the value requested and explain the meaning of the quantity.


```{r mmac-1_3-Essay3, echo=FALSE}
etude2::etudeEssay(
  prompt = "55) The value $f(2013)$  and its interpretation"
)
```

```{r mmac-1_3-Essay4, echo=FALSE}
etude2::etudeEssay(
  prompt = "56) Slope of $f()$ with respect to $x$ and its interpretation."
)
```

```{r mmac-1_3-Essay5, echo=FALSE}
etude2::etudeEssay(
  prompt = "57) Vertical intercept  of $f()$ and its interpretation."
)
```

```{r mmac-1_3-Essay6, echo=FALSE}
etude2::etudeEssay(
  prompt = "58) Horizontal intercept of $f()$ and its interpretation."
)
```

```{r mmac-1_3-Essay7, echo=FALSE}
etude2::etudeEssay(
  prompt = "Reality check) Are any of the answers to 55-58 quantities that would not be taken seriously by a historian of  the US economy?"
)
```

**MMAC Exercises 67-74**


Refer to the 8 function definitions on page 51 at the top of the right column. Each of the functions is piecewise linear. Over each piecewise domain the function is either increasing, constant, or decreasing with respect to input $x$.  Select the right answer for each of these domain pieces.



```{r mmac-1_3-QA10, echo=FALSE}
etude2::etudeQ(
  prompt = "The function in question 68 on the domain piece $x < 0$.",
  "+increasing+", "constant", "decreasing",
  random_answer_order = FALSE
)
```

```{r mmac-1_3-QA11, echo=FALSE}
etude2::etudeQ(
  prompt = "The function in question 69 on the domain piece $x \\leq -2$.",
  "increasing", "constant", "+decreasing+",
  random_answer_order = FALSE
)
```

```{r mmac-1_3-QA12, echo=FALSE}
etude2::etudeQ(
  prompt = "The function in question 70 on the domain piece $3  <  x$.",
  "increasing", "constant", "+decreasing+",
  random_answer_order = FALSE
)
```

```{r mmac-1_3-QA13, echo=FALSE}
etude2::etudeQ(
  prompt = "The function in question 71 on the domain piece $x  \\leq 8$.",
  "increasing", "+constant+", "decreasing",
  random_answer_order = FALSE
)
```

```{r mmac-1_3-QA14, echo=FALSE}
etude2::etudeQ(
  prompt = "The function in question 73 on the domain piece $0 < x < 5$.",
  "increasing", "constant", "+decreasing+",
  random_answer_order = FALSE
)
```
