---
title: "Allometrics"
author: "Danny Kaplan"
date: 2020-06-03 
tags: [power-law, allometrics, third]
id: crow-mean-dress
global-id: "KGYKY"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```

The surface area $S$ of a mammal is reasonably well approximated by the function $$S(M) \equiv k M^{2/3}$$ where $M$ is the body mass (in kg) and the constant $k$ depends on the particular species under consideration. 

Note that $M^{2/3}$ is **not an allowed arithmetic operation**. $[M] = \mbox{mass}$, and mass cannot be raised to a non-integer power. More properly, the expression should be written $$\left(\frac{M}{1\  kg}\right)^{2/3}$$ The division by "1 kg" renders dimensionless the quantity in the parentheses: $$\left[\frac{M}{1\  kg}\right] = 1$$ 
In order to render the quantity both dimensionless and *unitless*, $M$ should be specified in kg. The usual practice is to skip the "1 kg" business and simply say, "Where $M$ is in kg." You will see such notation frequently in your career and should take care to use the indicated units.

You'll need to open a computing sandbox to do the calculations.

```{r crow-mean-dress-1, echo=FALSE, results="markup"}
askMC(
  prompt = "A) Consider a baby and an adult. The adult's mass is $8$ times greater than the baby's. Then the adult's surface area is ...?",
  "The same as the baby's", "1.5 times of the baby's", "+4 times the baby's+", "8 times the baby's",
  random_answer_order = FALSE,
  inline = FALSE
)
```




```{r crow-mean-dress-2, echo=FALSE, results="markup"}
# indicate correct choices with +_+ in the name of the list item.
askMC(
  prompt = "B) Consider a human of body mass 70 kg with a skin surface area of 18,600 cm^2^. Which of the following units for the constant of proportionality $k$ is correct?",
    "cm$^2$ kg$^{-2/3}$" = "Kilograms to a fractional power is not a sensible unit.",
    "+cm$^2$+" = "",
    "cm$^2$ kg$^{2/3}$" = "When you multiply $(70 kg/kg)^{2/3}$ by $k$, you need to get a result in $cm^2$.",
    "kg$^{-1}$" = "Would this produce $cm^2$ for the result?"
)
```

```{r crow-mean-dress-3, echo=FALSE, results="markup"}
askMC(
  prompt = "C)  In the units of part (B), which value is $k$ closest to?",
  1, 10, 100, "+1000+",
  random_answer_order = FALSE
)
```

The numerical value of the constant $k$ changes depending on what units you want to express it in. The value you found in part (C) works for masses stated in kg and skin areas in cm$^2$. Suppose you want to present $k$ in units that will make sense to people who are used to talking about skin area in square inches and mass in pounds. 1 inch $\approx$ 2.6 cm.  In the rough-and-ready way everyday people express themselves, 1 kg $\approx$ 2.2 lbs. (Of course, pounds is a measure of force, not mass. But people use it as if it were mass. A mass of 70 kg corresponds to about 4.8 slugs. In Earth's gravity, 4.8 slugs produces a force of 154 pounds.)

We have to be very careful in dealing with the $\left(\frac{M}{1 kg}\right)^{2/3}$. Plugging in $M=154$ lbs makes this 
$$\left(\frac{154\ \text{pounds}}{1\ \text{kg}}\right)^{2/3}$$
You can't take (pounds)$^{2/3}$ or (kg)$^{2/3}$; you won't get a sensible unit in either case. But [pounds/kg] = [1], so taking the two-thirds power of the ratio is perfectly legitimate. 

Still, there's a problem. Multiplying $k\ 154^{2/3} (\text{pound}/\text{kg})^{2/3}$ has the right dimension, but strange-looking units.

The resolution to this paradox is to multiply $\frac{154\ \text{pounds}}{1\  \text{kg}}$ by an appropriate flavor of 1 to render the dimensionless quantity unitless as well. This factor will be $\frac{1 \text{kg}}{2.2 pounds}$, giving 

$$\left(\frac{154}{2.2}\right)^{2/3} = 0.591 \times 154^{2/3}$$

To find a constant of proportionality $k_{lbs}$ that we can simply multiply 

```{r crow-mean-dress-5, echo=FALSE, results="markup"}
askMC(
  prompt = "Optional challenge) What is the rough numerical value of the conversion factor which would produce a constant of proportionality that would be suitable for converting body mass in pounds to skin area in square-inches?",
  0.03, 
  "+0.3+" = "$k$ is about 1000 cm$^2$ kg$^{-2/3}$. To convert the cm to inches, we need to multiply by (1 in/2.6 cm)$^2$. To convert the kg to pounds, we need to multiply by (2.2 pounds/ 1 kg)$^{-2/3}$.",
  3, 30, 300,
  random_answer_order = FALSE
)
```




