---
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
etude2::load_learnr_safely()
library(gradethis)
library(mosaic)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```

```{r context="setup"}
## SET UP THE PROBLEM HERE. 
# Note use of double \\ for latex in quotation marks
problemLatex <- "\\int  A \\sin(Bx + C) + D\\ dt?"
answer_fun <- function(t,x,A,B,C,D) t*(A*sin(B*x + C) + D)
evaluate_at <- 1:10
correct_arguments <- c("x","t", "A", "B", "C", "D")
## Remember to set the CHUNK NAMES to match the file name
```

Work out this anti-derivative on paper.

<div style="font-size: 40px; text-align:center;">$`r problemLatex`$ </div>

Once you have your result, translate it into a correct R *function*. 

```{r IOD-H, exercise=TRUE, exercise.lines=5, exercise.cap="Integral of the day"}
F <- makeFun(  ~ x)
```

```{r IOD-H-check}
library(magrittr)
gradethis::grade_result(
  fail_if( ~ !is.function(.result), message = "Your result should be in the form of a function."),
  fail_if( ~ !"t" %in% names(formals(.result)), 
           message = "Shouldn't t be involved in this?"),
  fail_if( ~ !all(correct_arguments %in% names(formals(.result))),
           message = "You don't seem to have the right argument name."),
  pass_if( ~ all(.result(t=1:10, x=5, A=1,B=2,C=3, D=4) ==
                          answer_fun(t=1:10,x=5, A=1,B=2, C=3, D=4))),
  fail_if( ~ TRUE, message="Correct arguments, but not the right function.")
)  
```


Now that you have the correct answer, use Latex to typeset the formula in the function here. Some Latex tips: `\sin()` will produce $\sin()$. Note that the backslash at the start of `\sin()` does something. Without the backslash, that is `sin()`, you get $sin()$. If you don't see the difference, don't worry about it. But fastidious typesetters (like the author) are obsessive about such things.

```{r results="asis"}
etude2::latex_entry_box("latex-H")
```

```{r context="server"}
id <- "latex-H"
answers <- c("t(A \\sin(Bx + C) + D)",
             "(A \\sin(Bx + C) + D)t",
             "t(A sin(Bx + C) + D)",
             "(A sin(Bx + C) + D)t",
             "A sin(Bx + C)t + Dt",
             "A \\sin(Bx + C)t + Dt") 
squash_str <- "\\{|\\}| |\t|\n"

# To return the state of the app
if (!exists("apps_state")) apps_state <- reactiveValues() 

output[[NS(id, "rendered")]] <- renderUI({
  withMathJax(paste0("$$", input[[NS(id, "latex")]], "$$"))
})
output[[NS(id, "feedback")]] <- renderUI({
  compare <- gsub(squash_str, "", answers)
  the_answer <- input[[NS(id, "latex")]]
  if (is.null(the_answer)) the_answer <- "missing"
  correct <- gsub(squash_str, "", 
           input[[NS(id, "latex")]]) %in% compare
  if (exists("apps_state")) {
    registration_name <- NS(id, "result")
    apps_state[[registration_name]] <- 
      tibble::tibble(id = registration_name, 
                     question = "latex input",
                     answer = the_answer,
                     time = as.character(Sys.time()),
                     correct = correct,
                     attempts = 1,
                     history = "")
    
  }
  ifelse(correct, "✓", "☞")
})
```

