---
title: "Random walks"
author: "Daniel Kaplan"
runtime: shiny_prerendered
---

```{r include=FALSE, context="setup"}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(viridis)
```

The solution to diffusion differential equation gives the concentration of the diffusing particles $C(x, t)$ at any $x$ and $t$. At any given value of $t= t_1$, the shape of $C(x,t=t_1)$ tends to a smooth bell-shaped curve: the gaussian function. The width of the gaussian is a function of $t$. Conventionally, the width corresponds to the variance parameter $\sigma^2$. This changes with $t$ in a very simple way: $$\sigma(t)^2 = D t$$

Everything is smooth and nicely behaved. But this is an abstraction which summarizes the concentation of a theoretically infinite population of particles. Looking at the position of a single particle as a function of time gives a very different impression. The figure shows individually the trajectories of 50 (randomly selected) particles.

```{r echo=FALSE}
set.seed(102)
ntrajs <- 50
dt <- 0.05
step_size <- 9/sqrt(40/dt) # to match time t=40 in the other graphs
tpts <- seq(0,40, by=dt)
colors <- hcl.colors(ntrajs)
Trajs <- list()

for (k in 1:ntrajs) {
  Trajs[[k]] <- tibble::tibble(
    t = tpts,
    dx = rnorm(length(t), 0, step_size),
    x = cumsum(dx),
    color = colors[k]
  )
}

gf_line(x ~ t, data = bind_rows(Trajs), 
                     color = ~ color, alpha=0.25) %>%
  gf_line(x ~ t, data = bind_rows(Trajs[c(1, round(ntrajs/4), round(ntrajs/2))]), color=~color) %>%
  gf_refine(scale_color_identity())
  
```

Do take note that the position of each particle in the pipe at time $t$ is shown on the *vertical axis*.

Each of the trajectories is called a **random walk**, as if a walker were randomly taking steps forward or backward.

