---
title: "Daily Digital 01"
subtitle: "Math 142Z"
date: "Due 11 Jan 2021"
output: learnr::tutorial
runtime: shiny_prerendered
tutorial:
  version: 0.1
tags:
- vectors
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
```

```{r context="server"}
apps_state <- reactiveValues()
```


```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD-01") # Add the DD number
etude2::etude_document_name("DD-142Z-01") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
# Leave this safety check here
if (etude2::etude_document_name() == "DD-") 
  stop("You must set the master_id() and the document name specifically and uniquely for this document")
```

## A new Daily Digital system

```{r child="new-DD-system.Rmd"}
```

## Basic modeling functions

```{r child="basic-modeling-functions.Rmd"}
```

## Drill

You can reset the drill counter to start over at any time. When you can get 8 out of 10 correct, you can consider the drill complete. You'll have to hand in your work during *this session* to get credit. If you come back to the Daily Digital document at another time, the drill will have reset to 0/0, but your earlier submitted scores will still be in our records.

```{r child="Drill-1.Rmd"}
```



## Integral of the day

```{r child="IOD-A.Rmd"}
```

## Interior Functions

Very often, functions used in modeling have a basic form, e.g. $\exp()$ or $\sin()$ composed with an *interior function*. For example, in $\exp(kx)$ the interior function is $k x$, while in $\sin(\omega x + \phi)$ the interior function is $\omega x + \phi$. Our basic modeling functions often have an interior function of the form $ax + b$, which is to say, a straight-line function. Functions with a straight-line interior function (e.g. $e^{2x+3}$ and $e^{-x}$ are much easier to anti-differentiate than functions with a nonlinear interior function (example: $\sin(x^2)$). 

```{r child="interior_funs-1.Rmd"}

```

## Energy-Maneuverability Theory

```{r child="Boyd-1.Rmd"}

```

## Parent, child, grandchild

```{r child="chain-of-differentiation.Rmd"}

```


## Hand in your work

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>

```{r child=system.file("submission-handler.Rmd", package="etude2")}

```


-----

Location: `r knitr::current_input(dir = TRUE)` `r etude2::etude_document_name("142Z-DD-01.Rmd")`

