---
title: "Compose Latex Math"
date: "MMM DD YYYY"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r include=FALSE}
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(shiny)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60,
                 exercise.checker = gradethis::grade_learnr)
```

One of the challenges of calculus is making sense of mathematical notation. This requires careful reading, which is a skill you can develop. To that end, Daily Digitals will include typesetting challenges. These require you to read the traditional notation, take it apart, and re-assemble it using a famous and widely used notation called "Latex."

Latex uses ordinary typewriter characters to describe the mathematical expression. There are  Latex commands for special symbols. For instance, `\beta` produces the letter $\beta$ and `\sum` produces the $\sum$ symbol. Today's challenge introduces subscripts, superscripts, and fractions.

- Superscript: Example: `\beta^4` produces $\beta^4$. 
- Subscript: Like superscript but use `_` (underscore) instead of `^` (caret).
- Fractions: Example: `\frac{a + b}{c + d}` produces $\frac{a+b}{c+d}$


```{r echo=FALSE}
render_latex_ui <- function(id){
HTML(glue::glue('<div style="font-size: 25px;">
  <div id={NS(id, "prompt")} class="shiny-html-output"></div>
</div>
<hr/>
<div style="font-size: 25px;">
  <div class="form-group shiny-input-container">
    <label class="control-label" for="latex">Type latex math here.</label>
    <textarea id={NS(id, "latex")} class="form-control" rows="5" cols="50" style="font-size: 25px"></textarea>
  </div>
</div>
<div id={NS(id, "rendered")} class="shiny-html-output" style="font-size:25px;"></div>
<div id={NS(id, "feedback")} class="shiny-html-output" style="color: green; font-size:30px;"></div>'))
}


```

```{r}
render_latex_ui("latex-A")
```

```{r context="server"}
id <- "latex-A"
answers <- c("b_0 \\frac{x^3}{3}") 
squash_str <- "\\{|\\}| |\t|\n"

output[[NS(id, "rendered")]] <- renderUI({
  withMathJax(paste0("$$", input[[NS(id, "latex")]], "$$"))
})
output[[NS(id, "prompt")]] <- renderUI({
  withMathJax(paste0("Typeset this: $$",answers[1],"$$"))
})
output[[NS(id, "feedback")]] <- renderUI({
  compare <- gsub(squash_str, "", answers)
  if (gsub(squash_str, "", 
           input[[NS(id, "latex")]]) %in% compare) "Success!"
  else ""
})


```


