---
title: "R Graphics"
author: "Daniel Kaplan"
---

<a href="https://maa-statprep.shinyapps.io/141Z-R_command_guide" target="_blank">*R Command Guide*</a>


Starting in this *Daily Digital* you will have some exercises that involve your creating R commands. Please refer to the <a href="https://maa-statprep.shinyapps.io/141Z-R_command_guide" target="_blank">*R Command Guide*</a> for an explanation of the specific commands we will use in this course. A good approach is to scan quickly through  the *Guide* to find a command that looks like what you need, and try it out yourself.


### Contour plots

A function $\text{twins}(x,y)$ is already provided to you as `twins()` in the code box below. Use the code box to create a contour plot of `twins()`.  The domain of the plot should be large enough to show a mountain next to  a deep hole.


```{r daily-digital-03-E1-setup,echo=FALSE, message=FALSE}
# Setup code goes here
twins <- mosaic::rfun(~ x + y, seed = 302, n=5)
```

Use `contour_plot()` to create the graphic. You'll see an example in the <a href="https://maa-statprep.shinyapps.io/141Z-R_command_guide" target="_blank">*R Command Guide*</a>.


```{r daily-digital-03-E1, exercise=TRUE, exercise.cap="Contour plot of twins(x,y)", exercise.setup="daily-digital-03-E1-setup", exercise.eval=FALSE, exercise.lines=3, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE, eval=FALSE}

```


```{r daily-digital-03-E1-check,eval=FALSE}
gradethis::grade_result(
  fail_if( ~ ! etude::is_ggplot(.result), "You should be making a plot."),
  fail_if( ~ ! etude::is_contour(.result), "You should make a contour plot."),
  fail_if( ~ max(abs(.result$data$x)) < 5, "Your x-domain isn't big enough"),
  fail_if( ~ max(abs(.result$data$y)) < 5, "Your y-domain isn't big enough"),
  pass_if( ~ TRUE)

)
```

### Slice plots

Continuing to examine the function $\text{twins}(x,y)$, take a **slice** through the function along the line $y=0$. Plot the function output versus $x$ using `slice_plot()`. Use the same $x$ domain as you did in  the contour plot.

```{r daily-digital-03-E2, exercise=TRUE, exercise.cap="Slice plot  of twins(x, y) along the slice y=0", exercise.setup="daily-digital-03-E1-setup", exercise.eval=FALSE, exercise.lines=2, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE, eval=FALSE}

```


```{r daily-digital-03-E2-check,eval=FALSE}
gradethis::grade_result(
  fail_if( ~ ! etude::is_ggplot(.result), "You should be making a plot."),
  fail_if( ~ ! etude::is_slice_plot(.result), "You should be using `slice_plot()`."),
  pass_if( ~ TRUE)
)
```

```{r daily-digital-03-Essay2, echo=FALSE}
etude2::etudeEssay(
  prompt = "Explain what you see in the slice plot shape that corresponds to what the contour plot has to say along the slice $y=0$."
)
```
