---
title: "MMAC ยง1.4 Exercises"
---

```{r include=FALSE}
library(etude2)
```

### MAC Exercises 38-40

Refer to the graphs on  MMAC page 68 in the right column.

For each of the following, determine whether the curve is exponential or not by determining if the curve exhibits a constant doubling or halving time, as appropriate. Imagine that the horizontal axis is labeled $t$, for time in seconds.

```{r mmac-1_4-QA15, echo=FALSE}
askMC(
  prompt = "For graph (39) the function  has ...",
    "+A doubling time of 1 second,+" = "",
    "A doubling time of 2 seconds." = "",
    "A halving time (half-life) of 1 second," = "",
    "A halving time (half-life) of 2 seconds," = "",
    "The time to double or halve changes depending on $t$, so the function is not exponential." = "",
  random_answer_order = FALSE
)
```

```{r mmac-1_4-QA16, echo=FALSE}
askMC(
  prompt = "For graph (40) the function  has ...",
    "A doubling time of 1 second." = "",
    "A doubling time of 2 seconds." = "",
    "A halving time (half-life) of 1 second" = "",
    "A halving time (half-life) of 2 seconds" = "",
    "+The time to double or halve changes depending on $t$, so the function is not exponential.+" = "",
  random_answer_order = FALSE
)
```

```{r mmac-1_4-QA17, echo=FALSE}
askMC(
  prompt = "For graph (41) the function  has ...",
    "A doubling time of 1 second." = "",
    "A doubling time of 2 seconds." = "",
    "+A halving time (half-life) of 1 second.+" = "",
    "A halving time (half-life) of 2 seconds." = "",
    "The time to double or halve changes depending on $t$, so the function is not exponential." = "",
  random_answer_order = FALSE
)
```

### MMAC Exercises 47 & 48

Refer to the graphs on MMAC page 69 in the left column. For each, use properties of exponentials to explain briefly why the curve is not exponential. Have in mind the mathematical concepts of "domain", "concave up", and "concave down."

```{r mmac-1_4-Essay8, echo=FALSE}
etude2::etudeEssay(
  prompt = "Why  is the function in graph (47) not exponential?"
)
```

```{r mmac-1_4-Essay9, echo=FALSE}
etude2::etudeEssay(
  prompt = "Why  is the function in graph (48) not exponential?"
)
```



