---
title: "Dynamical Systems: State and Dynamics"
author: "Daniel Kaplan"
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r include=FALSE}
etude2::load_learnr_safely()
library(etude2)
```

Differential equations provide an important type of model for **systems** that change in time.

A **system** is something made up of two or more interacting components. A dynamical system has a **state** and a **dynamical rule**. Typically the state is a set of numbers. In CalcZ, we consider states consisting of one number, and states consisting of two numbers. We call these *one-dimension* and *two-dimensional* systems respectively.

A **change of state** is a set of numbers, one for each component of the state, which tells how fast the state is changing. In other words, the change of state is the derivative with respect to time of the state.

A **dynamical rule** consists of one function for each dimension of state. The function takes the state as input, and returns the **change-of-state with respect to time**.

A **differential equation** is a kind of sentence saying, "This is the state and dynamical rule for a dynamical system." The grammar of a differential equation is 

> [Derivative of state with respect to time] = [Dynamical function(s)] 

The two main notations we use for "derivative with respect to time" use $\partial_t$ as in $\partial_t y$ or the dot $\dot{ }$ as in $\dot{y}$. There's no fundamental difference between the two, the dot is more common in physics and engineering.

Here are six examples, labeled (A) through (F).

A) $\partial_t x = b\, \sin(x)$

B) $\dot{x} = r\, x$

C) $\dot{x} = r\, x\, (1 - x/K)$

D) $\left\{\begin{array}{ll}\dot{x} = 2\, x -3\, y\\\dot{y} = 1\, x + 2\, y \end{array}\right.$

E) $\dot{\mathbf X} = {\mathbf A} {\mathbf X}$, for instance 
${\dot{\mathbf X}} = \left[\begin{array}{cc}a & b\\c & d\end{array}\right] {\mathbf X}$

F) $\left\{\begin{array}{ll}\dot{S} = -\beta\, S\, I\\ \dot{I} = \beta\, S\, I - \alpha\, I\\\dot{R} = \alpha\, R\end{array}\right.$

It's helpful to have a systematic process for reading a differential equation.

**Step 1**) Identify the symbol(s) representing the *state* and what is the dimension of the state.

- Equations (A-C): The state is $x$ and has one dimension.

- Equation (D): There are two derivatives on the left-hand side of the equation, $\dot{x}$ and $\dot{y}$. This indicates that the state is the pair $(x, y)$, a two-dimensional state. 

- Equation (E): The very compact form $\dot{\mathbf X} = {\mathbf A} {\mathbf X}$ is using vector/matrix notation (the **symbols in bold face**). We cannot tell yet what the dimension of vector $\mathbf X$ is. The example ${\dot{\mathbf X}} = \left[\begin{array}{cc}a & b\\c & d\end{array}\right] {\mathbf X}$ shows the shape of the matrix $\mathbf A$. Here we can see that the state must be two-dimensional. 
Is the state one-dimensional or two-dimensional?

```{r dst-1, echo=FALSE, results="markup"}
etude2::etudeQ(
  "What symbol(s) represent the state in Equation (F)?",
  "$x$ and $y$" = "Do you see an $x$ or $y$ anywhere in equation (F)?",
  "Just $y$",
  "$S$ and $I$" = "But there are three equations.",
  "+$S$, $I$, and $R$+"
)
```


```{r dst-2, echo=FALSE, results="markup"}
etude2::etudeQ(
  "What is the dimension of the state in Equation (F)?",
  1, 2, 3, 4, "Not enough information",
  right_one = 3, random_answer_order = FALSE
)
```

**Step 2**) Identify the dynamical function (s). These will always be functions of the state and may include *parameters*, which are simply symbols standing in for numerical quantities.

Equation A) The state is $x$ and the dynamical function is $a \sin(x)$. Since $a$ is not part of the state, it must be a parameter.

Equation B) The state is $x$ and the dynamical function is $a x$. That's a very simple function, but it's also very common in models.

Equation C) The state is $x$. The dynamical function $r x (1 - x/K)$ has two parameters: $r$ and $K$.

```{r dst-3, echo=FALSE, results="markup"}
etude2::etudeQ(
  "How many dynamical functions are there in Equation (D)? (Hint: What is the dimension of the state in Equation (D)?)",
  1, 2, 3, 4, "Not enough information",
  right_one = 2,
  random_answer_order = FALSE
)
```

In vector/matrix notation such as $\dot{\mathbf X} = {\mathbf A} {\mathbf X}$ , it can seem that there is only one one function. Here, that's ${\mathbf A} {\mathbf X}$. But remember, such multiplication of a matrix times a vector amounts to a *vector of functions*, as in Equation (D).


```{r dst-4, echo=FALSE, results="markup"}
etude2::etudeQ(
  "In Equation (F) there are three dynamical functions, one of each of the components of the state. Which of these is **not** one of those functions?",
  "$-\\beta S I$",
  "$\\beta S I - \\alpha I$",
  "$\\alpha I$", 
  "+$-\\beta S I + \\alpha I$+"
)
```

-----


```{rrrr dst-4, echo=FALSE, results="markup"}
etude2::etudeQ(
  
)
```
