---
title: "Integral of the day: TEMPLATE"
date: "MMM DD YYYY"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
etude2::load_learnr_safely()
library(gradethis)
library(mosaic)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```

```{r context="setup"}
## SET UP THE PROBLEM HERE. 
# Note use of double \\ for latex in quotation marks
problemLatex <- "\\int  c_0 + c_1(x-x_0) + c_2 (x-x_0)^2\\ dx?"
answer_fun <- function(x,x0,c0,c1,c2) c0 * x + c1* (x-x0)^2 / 2 + c2*(x-x0)^3/3
evaluate_at <- 1:10
correct_arguments <- c("x","c0", "c1", "c2", "x0")
## Remember to set the CHUNK NAMES to match the file name
```

Work out this anti-derivative on paper.

<div style="font-size: 30px; text-align:center;">$`r problemLatex`$ </div>

Once you have your result, translate it into a correct R *function*. Use `c0` as the R name for $c_0$, `c1` for $c_1$, `c2` for $c_2$, and `x0` for $x_0$. Although the constant of integration $+C$ is properly a part of the definition, don't include it in your R function. This is basically the same as selecting $C=0$.

```{r IOD-G, exercise=TRUE, exercise.lines=5, exercise.cap="Integral of the day"}
F <- makeFun(  ~ x)
```

```{r IOD-G-check}
library(magrittr)
gradethis::grade_result(
  fail_if( ~ !is.function(.result), message = "Your result should be in the form of a function."),
  fail_if( ~ !all(correct_arguments %in% names(formals(.result))),
           message = "You don't seem to have the right argument name."),
  pass_if( ~ all(.result(x=1:10, x0=5, c0=1,c1=2,c2=3) ==
                          answer_fun(x=1:10,x0=5, c0=1,c1=2,c2=3))),
  fail_if( ~ TRUE, message="Correct arguments, but not the right function.")
)  
```


Now that you have the correct answer, use Latex to typeset the formula in the function here. Some Latex tips: `c_0` will produce $c_0$. You do not need to do the LaTeX code until you see the green checkmark. If your integral was validated by the codebox above, then you have the correct answer, we now want you to explore formatting correctly using LaTeX. The system to provide you with a green checkmark looks for the answer the way your instructors would type it. In this case, write your anti-derivative in the same order of coefficients as used at the top of this problem and use $C$ at the end if you want to include an explicit constant of integration. If there is a fraction multiplying a coefficient, write it like `\frac{1}{5} c_5` which would give $\frac{1}{5} c_5$. (Just an example; the actual problem doesn't have any $c_5$ in it.)

```{r results="asis"}
etude2::latex_entry_box("latex-G")
```

```{r context="server"}
id <- "latex-G"
answers <- c("c_0 x + c_1(x-x_0)^2/2 + c_2(x-x_0)^3/3", 
             "c_0 x + \frac{1}{2} c_1(x-x_0)^2 + \frac{1}{3} c_2(x-x_0)^3",
             "c_0 x + c_1(x-x_0)^2/2 + c_2(x-x_0)^3/3 + C", 
             "c_0 x + \frac{1}{2} c_1(x-x_0)^2 + \frac{1}{3} c_2(x-x_0)^3 + C") 
squash_str <- "\\{|\\}| |\t|\n"

# To return the state of the app
if (!exists("apps_state")) apps_state <- reactiveValues() 

output[[NS(id, "rendered")]] <- renderUI({
  withMathJax(paste0("$$", input[[NS(id, "latex")]], "$$"))
})
output[[NS(id, "feedback")]] <- renderUI({
  compare <- gsub(squash_str, "", answers)
  the_answer <- input[[NS(id, "latex")]]
  if (is.null(the_answer)) the_answer <- "missing"
  correct <- gsub(squash_str, "", 
           input[[NS(id, "latex")]]) %in% compare
  if (exists("apps_state")) {
    registration_name <- NS(id, "result")
    apps_state[[registration_name]] <- 
      tibble::tibble(id = registration_name, 
                     question = "latex input",
                     answer = the_answer,
                     time = as.character(Sys.time()),
                     correct = correct,
                     attempts = 1,
                     history = "")
    
  }
  ifelse(correct, "✓", "☞")
})
```

