---
title: "Integral of the day: TEMPLATE"
date: "MMM DD YYYY"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r context="setup", include=FALSE}
etude2::load_learnr_safely()
library(gradethis)
library(mosaic)
library(mosaicCalc)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```

```{r context="setup"}
## SET UP THE PROBLEM HERE. 
# Note use of double \\ for latex in quotation marks
problemLatex <- "\\int \\frac{1}{x^2} dx"
answer_fun <- makeFun(-1/x ~ x)
evaluate_at <- 1:10
correct_arguments <- c("x")
## Remember to set the CHUNK NAMES to match the file name
```

Work out this anti-derivative on paper, translate it to R, and enter it in the sandbox

<div style="font-size: 40px; text-align:center;">$`r problemLatex`$ </div>


```{r IOD-F, exercise=TRUE, exercise.lines=5, exercise.cap="Integral of the day"}
F <- makeFun()
```

```{r IOD-F-check}
library(magrittr)
gradethis::grade_result(
  fail_if( ~ !is.function(.result), message = "Your result should be in the form of a function."),
  fail_if( ~ !all(correct_arguments %in% names(formals(.result))),
           message = "You don't seem to have all the right argument names."),
  pass_if( ~ all(.result(evaluate_at) ==
                          answer_fun(evaluate_at))),
  fail_if( ~ TRUE, message="Correct arguments, but not the right function.")
)  
```


Now that you have the correct answer, use Latex to typeset the formula in the function here. Some Latex tips:  `x^{-2}` will produce $x^{-2}$.

```{r results="asis"}
etude2::latex_entry_box("latex-F")
```



```{r context="server"}
id <- "latex-F"
answers <- c("-1/x", "- \frac{1}{x}", "- x^{-1}") 
squash_str <- "\\{|\\}| |\t|\n"

# To return the state of the app
if (!exists("apps_state")) apps_state <- reactiveValues() 

output[[NS(id, "rendered")]] <- renderUI({
  withMathJax(paste0("$$", input$`latex-F-latex`, "$$"))
})
output[[NS(id, "feedback")]] <- renderUI({
  compare <- gsub(squash_str, "", answers)
  the_answer <- input$`latex-F-latex`
  if (is.null(the_answer)) the_answer <- "missing"
  correct <- gsub(squash_str, "", 
           the_answer) %in% compare
  if (exists("apps_state")) {
    registration_name <- NS(id, "result")
    apps_state[[registration_name]] <- 
      tibble::tibble(id = registration_name, 
                     question = "latex input",
                     answer = the_answer,
                     time = as.character(Sys.time()),
                     correct = correct,
                     attempts = 1,
                     history = "")
    
  }
  ifelse(correct, "✓", "☞")
})
```

 

 



The Second Fundamental Theorem of Calculus says that if $f(x)$ is a smooth function, you can calculate $\int_a^b f(x) dx$ by finding an anti-derivative $F(x)$ and evaluating it at the bounds of integration as $F(b) - F(a)$. 

Using the Second Fundamental Theorem, compute the value of the definite integral $$\int_{-2}^4 \frac{1}{x^2} dx$$
This looks perfectly ordinary, but it is wrong. The `antiD()`-constructed anti-derivative gets it right:

```{r antiD, exercise=TRUE, exercise.cap="Using antiD().", exercise.nlines=4}
F <- antiD(1/x^2 ~ x)
F(4) - F(-2)
```
```{r iodwordbox, echo=FALSE, results="markup"}
etude2::etudeEssay("Why would you need to be careful about integrating this function? Hint: What's special about $f(x)$ somewhere on the interval [-2, 4] that tells you the fundamental theorem may not apply?")
```




