---
title: "Piston power"
author: "Transcribed by Daniel Kaplan"
runtime: shiny_prerendered
---

```{r include=FALSE}
library(etude2)
```

An important setting in thermodynamics is the compression and expansion of gasses in a closed container such as a piston/cylinder arrangement. If the piston is moved slowly, the temperature of the system will stay at its initial value. 

Mechanical work needs to be done to push the piston into the cylinder. The work done, for instance, to compress from an initial volume of 50cc to a final volume of 10cc is:
$$\int_{50}^{10} b \frac{T_c}{V}dV$$ where $b$ is a constant and $T_c$ is the initial temperature of the gas begin compressed.

On the other hand, in expanding back to its original volume, the system pushes back as it moves. The work done is
$$\int_{10}^{50} b \frac{T_e}{V}dV$$. 

When $T_e = T_c$, no net work is done during a compression/expansion cycle at a steady temperature. The work put into the system when compressing is the same as the work taken out of the system when expansion. The trick in getting the cylinder/piston to do net work is to compress at a relatively cool temperature (say 500K), then heat the gas (to, say 1000K) before starting the expansion. The total work done by the system will be $$\int_{10}^{50} b \left[\frac{T_e}{V} + \frac{T_c}{V}\right] dV$$

```{r piston1, echo=FALSE}
askMC(
  "Which of these expressions gives the net work done by the cylinder/piston system in one cycle?",
  "0" = "This would be true if $T_e = T_c$, but remember that we heat the gas before expansion.",
  "$b (T_e - T_c) \\left(\\frac{1}{50} + \\frac{1}{10}\\right)$" = "You forgot to take the anti-derivative before evaluating at the limits of integration.",
  "+$b (T_e - T_c) \\ln(5)$+",
  "$b \\ln(T_e - T_c) 40$" = "The integral is being taken with respect to $V$, not $T$.",
  "$b (T_e/T_c) \\ln(40)$",
  random_answer_order=FALSE
)
```

```{r piston2, echo=FALSE}
askMC(
  "Suppose the engine runs at 2000 rpm, that is, revolutions per minute. At this speed, the engine generates 1 kW of power. What are the dimensions of $b$?",
  "kW",
  "K/kW",
  "K",
  "+kW/K+")
```
