---
title: "Daily Digital 15"
subtitle: "Math 142Z"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
library(mosaicCalc)
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```

```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD-15") # Add the DD number
etude2::etude_document_name("DD-142Z-15") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
```

## Objectives

* Understand the difference between state, dynamics, trajectory 
* Understand the difference between continuous and discrete time
* [Recognize the form of] differential and finite-difference equations

## Step-by-step

```{r child="mechanics.Rmd"}
```

## Trajectories

```{r child="mechanics2.Rmd"}
```

## `Iterate()`

```{r child="mechanics3.Rmd"}
```

## Solutions

```{r child="Solutions.Rmd"}
```

## Greek ~~to~~ for you

```{r child="Greek.Rmd"}
```




## Hand in your work

<div style="color: red;">USAFA CalcZ students should hand in their work via GradeScope</div>

<!--

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>

```{rr echo=FALSE, messages=FALSE}
etude2::submissions_ui()
```

```{rr context="server", echo=FALSE, messages=FALSE}
etude2::submissions_logic()
```
-->

-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name()`

