---
title: "Mechanics of Iteration: Trajectories 2"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
library(math141Z)
library(etude2)
```

```{r include=FALSE, context="setup"}
library(mosaic)
library(mosaicCalc)
library(math141Z)
```

The ingredients for calculating a trajectory are:

a. A function $g(S)$
b. An initial condition ${\mathbf S}_0$.
c. A choice of how many steps to take.

<!--

There are two basic styles to computing trajectories on the computer. The first is a **calculator style* series of commands. For instance, the code in the sandbox computes a trajectory of length 6 from the given $g()$ and initial condition.

```{r fdec1, exercise=TRUE, exercise.cap="Calculator style", exercise.nline=10, eval=FALSE}
g <- makeFun(5 + x/2 ~ x)
x0 <- 7
x1 <- g(x0)
x2 <- g(x1)
x3 <- g(x2)
x4 <- g(x3)
x5 <- g(x4)
x6 <- g(x5)

```

Computer programmers learn that his is a bad style for several reasons.

i. The information about all these closely related things `x0`, `x1`, `x2` and so on is spread out over a bunch of different names. Most programming language provide no easy way to handle these as a group. An example of "handling as a group" is to plot out the trajectory versus $n$.  
ii. The statements are highly repetitive. Some people like this, because it gives them a sense of order. But programmers learn (through hard experience) that it is difficult for a human reader to confirm that every line is repeating exactly the same pattern. For instance, `x5 <- g(x3)` is a perfectly valid computer statement, but inconsistent with the dynamics that are supposed to be calculated.

A more effective style of programming is illustrated in the next sandbox which uses an "iteration operator":

-->

To facilitate calculating a trajectory, use the `Iterate()` operator. 

```{r fdec3-2, exercise=TRUE, exercise.cap="An iteration operator", exercise.nlines=8, eval=FALSE}
g <- makeFun(5 + x/2 ~ x)
Traj <- Iterate(g, x0 = 7, n = 6)
Traj
```

```{r fdec3-2-solution, echo=FALSE}
g <- makeFun(-0.8*x + 2 ~ x)
Traj <- Iterate(g, x0=10, 50) 
gf_point(x ~ n, data = Traj) %>%
  gf_line(alpha=0.2)
```

The `Iterate()` operator takes as a first argument the dynamics in the form of the name of a function or a tilde-expression of the sort accepted by `makeFun()`. The next argument is the initial condition. (This argument is named `x0` regardless of the name used in the dynamics).  The third argument is the number of steps to iterate.

`Iterate()` returns a data frame with two (or more) columns. The first, named `n`, is the step number. The second is given the same name as the variable used in the dynamical function. Because the output of `Iterate()` is in the form of a data frame, you can employ any function set up to accept a data frame. `gf_point()` is particularly useful for plotting the trajectory as a function of $n$.  

`Iterate()` is for finite-difference equations in discrete time, so $n$ will always be an integer and the trajectory is appropriately plotted as a series of isolated points. Even so, it can be helpful to the human viewer to connect the points faintly with straight lines. This makes it a bit easier to see the sequence. For instance:

```{r eval=FALSE}
# Make the Traj then ...
gf_point(x ~ n, data = Traj) %>%
  gf_line(alpha=0.2)
```

```{r fdec3-3, echo=FALSE, results="markup"}
askMC(
  "Use the sandbox above to construct 50 steps of the dynamics of $g(x) \\equiv -0.8 x + 2$ starting at $x_0 = 10$. Which of these best describes the graph $x_n$ versus $n$? (Hint: It's always best to try to figure out the command on your own. If you get stuck, press the 'Solution' button in the sandbox.)",
  "A steady exponential decay to about $x=3$.",
  "An oscillating exponential growth.",
  "+An oscillating exponential decay to about $x=1.1$.+",
  "A sine wave with period $P=\\pi$."
)
```


