---
title: "Integration by parts"
author: "Mike Brilleslyper"
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r include=FALSE}
library(gradethis)
library(etude2)
library(mosaic)
library(mosaicCalc)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```

So far in CalcZ we have developed methods for algebraically finding anti-derivatives for a small set of functions. These include basic modeling functions such as $$\int\cos(3x)dx = \frac{1}{3}\sin(3x) + C$$ and more complicated functions where we reverse the chain rule, such as $$\int x^2 \exp(x^3) dx = \frac{1}{3} \exp(x^3) + C .$$

With a bit more work, we can extend to a broader class of functions that do not fit into the two classes of examples above. For example, consider finding the anti-derivatives of the following functions:
$$f(x) \equiv x e^x, \ \ \ g(x)\equiv x \cos(x),\ \ \text{and}\ \ h(x) = x \ln(x)$$ None of these are basic modeling functions and none of these fit the "came from the chain rule" pattern.

Now, we're going to explore a “guess and correct” method for these types of anti-derivatives. It doesn't always work, but when it does it's impressive.

Note that each of the function above are products of $x$ with a basic modeling functions.  Finding the anti-derivatives of these functions will, in some sense, involve reversing the product rule. 

Consider $f(x) = x e^x$. This will seem to come out of the blue, but notice that $f(x) = x \partial_x(e^x)$. Our strategy has three steps. 

First, we will find a helper function by substituting $\partial_x(e^x)$ with the value produced by differentiation. In this case $\partial_x (e^x) = e^x$, so our helper function will be $x e^x$. 

Second, differentiate the helper function, which we can do using the product rule: $$\partial_x (x e^x) = x e^x + e^x$$

Third, take the anti-derivative of each term in the above equation. Some of them are easy, but one of them is not.
$$\int \partial_x (x e^x)
 = \int x e^x dx + \int e^x dx\\
 \text{integrating what we can ...}\\
x e^x\ \ \ \ = \int x e^x dx +\ \ \ \ e^x$$ The one we don't know how to do is the same as the original problem. So now we know how to do it, by re-arranging the previous equation:
$$\int x e^x dx = x e^x - e^x + C$$
We added in a constant $C$ at the end to get the entire family of anti-derivatives.

Aren't sure this answer is right? You can check it by computing the derivative of the answer to see that it gives $x e^x$.

**Problem 1**: Go through the steps above to find the anti-derivative of $g(x) \equiv x \cos(x)$. 

Step 1 hint: We know the anti-derivative of $\cos(x)$ is $\sin(x)$, so an appropriate helper function is the function $x\, \sin(x)$. Now do steps (2) and (3): (2) take the derivative of the helper function and then (3) integrate each term in the result. 

```{r parts1, echo=FALSE}
askMC(
  "What is the derivative of the helper function with respect to $x$?",
  "+$\\partial_x \\text{helper}(x) = \\sin(x) + x \\cos(x)$+",
  "$\\partial_x \\text{helper}(x) = \\sin(x) + x \\sin(x)$",
  "$\\partial_x \\text{helper}(x) = \\sin(x) + \\cos(x)$",
  "$\\partial_x \\text{helper}(x) = \\sin(x)\\cos(x)$"
)
```

```{r parts2, echo=FALSE}
askMC(
  "What is $$\\int \\partial_x \\text{helper}(x)\\ ?$$",
  "+$\\text{helper}(x) + C$+" = "We included the constant of integration $C$ just as a reminder.",
  "$\\frac{1}{2} \\text{helper}^2(x)$",
  "$1 / \\text{helper}(x)$",
  "Whatever it is, it's just as complicated as the original integral. No obvious way to do it." = "Actually, the answer is simple, even if it doesn't seem to get us anywhere. Wait for the next problem!")
```

```{r parts3, echo=FALSE}
askMC(
  "What is the integral with respect to $x$ of the first part of the expanded form of the helper function, that is, $\\int \\sin(x) dx$?",
  "+$-\\cos(x)$+" = "This is one of our basic modeling functions.",
  "$\\cos(x)$" = "Close. $\\cos(x) = \\partial_x \\sin(x)$, but we want $\\int \\sin(x) dx.$",
  "$e^x \\sin(x)$", 
  "$e^x \\cos(x)$"
)
```

```{r parts4, echo=FALSE}
askMC(
  "What is the integral with respect to $x$ of the second part of the expanded form of the helper function, that is, $\\int x\\, \\cos(x)$?",
  "It's the same as the original problem! I thought you were showing us how to do the problem. If we didn't know the answer when we started, why should we be able to do it now?" = "An understandable frustration. But think! Now you have the original problem written out as part of an equation with two integrals that you *do* know.",
  "+It's the same as the original problem. I've got an equation involving the original problem and two bits of algebra/calculus that I know how to do. Thanks!+",
  "$\\sin(x)$" = "If this were true, then $\\partial_x \\sin(x)$ would give us $x\\, \\cos(x)$. But, as you know, $\\partial_x \\sin(x) = \\cos(x)$, so this answer must be wrong."
)
```

Solve for the answer to the original function and write the function in R notation here:

```{r parts5, exercise=TRUE, exercise.cap="Answer to the original problem.", exercise.nlines=3, eval=FALSE}
makeFun( ...your stuff here... ~ x)
```

```{r parts5-check, echo=FALSE}
grade_result(

  pass_if( ~ sd(.result(1:10) -  ((1:10)*sin(1:10) + cos(1:10))) < 0.001),
  fail_if( ~ TRUE, message = "No!")
)
```


**Problem 2**: Use the same procedure to find the anti-derivative of $x\, \cos(2x)$. Since $$\cos(2x) = \frac{1}{2}\partial_x \sin(2 x)$$ a reasonable guess for a helper function will be $x \sin(2x)$. 


(We have intentionally dropped the $1/2$ to simplify the rest of the procedure. You'll see that such multiplicative constants don't matter, since they will be on both sides of the equation showing the derivative of the helper function. You can see this by keeping the $1/2$ in the helper function and watching what happens to it.)

As you work through the steps be very careful about the constants and make sure you check your final answer by differentiating.

```{r parts6, echo=FALSE, results="markup"}
askMC(
  "What is $\\partial_x x\\, \\sin(x)$?",
  "+$\\sin(x) + x\\, \\cos(x)$+",
  "$\\sin(x) + x\\, \\sin(x)$",
  "$\\cos(x) + x\\, \\sin(x)$",
  "$\\cos(x) + x\\, \\cos(x)$"
)
```

```{r parts7, echo=FALSE, results="markup"}
askMC(
  "What is $\\int \\partial_x [ x\\, \\sin(x)] dx$?",
  "+$x\\, \\sin(x)+C$+" = "Integration undoes differentiation!",
  "$\\sin(x)+C$",
  "$\\cos(x)+C$",
  "$x\\, \\cos(x)+C$"
)
```

```{r parts8, echo=FALSE, results="markup"}
askMC(
  "What is $\\int x\\, \\cos(x) dx$?",
  "+$x\\, \\sin(x) + \\cos(x)$+", 
  "$x\\, \\cos(x) + \\cos(x)$",
  "$x\\, \\cos(x) + \\sin(x)$",
  "$x\\, \\sin(x) + \\sin(x)$"
)
```


**Problem 3**: A giant tortoise (with very good eyesight and standing on an unobstructed plane!) spies a head of lettuce on the ground 65 meters away.  Being hungry (and knowing the shortest path between two points on the plane!), the tortoise takes off in a straight line for the lettuce. She pretty quickly reaches her top speed, but then starts to tire. If her velocity as a function of time (in meters per minute) is modeled by $v(t) = 7 t e^{−0.3t}$, how long does it take the tortoise to reach her lunch? Answer this question by finding an calculus/algebra formula for the tortoise’s displacement and then use it to approximate how long it takes to get to the lettuce. 

We're going to be looking at $\int v(t) dt = 7 \int t e^{-0.3 t} dt$.

We'll call the left side of the equation "displacement(t)". 
Use integration by parts to find displacement(t) as a simple formula in $t$.

The tortoise to reach the cabbage at time $t^\star$ such that $\text{displacement}(t^\star) = 65$ meters. Graph your displacement function to find $t^\star$. You can use the sandbox. (Note that the graphics domain isn't necessarily the best choice for answering the question.)

```{r parts12, exercise=TRUE, exercise.nlines=5, exercise.cap="Solving for t_star.", eval=FALSE}
displacement <- makeFun(77.77 * WHAT ~ t)
slice_plot(displacement(t) ~ t, domain(t=c(0,5)))
```

```{r parts13, echo=FALSE, results="markup"}
askMC(
  "At what time $t^\\star$ does the tortoise reach the cabbage?",
  "5.95 sec",
  "+10.85 sec+",
  "15.75 sec",
  "Never! (That is, $t^\\star$ is infinite.",
  random_answer_order = FALSE
)
```


**Commentary**: The procedure we have been using to find these anti-derivatives can be formalized into a method called *integration by parts* (IBP).  However, rather than simply giving a formula and an algorithm, we wanted you to understand what is actually going on behind the scenes.  In a more formal IBP approach, you try to re-arrange the function you are integrating into the product of two simple functions: $$\int f(x)\, \partial_x g(x)\, dx$$

That is, you assume part of the original function is the derivative of some function $g(x)$. Once you’ve decided how to make the re-arrangement into $f(x)$ and $\partial_x g(x)$, and calculated $g(x)$ from $\partial_x g(x)$, you can write the original integral into a new, and possibly simpler, way:
$$\int f(x)\, \partial_x g(x)\, dx = f(x)\, g(x) - \int [\partial_x f(x)]\, g(x) dx$$ Notice that the formula replaces the integral we want to compute with another integral.  The idea is that the new integral will be easier to compute than the one you started with.  This procedure works in much more generality than for the examples we did, but it is not so easy to use.  You need to correctly assign $f(x)$ and $\partial_x g(x)$ to the parts of the original function, and you need to be able to integrate $\partial_x g(x)$ to get $g(x)$. Finally, you need to be able to integrate $[\partial_x f(x)]\, g(x)$. Perhaps you can go back and look at the previous problems with this more general description of the procedure.   


