---
title: "Class size"
author: "Daniel Kaplan"
---

```{r include=FALSE, contect="setup"}
library(mosaic)
library(mosaicCalc)
library(etude2)
```

A school district is considering how to balance class size against teachers' salaries.  At issue, of course, is the elemental fact that you need to hire more teachers in order to have smaller classes, but since funding is fixed, you won't be able to pay those teachers as well.  On the other hand, by having large classes, you need fewer teachers who can be paid more.

A consulting firm has used national data on test scores $T$ as a function of teacher salary $S$ and class size $C$ to build a model of the function $T = f(C,S)$
```{r echo = FALSE}
score <- function(C,S) {
  400*(C+ 10)^(-1.1) * S^(0.5) + 300
}
```

Here is a contour plot of that model function: expected student test scores, $T$ as a function of class size $C$ and teacher salary $S$:

```{r score1,echo = FALSE}
contour_plot(score(C, S) ~ C + S,
         domain(C=c(10,50),S=c(30,100)),
         filled=FALSE) %>%
gf_labs(x="Class Size (C)",
        y="Teacher Salary (S, in $1000s)",
        title="Test Scores as function of Teacher Salary and Class Size")
```


1. What's the value $T$ of this function when class size $=20$ and   teacher salary  $=50$?

#. Give two different values of class size and teacher salary for which the resulting test score $T=400$.

#. This isn't a linear function.  But if you focus on a small region of the Salary/Class-size plane, it is approximately linear. Draw two  boxes on the graph. The first should enclose a region where the function is approximately linear and where the output depends weakly on class size.  The second should enclose a region where the function is approximately   linear and the output depends strongly on class size.  Label the regions "weak" and "strong'' respectively.

For the school district, there is a straightforward relationship between salary (in $1000s) and class size.  $$S = 4.5 C - 80 .$$ (This is explained at the bottom of the problem, but you don't need to know about it to do the problem.)


4. Two lines are shown on the figure.  Only one of them matches the  relationship given in the equation.  Mark clearly  which line is correct. (Use the label ``correct").

#. Keeping in mind that the only allowed combinations of class size   and teacher salary that fit your budget lie on the line of the budget   constraint, choose the best possible class size to maximize the test   scores.  Mark this best-possible class size   with a $\star$ on the graph. Next to the $\star$, write the   (approximate) value of the output at this point.


```{r scores2,echo = FALSE}
contour_plot(score(C, S) ~ C + S,
         domain(C=c(10,50), S=c(30,100)), filled=FALSE) %>%
  gf_labs(x="Class Size (C)", y="Teacher Salary (S, in $1000s)",
          title="Test Scores as function of Teacher Salary and Class Size") %>%
  gf_abline(slope=4.5, intercept = -80, alpha=.5, lwd=4) %>%
  gf_abline(slope=2, intercept = -5, alpha=.5, lwd=4)

```



*Just leisure reading for your edification at a later date.*  Here are the factors that go into shaping the relationship between class size and teacher salary:

* The district receives $10,000 per year per student from the state.
* They have non-teaching costs of $5500 per student for testing, registration, special education, bussing, etc.
* It costs approximately $80,000 per year to provide a classroom -- this includes heating, the common spaces in the building, etc.
* The rest goes to teacher salary, benefits (such as health insurance and retirement), etc.  Call all of this salary $S$, recognizing that the teacher doesn't receive all of it directly.

Put together, the relationship between teacher salary $S$ (in thousands) and class size $C$ is $10 C = 5.5 C + 80 + S$.
