---
title: "Making polynomials"
author: "Daniel Kaplan"
---

There's no linear combination of lower-order polynomials to give a higher-order polynomial (or any linear combination of higher-order polynomials).

Example: Suppose $a_0 + a_1 x + a_2 x^2 + a_3 x^3  = x^n$ Differentiate both sides giving $a_1 + 2 a_2 x + 3 a_3 x^2 = n x^{n-1}$ 

Differentiate again $2 a_2 + 3! a_3 x = n(n-1)x^{n-2}$$

Suppose you wanted to use the polynomial computer to calculate values of $f(x) \equiv e^x$. 

Take 
$$p(x) = 1 + b_1 x + b_2 x^2 + b_3 x^3 + b4 x^4 + \cdots$$ where we have arbitrarily set the function value $p(0) = 1$.

The handle we have on the problem is that $\partial_x e^x = e^x$

$$\partial_x p(x) = b_1 + 2 b_2 x + 3 b_3 x^2 + 4 b_4 x^3 + 5 b_5 x^4$$

In order for $\partial_x p(x) = p(x)$ for all $x$ then we can equate the coefficients.

$$b_1 = 1\\
2 b_2 = b_1 \ \ \implies b_2 = \frac{1}{2}\\
3 b_3 = b_2 \ \ \implies b_3 = \frac{1}{3 \cdot 2}\\
4 b_4 = b_3 \ \ \implies b_4 = \frac{1}{4 \cdot 3 \cdot 2} = \frac{1}/{4!}$$
So our function $p(x)$ will be

In order to make the pattern glaringly obvious, we'll write the function $g(x) \equiv 1$ as $g(x) \equiv x^0$ and we'll use the fact that $0! = 1! = 1$. The result is:

$$p(x) = \frac{x^0}{0!} + \frac{x^1}{1!} + \frac{x^2}{2!} + \frac{x^3}{3!} + \frac{x^4}{4!} + \cdots$$
We'd like to plot out the function. Suppose we do this over the domain $-1 \leq x \leq 1$. How many terms do we need to get a plausible approximation.

We know that the next term will be $x^5/5!$ which, when $x=1$ is $1/5! = 1/120$. The term after that will be $x^6/6!$ which will be even smaller: $1/6! = 1/720$. The sum over the first five terms will be $1 + 1 + 1/2 + 1/6 + 1/24 = 2 17/24$. The next $x^5$ term would only add 0.3% to the sum, a displacement of 3 pixels in a graph of height 1000 pixes.

```{r}
p <- function(x) {
  1 + x + x^2/factorial(2) + 
  x^3/factorial(3) + x^4/factorial(4)
}
slice_plot(p(x) ~ x, domain(x=c(-1,1)))
```
Let's also look at the error at $x=-1$. Using the same logic, the sum of the previous terms will be $1 - 1 + 1/2 - 1/6 + 1/24  = 9/24$. 

$1 + 1/2 + 1/8 + 1/48 + 1/384 $ with the next terms being $1/3840\approx 0.0003$  and the one after that being $1/46080 = 0.00002$. 

But at $x=2$ the situation is not so good. The series will be $1 + 1 + 4/2! + 8/3! + 16/4! = 9.25$ and the next term will be $32/5! = 6.4$. 

## With sine

$\partial_{xx} sin(x) = - sin(x)$

$$s(x) \equiv c_0 + c_1 x + c_2 x^2 + c_3 x^3 + c_4 x^4 + (c_5 x^5 + c_6 x^6)$$

$$\partial_{xx} s(x) = 2 c_2 + 3\cdot 2 c_3 x + 4\cdot 3 c_4 x^2 + (5\cdot 4 c_5 x^3 + 6 \cdot 5 c_6 x^4)$$

This implies $$-2 c_2 = c_0  \ \ \implies\ \ \ c_2 = -\frac{c_0}{2!}\\
-6 c_3 = c_1  \ \ \implies\ \ \ c_3 = -\frac{c_1}{3!}\\
-12 c_4 = c_2  \ \ \implies\ \ \ c_4 = -\frac{c_2}{4\cdot 3} = -\frac{c_0}{4!}\\
-20 c_5 = c_3  \ \ \implies\ \ c_5 = \frac{c_1}{5!}\\
-30 c_6 = c_4 \ \ \implies\ \ c_6 = = \frac{c_0}{6!}$$

```{r}
s <- function(x, c0=0, c1=1) {
  c0 + c1*x - c0*x^2/2 - c1*x^3/factorial(3) +
    c0*x^4/factorial(4) 
}
slice_plot(s(x, 1, 1) ~ x, domain(x=c(-5,6)))
```

Actually, there are many functions whose 2nd derivative equals the negative of the function itself. You get to choose any value of $c_0$ and $c_1$ you like. 
