---
title: "Ebola sigmoid"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(MMAC)
```

A data frame named `EbolaSierraLeone`^[From the `{MMAC}` R package] records the cumulative number of Ebola cases in  Sierra Leone from May 1, 2014 to December 16, 2015.

Let's find how well a sigmoidal function models the data.


```{r}
gf_point(Cases ~ Day, data = MMAC::EbolaSierraLeone)
```

It's straightforward to estimate the values of the parameters from data that have a sigmoidal form. To walk you through the process, the next plot shows simulated data from a genuine sigmoid function.

```{r echo=FALSE, warning=FALSE}
sigmoid <- makeFun(A * pnorm(t, center,  width) ~ t)
Points <- tibble::tibble(
  t = seq(0, 12, length=100),
  y = sigmoid(t, A=15, center=5, width = 2)
)

gf_jitter(y ~ t, data  = Points,  height = 0.5, alpha = 0.25) %>%
slice_plot(sigmoid(t, A=15, center = 5, width = 2) ~ t,
           domain(t = c(-1, 15))) %>%
  gf_hline(yintercept  = 15, color  = "red",  alpha = 0.5) %>%
  gf_vline(xintercept = 5, color = "blue", alpha = 0.5) %>%
  gf_vline(xintercept =  c(3, 7),  color = "green", alpha = 0.5) %>%
  gf_segment(7.5 + 7.5 ~ -1 + 5, color = "blue",  alpha = 0.5) %>%
  gf_labs(title  = "A genuine sigmoid and its parameters")
```

1. Sketch in a S-shaped curve that smoothly follows the data.

2. Find the top plateau of the S-curve. This is indicated by the red line in the graph above. The parameter `A` is simply the height of the plateau, in this case $y \approx 15$.

3. Come down half way from the  plateau. Here, that's 15/2 or 7.5, indicated by the horizontal blue line segment.  Find the inverse of the S-curve from that half-way point onto the horizontal-axis. Here, that gives $t \approx 5$. The  parameter `center` is that value.

4. From the center of the S-shaped curve, follow the curve upward about 2/3 of the way to the  plateau. In the diagram, that point is marked with a green line at $t \approx 7$. The `width` is the distance along the horizontal axis from the blue centerline to the green line. Here, that's $7 - 5$ giving 2 as the `width`.

5. You might also want  to trace the S-curve *downward* from the centerline about 2/3 of the way to zero. That's indicated by the  left green line.  In the standard sigmoid, the two green lines will be equally spaced around the centerline. Of course the data may not be in the shape of the standard sigmoid, so you might find the two green lines are not equally spaced from the center.

-------

*Put aside for the moment that the Ebola data doesn't look exactly like the standard sigmoid function. Follow the above procedure, nevertheless.*

```{r daily-digital-08-QA12, echo=FALSE}
askMC(
  prompt = "Where is the top plateau?",
    "About Day 600." = "Measure the height of the plateau, not where it starts horizontally.",
    "+About 14,000 cases+" = "",
    "About 20,000 cases" = "Read the vertical axis markings more carefully.",
    "None of the above" = "One of the above answers is pretty good.",
  random_answer_order = TRUE
)
```

```{r daily-digital-08-QA13, echo=FALSE}
askMC(
  prompt = "Where is the centerline?",
    "+Near Day 200+" = "",
    "Near Day 300" = "That's the center of the vertical scale, not the day at which the curve reaches half-way to the eventual plateau.",
    "At about 7000 cases" = "That's half-way up to the plateau, but the answer you  want is the day at which the curve reaches that point."
)
```

```{r daily-digital-08-QA14, echo=FALSE}
askMC(
  prompt = "Now to find the `width` parameter. The curve looks more classically sigmoidal to the left of the centerline than to the right, so follow the curve *downward* as in Step 4 of the algorithm to find the parameters. What's a good estimate for `width`?",
    "+About 50 days+" = "",
    "About 100 days" = "Too wide!",
    "About 10 days" = "Too small",
    "About 2500 cases" = "The width is measured along the horizontal axis, not the vertical"
)
```

Open a sandbox and use the following commands to plot the Ebola data along with the sigmoid that you matched to the data by eye. The commands also generate and plot a more "refined" estimate found with `fitModel()` starting with  your by-eye parameters. (Note: You'll need to fill in the blanks at the start of the scaffolding with your by-eye parameters before running the code.)


::: {.scaffolding}
```{r eval=FALSE}
A_estimate <-  _____
center_estimate <- _____
width_estimate <- ______

guessed_fun <- makeFun(A* pnorm(Day, center, width) ~ Day,
                       A = A_estimate,
                       center = center_estimate,
                       width = width_estimate)


refined_fun <- fitModel(Cases ~ A * pnorm(Day, center, width),
                 data = EbolaSierraLeone,
                 start = list(A=A_estimate, 
                              center=center_estimate,
                              width=width_estimate))

gf_point(Cases  ~ Day, data = MMAC::EbolaSierraLeone)  %>%
  slice_plot(guessed_fun(Day) ~ Day, color="red") %>%
  slice_plot(refined_fun(Day) ~ Day, color="blue")
```
:::


I can't predict what you see in your graph, since it depends on what starting estimates you used for `A`, `center` and `width`.

```{r daily-digital-08-Essay1, echo=FALSE}
etude2::etudeEssay(
  prompt = "Write down the estimates you  made by eye for `A`, `center` and `width`. Then  explain in words any major systematic deviations of the 'refined' model from the data."
)
```

This idea of fitting a sigmoidal curve *globally* to data that are only roughly sigmoidal might be forgiven since it's just a textbook exercise. But if this were a real-world problem, there are three rookie mistakes in the approach, despite its technical sophistication.

1. A model is built for a  purpose. There's no statement in  the exercise about  what the purpose is.
2. The data themselves are a  good enough representation  of the historical pattern. There's no need to adorn  them with a fitted  curve.
3. The mathematical processes that would justify the expectation  that a sigmoid curve could be useful are based on the idea that the  underlying mechanism for the phenomenon is unchanging. For an epidemic, that mechanism is contagion. By saying the "mechanism is unchanging," we mean that person-to-person transmission of the Ebola virus happens in the same way at the beginning of the epidemic as it does in the middle and in  the end. But strong public health interventions were introduced to control the epidemic. That's a change in  mechanism.

