---
title: "Magnitudes and engines"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```


Here is a plot of the power output (BHP) versus displacement (in cc) of 39 internal combustion engines.

::: {.scaffolding}
```{r}
gf_point(BHP ~ displacement, data = Engines) %>%
  gf_lims(y  = c(0, 30000))
```
:::

```{r EM1-1, echo=FALSE}
etude2::etudeQ(
  prompt = "Your study partner claims that the smallest engine in the data has a displacement of 2000 cc (that is, 2.0 liters) and 100 horsepower. Based only on the graph, is this claim plausible?",
    "+Yes, because 2000  cc and 100 hp would look like (0, 0) on  the scale of this graph.+" = "",
    "Yes, because that size engine is typical for a small car." = "That may be, but certainly you've encountered lawn  mower engines that are much smaller.",
    "No, the smallest engine is close to 0 cc." = "Would you  be able to distinguish visually an engine of 1 cc from an engine of 1000cc on this graph? Both these values would lie on the same horizontal pixel in the graph.",
    "No, my study partner is always wrong." = "Be that as it may, we're looking for a principled answer, not an *ad hominem* one."
)
```

**Semi-log scales**

The next command will make a graph of the same engine data as before, but with a log scale on the horizontal axis. The vertical axis is still linear. 

::: {.scaffolding}
```{r}
gf_point(BHP ~ displacement, data = Engines) %>%
  gf_refine(scale_x_log10())
```
:::


```{r EM1-2, echo=FALSE}
etude2::etudeQ(
  prompt = "Using just the graph, answer this question: The engines range over how many decades of displacement? (Remember, a decade is a factor of 10.)",
    "+7 decades+" = "",
    "Can't tell" = "Yes, you can. Figure out what one decade corresponds to in terms of distance on  the log axes.",
    "$10^7$ decades" = "The estimated volume of the entire universe is about $4 \\times 10^{86}$ cc. The volume of a neutron is about $6 \\times 10^{-81}$ cc. The range between a neutron and the universe is therefore about $86  - -81 = 167$ decades. Do you think it likely that there is an internal combustion engine smaller than a neutron or larger than the universe?",
    "About 3.5 decades" = "Perhaps you're treating the distance between axis labels as one decade. Look carefully and you see that it's actually a factor of 100, that is, two decades."
)
```

**Log-log scales**

In  order to make a graph with log-log scales, we refine  the graph with both `scale_y_log10()` and `scale_x_log10()`, as in the following  ...

```{r}
gf_point(BHP ~ displacement, data = Engines) %>%
  gf_refine(scale_x_log10(), scale_y_log10())
```

Based on the log-log plot, answer these questions.

```{r EM1-3, echo=FALSE, results="markup"}
etude2::etudeQ(
  prompt = "How many engines have a displacement of 1 liter or less?",
    "none" = "Perhaps you recognized that the left-most tick mark corresponds to a value of 1, and that no data points are 1 or smaller. But one liter corresponds to 1000 cc.",
    "7" = "This is the number of engines with displacement of 10 cc or smaller. But one liter corresponds to 1000 cc.",
    "+14+" = "Right. It's the $10^3$ tick that marks 1 liter, since 1 liter is 1000 cc.",
    "25" = "That would be true if the cut-off were 10 liters. But it's not.",
  randomAnswerOrder = FALSE
)
```

```{r EM1-4, echo=FALSE, results="markup"}
etude2::etudeQ(
  prompt = "Using the log-log plot, how many decades of BHP are spanned by the data?",
  choices = list(
     "4" = "Not a bad answer, but not the best one either. Notice that the smallest engine is about half a decade below 1 BPM, and the largest is  about half a decade above 10,000 BPH",
    "+5+" = "",
    "100" = "The number $10^{100}$ is called a *googol* and is roughly how many particles (including photons, neutrinos, etc.)  are in the universe. Imagine,  quite contrary to fact, that 1 BHP could be generated by burning one molecule of fuel per second. Then as many fuel molecules as there are particles in the universe would have  to  be burned each second to power an engine at the high end of a span of 100 decades."
  ),
  random_answer_order = FALSE
)
```




