---
id: "fish-sees-tree"
author: "Daniel Kaplan"
---

Open a `r sandbox_link()`. (Just click on that link, although you may eventually be given other ways to open a sandbox.)

::: {.scaffolding}
When you see a breakout box like this, it means that we're providing some computer code that you can paste into a sandbox and run. For this exercise, that code is

```{r results="hide"}
x <- 2
sin(x)*sqrt(x)
```

Paste this into the sandbox and press "Run code." Verify that you get this as a result:

`[1] 1.285941`
:::

The sandbox has two commands, one on each line. The first command gives a value to $x$. The second command uses that value for $x$ to calculate a function output. The function is $g(x)\equiv \sin(x) \times \sqrt{\strut x}$. You'll see how to give a function a name like $g()$ in Section \@ref(objects-and-actions). 

::: {.why}
Why not simplify the above code to `sin(2)*sqrt(2)`? We want to make it clear to the reader (and the computer) that whatever $x$ might be, it should be used as inputs to **both** the $\sin()$ and the $\sqrt{\strut\ \ \ }$ functions.
:::

```{r fst9-1, echo=FALSE, results="markup"}
askMC(
  "Change $x$ to 1. What's the output of $\\sin(x) \\times \\sqrt{\\strut x}$",
  "-1.51", "0.244", "+0.84+", "0.99", "2.14", "`NaN`",
  random_answer_order=FALSE
)
```

```{r fst9-2, echo=FALSE, results="markup"}
askMC(
  "Change $x$ to 3. What's the output of $\\sin(x) \\times \\sqrt{\\strut x}$",
  "-1.51", "+0.244+", "0.84", "0.99", "2.14", "`NaN`",
  random_answer_order=FALSE
)
```

```{r fst9-3, echo=FALSE, results="markup"}
askMC(
  "Change $x$ to $-5$. What's the output of $\\sin(x) \\times \\sqrt{\\strut x}$",
  "-1.51", "0.244", "0.84", "0.99", "2.14", "+`NaN`+" = "This stands for Not-a-Number, which is what you get when you calculate the square root of a negative input.",
  random_answer_order=FALSE
)
```

In the sandbox, change the **function** to be $\sqrt{\strut\pnorm(x)}$.

```{r fst9-4, echo=FALSE, results="markup"}
askMC(
  "For $x=2$, what's the output of $\\sqrt{\\strut\\pnorm(x)}$",
  "-1.51", "0.244", "0.84", "+0.99+", "2.14", "+NaN`",
  random_answer_order=FALSE
)
```
