---
chapter: "Modeling Functions"
author: "Danny Kaplan"
date: 2020-07-17 
tags: [COVID-19, semi-log axes, third]
id: maple-hit-saucer
global-id: "j3xe"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```


You have likely heard the phrase "exponential growth" used to describe the COVID-19 pandemic. Let's explore this idea using actual data. 

The [COVID-19 Data Hub](https://covid19datahub.io/) is  a collaborative effort of universities, government agencies,  and NGOs to provide up-to-date information about the pandemic. We're going to use the data about the US at the whole-country level. (There's also data at state and county levels.) 

Perhaps the simplest display is to show the number of cumulative cases (the `confirmed` variable) and deaths as a function of time. You can see the documentation for the dataset at the  [COVID-19 Data Hub](https://covid19datahub.io/) 

Open a sandbox and make the plot.

::: {.scaffolding}
We're going to plot confirmed cases in blue and deaths in red.
```{r results="hide"}
gf_point(confirmed ~ date, data = Covid_US, color = "blue") %>%
  gf_point(deaths ~ date, color = "red")
```
:::

```{r maple-hit-saucer-QA1, echo=FALSE}
message1 <- "The number 1e6 means 1,000,000, that is, six zeros following the 1."
message2 <- "Mid July is the tick mark *after* the mark labelled `Jul`."
askMC(
  prompt = "As of mid July, about how many confirmed cases were there? (Note that the labeled tick marks refer to the beginning of the month, so the point labeled `Feb` is February 1.)",
    "about 300,000" = message1,
    "about 350,000" = message1,
    "about 400,000" = message1,
    "about 3,000,000" = message2,
    "+about 3,500,000+" = "",
    "about 4,000,000" = message2, 
  random_answer_order = FALSE,
  inline = FALSE
)
```

We're going to do something to make the graph more informative. (At least, more informative if you understand about logarithms and exponentials!)

Here's the same graphic as above, but taking the *logarithm* of the number of cases (that is, `confirmed`) and of the number of deaths. Since we're taking the logarithm of only the y-variable, this is called a "semi-log" plot. 

::: {.scaffolding}
```{r results="hide"}
gf_point(log(confirmed) ~ date, data = Covid_US, color = "dodgerblue") %>%
  gf_point(log(deaths) ~ date, color = "orange3") 
```
::: 

Up through the beginning of March in the US, it is thought that most US cases were in people travelling into the US from hot spots such as China and Italy and the US, as opposed to contagion between people within the US. (Such contagion is called "community spread.") So let's focus on the data from the start of March onward.

Exponential growth appears as a straight-line pattern on a semi-log plot. Obviously, the overall pattern of the curves is not a straight line. The explanation for this is that the exponential growth rate changes over time, perhaps due to public health measures (like business closing, masks, etc.)

The first (official) US death from Covid-19 was recorded was recorded on Feb. 29, 2020. Five more deaths occurred two days later, bringing the cumulative number to 6.


```{r maple-hit-saucer-QA3, echo=FALSE}
askMC(
  prompt = "The red data points for Feb 29/March 1 show up at zero on the vertical scale. The red data point for March 2 is at around 2 on the vertical scale. Is this consistent with the facts stated above?",
    "No. The data contradict the facts." = "Think about what it means to be 0 on the vertical scale.",
    "+Yes. The vertical scale is in log units, so 0 corresponds to 1 death, since $\\ln 1 = 0$.+" = "",
    "No. The vertical scale doesn't mean anything." = "You can see from the plotting command what the quantity on the vertical axis is: `log(confirmed)` for the blue dots and `log(deaths)` for the red."
)
```


One of the purposes of making a semi-log plot is to enable you to compare very  large numbers with very small numbers on the same graph. For instance, in the semi-log plot, you can easily see when the first death occurred, a fact that is invisible in the plot of the raw counts (the first plot in this exercise).

Another feature of semi-log plots is that they preserve proportionality. Look at the plot of raw counts and note that the curve for the number of deaths is much shallower than the curve for the number of (confirmed) cases. Yet on the semi-log plot, the two curves are practically parallel.

On a semi-log plot, the arithmetic difference between the two curves tells you what the proportion is between those curves. The parallel curves mean that the proportion is practically constant. 
Calculate what the proportion between deaths and cases was in the month of May. Here's a mathematical hint: $\ln \frac{a}{b} == \ln a - \ln b$. We are interested in $\frac{a}{b}$.

```{r maple-hit-saucer-QA4, echo=FALSE}
askMC(
  prompt = "What is the proportion of deaths to cases during the month of May?",
    "about 0.2%" = "Remember, it's the natural logarithm that's being plotted, so the inverse function is `exp()`.",
    "about 1%" = "This would correspond to a (vertical) difference between the curves of about 4.6 log units. Is it really that big?",
    "about 2%" = "This would correspond to a (vertical) difference between the curves of about 4 log units. Is it really that big?",
    "+about 6%+" = "Good. You calculated `exp(log(deaths) - log(cases))` which gives an answer around 0.06, or 6%.",
    "about 25%" = "I'm not really sure what could lead you to this answer. You're making a mistake that I didn't anticipate.",
    "more than 50%" = "It's true that in May log(deaths) is about 12, and log(cases) is about 14, and 12/14 is indeed greater than 50%. But, on a log scale, the proportion relates to the difference between logs, not the ratio.",
  inline = FALSE,
  random_answer_order = FALSE
)
```

In many applications, people use semi-log plots to see whether a pattern is exponential or to compare very small and very large numbers. Often, people find it easier if the vertical scale is written in the original units rather than the log units. To accomplish both, the vertical scale can be ruled with raw units spaced logarithmically, like this:

::: {.scaffolding}
```{r}
gf_point(confirmed ~ date, data = Covid_US, color = "dodgerblue") %>%
  gf_point(deaths ~ date, color = "orange3") %>%
  gf_refine(scale_y_log10())
```
:::

The **labels** on the vertical axis show the raw numbers, while the **position** shows the logarithm of those numbers.

The next question has to do with the meaning of the interval between grid lines on the vertical axis. Note that on the *horizontal* axis, the spacing between adjacent grid lines is half a month.


```{r maple-hit-saucer-QA5, echo=FALSE}
askMC(
  prompt = "What is the numerical spacing (in terms of raw counts) between adjacent grid lines on the vertical axis? (Note: Two numbers are different by a \"factor of 10\" when one number is 10 times the other.\" Similarly, \"a factor of 100\" means that one number is 100 times the other.",
    "$10^1$" = "If this were true, moving up from the lowest label (`1e+01`, that is, 10) the next grid line would be at 20, then 30, then 40.",
    "$10^2$" = "If this were true, moving up from the lowest label (`1e+01`) the next grid line would be at 110, then 210, then 310.",
    "+A factor of 10.+" = "Right. Every time you move up by one grid line, the raw number increases ten-fold, so 10, 100, 1000, 10,000, and so on. The phrase `a factor of 10` means to multiply by 10, not to add 10.",
    "A factor of 100." = "You're thinking along the right lines, but this is the difference between every second grid line, not adjacent grid lines.",
  inline = FALSE,
  random_answer_order = FALSE
)
```


