---
id: "naked-axis-crossing"
author: "Daniel Kaplan"
---

For each of the naked modeling functions, the graph crosses either the vertical axis (that is $x=0$) or the horizontal axis (that is, $f(x) = 0$), or both. It's helpful to know the exact quantitative value for the input where the function graph crosses the horizontal axis, and the exact quantitative value for the output where the function graph crosses the vertical axis. 

These questions ask you, for each of the naked modeling functions, to find a specific input/output pair where you know the exact values. (Feel free to open a `r sandbox_link()` to try the various possibilities.)

```{r nac89-1, echo=FALSE, results="markup"}
askMC(
  "What is the exact output of the naked **exponential** function when the input is $x=0$?",
  0, "0.3989423", "1/2", "+1+",
  random_answer_order = FALSE
)
```

```{r nac89-2, echo=FALSE, results="markup"}
askMC(
  "What is the exact output of the naked **sine** function when the input is $x=0$?",
  "+0+", "0.3989423", "1/2", "1",
  random_answer_order = FALSE
)
```
```{r nac89-3, echo=FALSE, results="markup"}
askMC(
  "What is the exact input to  the naked **logarithm** function where the output is 0?",
  "0", "0.3989423", "1/2", "+1+",
  random_answer_order = FALSE
)
```

```{r nac89-4, echo=FALSE, results="markup"}
askMC(
  "What is the exact output of the naked **sigmoid** function when the input is $x=0$?",
  "0", "0.3989423", "+1/2+", "1",
  random_answer_order = FALSE
)
```

```{r nac89-5, echo=FALSE, results="markup"}
askMC(
  "What is the output (to several digits) of the naked **hump** function when the input is $x=0$?",
  "0", "+0.3989423+" = "Right. But 0.4 will do when you're sketching a graph.", "1/2", "1",
  random_answer_order = FALSE
)
```

```{r nac89-6, echo=FALSE, results="markup"}
askMC(
  "What is the exact input to  the naked **reciprocal* function where the output is 1? (That's not an axis crossing.)",
  0, "0.3989423", "1/2", "+1+",
  random_answer_order = FALSE
)
```

```{r nac89-7, echo=FALSE, results="markup"}
askMC(
  "What is the exact input to  the naked **proportional* function where the output is 1/2? (That's not an axis crossing.)",
  0, "0.3989423", "+1/2+", "1",
  random_answer_order = FALSE
)
```

```{r nac89-8, echo=FALSE, results="markup"}
askMC(
  "What is the exact output of the naked **constant** function?",
  0, "0.3989423", "1/2", "+1+",
  random_answer_order = FALSE
)
```

```{r nac89-9, echo=FALSE, results="markup"}
askMC(
  "The naked **square** function just touches the horizontal axis. What's the input where this happens?",
  "+0+", "0.3989423", "1/2", "1",
  random_answer_order = FALSE
)
```
