---
title: "Sigmoid intro"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

Hump functions and sigmoidal functions come in pairs. For  every possible sigmoid, there is a corresponding hump that gives, for each value of the input, the slope of the sigmoid.

Each of the following graphs shows a sigmoid and a hump function. The two might or might not correspond to one another. That is, the output of the hump might be the slope of the sigmoid, or the hump might correspond to  some other sigmoid. Remember, you're comparing the *output* of the hump to the *slope* of the sigmoid.

For each graph, say whether the hump and the sigmoid correspond to one another. If not, choose one of the reasons why not.

```{r echo=FALSE}
slice_plot(pnorm(x, mean = 5, sd = 3) ~ x, domain(x=c(-5,15)),
           label_text = "sigmoid") %>%
  slice_plot(dnorm(x, mean = 5, sd = 3) ~ x, color = "blue",  label_text = "hump") %>%
  gf_labs(title  = "(A)")
```

```{r daily-digital-08-QA7, echo=FALSE}
etude2::etudeQ(
  prompt = "Graph (A)",
    "+The hump and sigmoid correspond.+" = "",
    "The peak of the hump does not occur at the same value of $x$ at which the sigmoid is steepest." = "For what $x$ is the sigmoid the steepest? For what $x$ is the hump the highest?",
    "The numerical value of the output of the hump function is, for all $x$, much larger than the numerical value  of the *slope*  of the sigmoid." = "Did you  calculate the numerical value  of  the slope of the sigmoid?",
  inline = FALSE,
  random_answer_order = FALSE
)
```


```{r echo=FALSE}
slice_plot(pnorm(x, mean = 4, sd = 3) ~ x, domain(x=c(-5,15)),
           label_text = "sigmoid") %>%
  slice_plot(dnorm(x, mean = 2, sd = 1) ~ x, color = "blue",  label_text = "hump") %>%
  gf_labs(title = "(B)")
```

```{r daily-digital-08-QA8, echo=FALSE}
etude2::etudeQ(
  prompt = "Graph (B)",
    "The hump and sigmoid correspond." = "",
    "+The peak of the hump does not occur at the same value of $x$ at which the sigmoid is steepest.+" = "The hump peaks at about $x=2$ while the steepest part of the sigmoid is at about $x=4$",
    "The numerical value of the output of the hump function is much larger than the numerical value  of the *slope*  of the sigmoid." = "Did you  calculate the numerical value  of  the slope of the sigmoid?",
  inline = FALSE,
  random_answer_order = FALSE
)
```

```{r echo=FALSE}
slice_plot(pnorm(-x, mean = -5, sd = 3) ~ x, domain(x=c(-5,15)),
           label_text = "sigmoid") %>%
  slice_plot(-dnorm(x, mean = 5, sd = 3) ~ x, color = "blue",  label_text = "hump", label_x = 0.1) %>%
  gf_labs(title  = "(C)")
```


```{r daily-digital-08-QA9, echo=FALSE}
etude2::etudeQ(
  prompt = "Graph (C)",
    "+The hump and sigmoid correspond.+" = "",
    "The peak of the hump does not occur at the same value of $x$ at which the sigmoid is steepest." = "For what $x$ is the sigmoid the steepest? For what $x$ is the hump the highest?",
    "The numerical value of the output of the hump function is, for all $x$, much larger than the numerical value  of the *slope*  of the sigmoid." = "Did you  calculate the numerical value  of  the slope of the sigmoid?",
  inline = FALSE,
  random_answer_order = FALSE
)
```

In the  graph *D*, there are several hump functions shown, only one of which corresponds to the sigmoid.

```{r  echo=FALSE}
slice_plot(pnorm(x, mean = 1, sd = 3) ~ x, domain(x=c(-10,10)),
           label_text = "sigmoid", color = "blue") %>%
  slice_plot(7.5*dnorm(x, mean = 1,  sd = 3) ~ x, label_text = "hump A",
             label_x  = 0.6) %>%
  slice_plot(dnorm(x, mean = 1, sd = 3) ~ x,
             label_text = "hump B", label_x = 0.65) %>%
  slice_plot(dnorm(x, mean = 0, sd = 0.5) ~ x,
             label_text = "hump C", label_x = 0.475, npts = 200) %>%
  slice_plot(6*dnorm(x, mean = -1, sd = 4) ~ x,
             label_text = "hump D", label_x = 0.3) %>%
  gf_labs(title = "Graph (D)")
```

```{r daily-digital-08-QA10, echo=FALSE}
etude2::etudeQ(
  prompt = "Which hump corresponds to  the sigmoid?",
    "A" = "The value of the hump output is much larger than the slope of the sigmoid.",
    "+B+" = "Right! The hump is centered on the steepest part of the sigmoid and falls to zero where the sigmoid levels out.",
    "C" = "The hump is too narrow.",
    "D" = "The hump is too broad and shifted to the left.",
  random_answer_order = FALSE
)
```
