---
title: Sunset in LA
author: MMAC
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(MMAC)
```

The data frame `SunsetLA` records the number of minutes after 4 pm until the sun sets in Los Angeles, CA  over a 4-year interval from January 2010 (month 1) through December 2013 (month 48). 

<!-- (All times are already adjusted for daylight savings time. The book doesn't tell us whether the value for each month is an average over all days or for a particular day of the month.) -->

Open a sandbox and make a plot of sunset time versus month.

::: {.scaffolding}
```{r results="hide"}
gf_point(Minutes ~ Month, data = SunsetLA) %>%
  gf_line()
```

We're using both `gf_point()` and `gf_line()`. With data that oscillates up and down, connecting the data points with lines makes it easier to see the pattern.
:::

```{r dd-07-QA1, echo=FALSE}
askMC(
  prompt = "What is the range of the number of minutes until sunset over the whole 4-year period?",
    "+40 to 190 minutes+" = "",
    "120 minutes" = "A range is an interval spanned by *two* numbers.",
    "40 to 180 minutes" = "The largest values are half a tick mark up from 180. Tick marks are spaced by 20 minutes.",
    "0 to 48  months" = "That's the *domain*. The *range* is along the vertical axis."
)
```

```{r dd-07-QA2, echo=FALSE}
askMC(
  prompt = "The data fall nicely on a sine-shaped curve. What is the period of that sine?",
    "6 months"  = "Look at the number of months from one peak to another.",
    "11 months" = "Look more carefully. And remember that the change in length of day is an *annual* phenomenon.",
     "+12 months+" = "",
    "12 minutes" = "Period refers to an interval on  the *domain* of the function, not the range.",
  random_answer_order = FALSE
)
```

The function $$\mbox{sunset}(\mbox{Month}) \equiv A  \sin(2 \pi\, \mbox{Month} / 12) + C$$ is a *linear combination* of two functions:

1. The constant function `one(Month)`
2. The sine function `sin(2*pi*Month/12)`

The two functions are scaled by $C$ and $A$, respectively.

You can estimate $C$ by the mid-point of the range (in minutes) of the data. You can estimate $A$  as half the length of the range (that is, half the max minus the min.)

Make rough but reasonable numeric estimates for $C$ and $A$ from the data. Then, in the sandbox, define the `sunset()` function using the numerical estimates in the linear combination. Plot your function as a layer on  top of the data. (Pipe  the `gf_point()` layer  to `slice_plot()`.)

::: {.scaffolding}
```{r results="hide", eval=FALSE}
sunset <- makeFun(A + C*sin(2*pi*(Month - offset)/12), 
                  A = __your estimate__, 
                  C = __your estimate__,
                  offset=0)
gf_point(Minutes ~ Month, data = SunsetLA) %>%
  gf_line() %>%
  slice_plot(sunset(Month) ~ Month)
```
The domain for `slice_plot()` is inherited to that implied by the `SunsetLA` data. Notice that the input name in `slice_plot()` corresponds to that established in `gf_point()`.
:::

```{r dd-07-QA3, echo=FALSE}
askMC(
  prompt = "Your `sunset()` function should be a pretty good match to the data except for one thing. What is that thing?",
    "The `sunset()` function has a completely different range than the data." = "This won't be the case if you have estimated $C$ and $A$ correctly.",
    "The period of the `sunset()` function doesn't match the data." = "Did you use `sin(2*pi*Month/12)`? If so, the period should be right.",
    "+There is a horizontal time shift between `sunset()` and the data.+",
  inline = FALSE
)
```

We're going to fix the problem with `sunset()` by defining a *time offset* to use as a reference. For a sine function, a suitable time offset is the value along  the horizontal axis when the phenomenon being modeled crosses $C$ with  a positive slope. There are 4 such points along the horizontal axis readily identifiable in the data. (They may not be at an integer value of `Month`.)

```{r dd-07-QA4, echo=FALSE}
askMC(
  prompt = "Which of these is a suitable value for the time offset?",
    "0 months" = "That's not a time when the data suggest that $C$ is being crossed.",
    "19 months" = "That's a maximum, not a crossing  of $C$.",
    "21.5 months" = "That's a crossing of $C$, but not one with  a positive slope.",
    "+15.5 months+" = "This is a good rough value. Since  the  period is 12 months, you  could equally  well  have said the offset  is 3.5 months."
)
```

In the original scaffolding, the value of `offset` was zero. Change that to match your answer to the previous question.

Plot out the modified `sunset()` function and confirm that it is a much better match to the data than the original  (that is, the one  without the time offset). You can "tune" your function by tweaking the numerical values of the $A$, $C$, and $offset$  parameters until you  get a solid match.

Alternatively, you can use `fitModel()` to do the tuning for you. Plug in your estimates (a.k.a. "guesses") for the parameters in place of the `___` in  the following. Then run the code. You'll  see your estimate  of  the function compared to the result of having  the computer refine your estimate. Chances are, the computer does a better job of stringing the function through the data.

::: {.scaffolding}
```{r eval=FALSE}
## rough estimates from graph
rough_A <- __estimated_A__
rough_C <- __estimated_C__
rough_offset <- __estimated_offset___
guessed_fun <-
  makeFun(A*sin(2*pi*(Month - offset)/12) + C ~ Month,
          A  = rough_A, C = rough_D,
          offset = rough_offset)

tuned_fun <-
  fitModel(Minutes ~ A*sin(2*pi*(Month - offset)/12) + C,
         data  = SunsetLA,
         start = list(A = rough_A,
                      C = rough_C,
                      offset = rough_offset) )

gf_point(Minutes ~ Month, data = SunsetLA) %>%
  gf_line(color  = "dodgerblue") %>%
  slice_plot(tuned_fun(Month) ~ Month) %>%
  slice_plot(guessed_fun(Month) ~ Month,  color = "orange3")

```
:::


Perhaps you were expecting the tuned sine function to match the data exactly. It does not. One reason for this is that the Earth's orbit around the Sun is not exactly circular.  The sine function is only a *model* of the phenomenon, good for some purposes and not for others. For a more complete explanation, see [this article on Wikipedia](https://en.wikipedia.org/wiki/Sunrise_equation).

(*Thomas Swalm* contributed to this exercise.)
