---
title: "Solar Panel App"
author: "Daniel Kaplan"
date: "7/20/2021"
output: html_document
runtime: shiny
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
knitr::opts_chunk$set(echo = TRUE)
```

The graphs show the situation over the course of a single day.The top graph shows net power produced as a function of time of day. The bottom graph shows net energy accumulated in the battery since 00:00.  

Since you are obsessed with your solar system, you have installed an app on your phone. It beeps each time the stored energy in the battery changes by one kW-minute (that is 1/60 of a kWh). The beeps are recorded in the middle graph. The beeps have a low, sad tone when the energy stored is falling (shown as a red line). The beeps have a high, happy tone when the energy stored is rising (shown as a black line). 

Note that by click-dragging the mouse over the beep graph---the one in the middle---you can highlight an interval of time. The top and bottom graphs will then show the information corresponding to the beeps during that interval.

```{r  echo=FALSE}
# construct the f(t) function
make_rand_fun <- function(seed = NULL , center=0, width=5, aveval=0) {
  rawf <- rfun(~ x, seed = seed)
  offset <- mean(rawf(seq(-3,3,100)))
  function(x) { rawf((x - center)/(width/7)) - offset + aveval}
}

cone_diagram <- function(f,
                         dom=list(x=c(0,10)), # endpoints of plot
                         total_area = 10,
                         events_per_unit_area = 60,
                         startx = 0,
                         startn=0, #initial stock at time startx
                         left = NA, # limits of integration
                         right = NA
                         ) {
  xmin <- dom[[1]][1] # for convenience
  xmax <- dom[[1]][2]
  total_events = total_area * events_per_unit_area
  # find total number of events, positive or negative
  Fabs <- antiD(abs(f(x)) ~ x)
  # number of actual events, according to f(x)
  n_events <- diff(Fabs(dom[[1]]))

  # scaling the derivative to produce the specified
  # total number of events
  Points <- tibble::tibble(
    x = seq(xmin, xmax, length = 200),
    y = f(x),
    absy = abs(y))
  top <- mean(Points$absy)
  unsigned_area_under_curve <- top*diff(dom[[1]])
  rate_fun <- function(x) (total_area*f(x)/unsigned_area_under_curve)
  Points$bigY <- Points$y * total_events/n_events
  rect_area <- top * diff(dom[[1]])
  # How many random numbers to generate to get
  # a total of <total_events> (roughly)
  npts <- ceiling(unsigned_area_under_curve * events_per_unit_area)
  ## Generate the actual events
  Points <- tibble::tibble(
    x = seq(xmin, xmax, length = npts),
    y = f(x),
    absy = abs(y),
    yval = top * runif(npts)
  ) %>%
    filter(yval <= absy) %>%
    mutate(val = sign(y),
           color = ifelse(val < 0, "orange3", "black"),
           shape = ifelse(val < 0, "-", "+"))
  # the anti-derivative to be plotted
  F <- antiD(rate_fun(x) ~ x)

  # make the plots
  P1 <- slice_plot(rate_fun(x) ~ x, dom, npts = 500)
  if (!(is.null(left) || is.na(left) ||
        is.null(right) || is.na(right))) {
    Keepers <- tibble::tibble(
      x = seq(left, right, length = 500),
      y = rate_fun(x))
    P1 <- P1 %>% gf_ribbon(0 + y ~ x, data = Keepers, inherit=FALSE,
                           fill = "dodgerblue", color=NA, alpha = 0.3) %>%
      gf_ribbon(0 + y ~ x, data = Keepers %>% filter(y < 0),
                color = NA, fill = "orange3", alpha = 0.3, inherit = FALSE)
  }

    # gf_labs(y = fname) %>%
    # gf_theme( axis.title.x = element_blank(),
    #           axis.text.x = element_blank())

  P2 <- gf_vline(xintercept = ~ x, data = Points,
                 color = ~ color,
                 alpha = 0.3) %>%
    gf_refine(scale_color_identity(),
              scale_shape_identity(),
              scale_y_continuous(limits = c(-1,1),
                                 breaks=0, labels="00")) %>%
    gf_theme( axis.title.x = element_blank(),
              axis.text.x = element_blank(),
              panel.grid = element_blank()
    )


  F2 <- function(x) F(x) - F(startx) + startn

  P3 <- slice_plot(F2(x) ~ x, dom)
    if (!(is.null(left) || is.na(left) ||
          is.null(right) || is.na(right))) {
     Dots <- tibble::tibble(x = c(left, right),
                            y = F2(x))
     P3 <- P3 %>% gf_point(y ~ x, data = Dots, color="dodgerblue")
    }

    return(list(fplot = P1, eplot = P2, Fplot = P3, f = rate_fun,
                F = F))
}


```

```{r echo=FALSE}
raw <- make_rand_fun(3222, center = 12, width = 4, aveval =4)
solar <- function(x) raw(x) * dnorm(x, 13, 2)
consumption <- make_rand_fun(231, center = 12, width =8)
total <- function(x) 1.9*solar(x) - abs(consumption(x)/3)

```
```{r, echo=FALSE}
plotOutput("fplot", height="200px")
plotOutput("eplot", height="50px", brush = brushOpts("interval", direction="x",
                                                     opacity=0.5))
plotOutput("Fplot", height="200px")
```

```{r echo=FALSE}
create_plots <- reactive({
  suppressWarnings(
    cone_diagram(total, dom=list(x=c(0, 24)), 
                 total_area = 11, 
                 events_per_unit_area = 60,
                 left = input$interval$xmin , 
                 right = input$interval$xmax)
  )
})

output$fplot <- renderPlot({
  tmp <- create_plots()
  tmp$fplot  %>%
    gf_hline(yintercept = 0, color = "dodgerblue") %>%
    gf_labs(y = "Net power production (kW)", x = "") %>%
    gf_theme( axis.title.x = element_blank(),
              axis.text.x = element_blank())

})
output$eplot <- renderPlot({
  isolate({tmp <- create_plots()
  tmp$eplot %>%
    gf_labs(y = "Beeps")})
})
output$Fplot <- renderPlot({
  tmp <- create_plots()
  tmp$Fplot  %>%
    gf_hline(yintercept = 0, color = "dodgerblue") %>%
    gf_labs(y = "Battery level (kWh)", x = "Hour of day")
})
```
