---
title: "Daily Digital 11"
subtitle: "Math 142Z"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r include=FALSE}
#gradescope <- TRUE
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
#gradescope <- TRUE # format multiple-choice output for Gradescope
```


```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD-11") # Add the DD number
etude2::etude_document_name("DD-142Z-11") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
```

## Objectives

a. Construct differentials to compute accumulations of length, area, and accumulation.
b. Recognize how integration (and differentiation) with respect to L, T, or M links quantities with different dimension. 
c. Construct differentials that are constructed by integrating with respect to another variable, as with the differential of volume being an area which is in turn an integral of a function over L.

## A Strategy for solving

```{r child="strategy.Rmd"}
```


## `antiD()`

```{r child="antiD.Rmd"}
```

## Length of a curve 

```{r child="length-differential.Rmd"}
```

## Volumes under functions

```{r child="function-volume.Rmd"}
```

## Integral of the day

```{r child="IOD-K.Rmd"}
```

## LaTeX of the day

```{r child="latex-11.Rmd"}
```



<!--

```{r child="curve-length.Rmd", eval=FALSE}
```

## Differential orbit

```{r child="differential-orbit.Rmd", eval=FALSE}
```

## Areas and Volumes

```{r child="geometry.Rmd", eval=FALSE}
```

-->

<!--

## Example

Distribution of times between one hundred-year storm and the *second* one that follows. Integral over the probability function $\frac{1}{k^2}\\exp(-(x+y)/k) dx$. Then do the expectation value by inserting $(x+y)$ as the outcome quantity and integrating over both $dx$ and $dy$.

-->

## Hand in your work

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>


```{r echo=FALSE, messages=FALSE}
etude2::submissions_ui()
```

```{r context="server", echo=FALSE, messages=FALSE}
etude2::submissions_logic()
```



-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name()`

