---
title: "Exponential bases"
author: "Daniel Kaplan"
runtime: shiny_prerendered
---

These three expressions 

$$e^{kt}\ \ \ \ \ 10^{t/d} \  \ \ \ \  2^{t/h}$$

produce the same value if $k$, $d$ and $h$ have corresponding numerical values.

The code sandbox below has an expression for plotting out  $2^{t/h}$ for $-4 \leq t \leq 12$ where $h = 4$. It also plots out $e^{kt}$ and $10^{t/d}$

```{r EB2-1, exercise=TRUE, exercise.cap="Sandbox", warning=FALSE, exercise.setup="daily-digital-04-sandbox3-setup", exercise.eval=FALSE, exercise.lines=8, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE, eval=FALSE}
fa <- makeFun(2^(t/h) ~ t, h = 4)
fb <- makeFun(10^(t/d) ~ t, d = 10)
fc <- makeFun(exp(k*t)  ~ t, k = 0.1)

slice_plot(fa(t) ~ t, domain(t = c(-4, 12))) %>%
  slice_plot(fb(t) ~ t, color="blue") %>%
  slice_plot(fc(t) ~ t, color  = "red") %>%
  gf_lims(y = c(0, 8))
```

Your task is to modify the values of `d` and `k` such that all three curves lie on top of one another. (Leave `h` at the value 4.) You can find the appropriate values of `d` and `k` to accomplish this by any means you like, say, by using the algebra of exponents or by using trial and error. (Trial and error is a perfectly valid strategy regardless of what your high-school math teachers might have said about "guess and check." The trick is to make each new guess systematically based on your previous ones and observation of how those previous ones performed.)

After you have found values of `k` and `d` that are suited to the task ...

```{r EB1-1, echo=FALSE}
etude2::etudeEssay(
  prompt = "Enter the numerical value of your best estimate of `k`."
)
```

```{r EB1-2, echo=FALSE}
etude2::etudeEssay(
  prompt = "Enter the numerical value of your beat estimate of `d`."
)
```
