---
title: "Daily Digital 03"
subtitle: "Math 142Z"
date: "Due 15 Jan 2021"
output: learnr::tutorial
runtime: shiny_prerendered
tutorial:
  version: 0.1
tags:
- vectors
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
```



```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("DD03") # Add the DD number
etude2::etude_document_name("DD-142Z-03") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
# Leave this safety check here
if (etude2::etude_document_name() == "DD-") 
  stop("You must set the master_id() and the document name specifically and uniquely for this document")
library(mosaic)
library(mosaicCalc)
```


## Objectives

a. Understand the notation of limits of integration within a definite integral 
b. Determine units of a definite integral (MMAC pg. 614) 
c. Use the algebraic properties of definite integrals (MMAC pg. 615-616) to calculate definite integrals 
d. Calculate definite integrals graphically 

## Accumulating change

```{r child="sailing-over-time.Rmd"}
```


## Power and energy

```{r child="solar-example.Rmd"}
```



## Algebraic Integral of the Day

```{r child="IOD-C.Rmd"}

```

## Careful with units!

```{r child="falcon-tell-mug.Rmd"}
```

## Hand in your work

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>

```{r child=system.file("submission-handler.Rmd", package="etude2")}

```


-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name("142Z-DD-03.Rmd")`

