---
title: "Mechanics of Iteration: Trajectories"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
library(math141Z)
library(etude2)
```

A **trajectory** is a sequence of successive states. For example, here is a trajectory of six successive states: $$\{ {\mathbf S_0}, {\mathbf S_1},  {\mathbf S_2}, {\mathbf S_3}, {\mathbf S_4}, {\mathbf S_5}\}$$

It's conventional to call ${\mathbf S}_0$ the "initial state" or, more often, the **initial condition**.

Usually, we calculate a trajectory one step at a time:
$$\left\{ {\mathbf S_0},\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \\ {\mathbf S_1}=g({\mathbf S_0}),\\ {\mathbf S_2} = g({\mathbf S_1}),\\ {\mathbf S_3} = g({\mathbf S_2}),\\ {\mathbf S_4} = g({\mathbf S_3}),\\ \ \ \ \ \ {\mathbf S_5} = g({\mathbf S_4}))\ \ \ \right\}$$
NOTE: We've written the above on several lines just to make it more readable. It could just as well have been written on one line.


```{r fdem2-1, echo=FALSE, results="markup"}
askMC(
  "Given the system ${\\mathbf S}_{n+1} = g( {\\mathbf S}_n)$ with $g(S) \\equiv 5 + S/2$, which of these is a trajectory of the system?",
  "+{16, 13, 11.5, 10.75, 10.375, 10.1875, 10.09375}+",
  "{16, 8, 4, 2, 1}",
  "{5, 7.5, 10, 12.5, 15}",
  "{0, 5, 10, 15, 20, 25}"
)
``` 

```{r fdem2-2, echo=FALSE, results="markup"}
askMC(
  "Given the system ${\\mathbf S}_{n+1} = g( {\\mathbf S}_n)$ with $g(S) \\equiv 5 + S/2$, which of these is **NOT** a trajectory of the system?",
  paste0("{",paste(Iterate(makeFun(5 + s/2 ~ s), x=0, 5)$s, collapse=", "),"}"),
  paste0("{",paste(Iterate(makeFun(5 + s/2 ~ s), x=1, 3)$s, collapse=", "),"}"),
  paste0("{",paste(Iterate(makeFun(5 + s/2 ~ s), x=-4, 4)$s, collapse=", "),"}"),
  paste0("{",paste(Iterate(makeFun(5 + s/2 ~ s), x=10, 5)$s, collapse=", "),"}"),
  "+{4, 7, 10, 13}+"
)
```  

