---
title: "Mechanics of Iteration"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
library(etude2)
```

One common mathematical task in using finite-difference equations is **iteration**. In iterating a finite-difference equation, you are running a machine to construct the "future state" from the "present state." In iteration, you repeatedly do this. By plugging in the "future state" into the machine, you generate a state further in the future. 

It can be helpful to think of "future" and "present" state in terms of a series of days. The system's state **today** is the present state. The system's state **tomorrow** is the next future state.  The system's state on the **day after tomorrow** is the future state after tomorrows's state. 

A finite-difference equation is a statement about how tomorrow's state is a function of today's state. We could write this in the form
$$\mbox{state}_{\small\mbox{tomorrow}} = g( \mbox{state}_{\small\mbox{ today}})$$
This notation is wordy and hard to read. So we simplify it. 

- Today is day $n$
- Tomorrow is day $n+1$
- The state on day $n$ is written $\mathbf S_n$. 
- The state on day $n+1$ is written $\mathbf S_{n+1}$
- The dynamics are represented by a function, say, $g({\mathbf S})$.
- To remind us of the role played by $g()$, we write $${\mathbf S}_{n+1} = g({\mathbf S}_n)$$

```{r fdem1, echo=FALSE, results="markup"}
etude2::etudeQ(
  "If today is $n$, what is the day after tomorrow?",
  "$n-1$" = "$n-1$ is yesterday.",
  "$n$" = "$n$ is today.",
  "$n+1$" = "$n+1$ is tomorrow.",
  "+$n+2$+",
  "$n=3$",
  random_answer_order = FALSE
)
```

```{r fdem2, echo=FALSE, results="markup"}
etude2::etudeQ(
  "If today is $n$, what is the day one week from now?",
  "$7n$",
  "$n + 1 week$" = "Although $n$ stands for \"today\", $n$ is a pure number.",
  "+$n+7$+"
)
```

```{r fdem3, echo=FALSE, results="markup"}
etude2::etudeQ(
  "If today is $n$, what is the day one week ago?",
  "$-7 n$",
  "$n - 1$ week" = "Although $n$ stands for \"today\", $n$ is a pure number.",
  "+$n-7$+"
)
```

```{r fdem4, echo=FALSE, results="markup"}
etude2::etudeQ(
"Given ${\\mathbf S}_{n+1} = g({\\mathbf S}_n)$, what is ${\\mathbf S}_{n+3}$ ?",
  "+$g(g(g({\\mathbf S}_n)))$+",
  "$3 g({\\mathbf S}_n)$" = "$3 g({\\mathbf S}_n)$ is \"three times ${\\mathbf S}_{n+1}$\", not ${\\mathbf S}_{n+3}$.",
  "$g({\\mathbf S}_n)^3$",
  "$g({\\mathbf S}_n) + 2 {\\mathbf S}_{n}$"
)
```

```{r fdem5, echo=FALSE, results="markup"}
etude2::etudeQ(
"Given ${\\mathbf S}_{n+1} = g({\\mathbf S}_n)$, which of these is **not** ${\\mathbf S}_{n+1}$ ?",
  "$g(g(g({\\mathbf S}_{n-2})))$",
  "+$g({\\mathbf S}_{n-1})$+" = "This is ${\\mathbf S}_{n}$.",
  "$g(g({\\mathbf S}_{n-1}))$",
  "$g(g(g(g(g(g(g(g({\\mathbf S}_{n-7}))))))))$" = "These 8 applications of $g()$ will take ${\\mathbf S}_{n-7}$ up to ${\\mathbf S}_{n+1}$"
)
```

