---
title: "Some Block 1 Review"
subtitle: "Math 142Z"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r mother1-setup, include=FALSE}
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(DT)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


```{r context="setup", include=FALSE}
### Boilerplate: Packages relating to `{learnr}` and 
### submission functionality.
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(learnrhash)
```


```{r context="setup", include=FALSE}
# Good practice to include these in any document bringing
# with the answer submission system
etude2::set_master_id("Block-1") # Add the DD number
etude2::etude_document_name("Review-Block-1") # Add the DD number
# Fill this data frame in when document is completed. See etude2 documentation
document_roster <- tibble::tribble(
  ~ id,
)
```

```{r eval=FALSE}
# Leave this safety check here
if (etude2::etude_document_name() == "DD-") 
  stop("You must set the master_id() and the document name specifically and uniquely for this document")
```

## Algebra-free calculus?

These problems deal with important aspects of calculus, but don't use either algebra or computing. Instead, they are about ideas that can help you understand the concepts of calculus (e.g., "signed area", dimension change) and provide insight into calculus operations in science and engineering.

## Boxing areas

### Problem 1

```{r child="beech-sleep-drawer.Rmd"}
```

### Problem 2

```{r child="beech-sleep-drawer2.Rmd"}
```

## Seeing slope, height, area in the same graph


```{r child="fox-make-stove.Rmd"}
```

## Jumping between dimensions

```{r child="goat-pitch-socks.Rmd"}
```

## Definite and indefinite integrals

[This is a repeat of a problem from DD-08]

```{r child="fundamental-theorem-duplicate.Rmd"}
```



<!--

## Hand in your work

REMEMBER: Press the REFRESH button before copying the hash code and submitting it at the <a href="https://forms.microsoft.com/Pages/ResponsePage.aspx?id=Bgq4einwwEWE0X2tGc48YcCKijCKVrhDla481YPh1UJUMVdERUtVT0VOSEpMMEdPQU1aWlpYTTA0RS4u" target = "_blank">USAFA Math 142Z site</a>


```{r echo=FALSE, messages=FALSE}
etude2::submissions_ui()
```

```{r context="server", echo=FALSE, messages=FALSE}
etude2::submissions_logic()
```

-->

-----

Location: `r gsub("^.*USAFA/", "", knitr::current_input(dir = TRUE))` `r  etude2::etude_document_name()`

