---
title: "Counting squares 1"
author: "Daniel Kaplan"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)

knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


At a glance, give a rough numerical value for the integral of the function shown in the plot.

```{r echo = FALSE}
xpts = c(10,21,33,45,68,82,93)
ypts = c(4,6,7,2,3.5,5.5,6.5) - 3
f = splinefun(xpts,ypts)
g1 = function(x){f(x) + .3*sin(2*pi*x/9)}
newx = seq(10,93,by=10)
set.seed(65)
g2 = approxfun(newx,f(newx))
g3 = splinefun( newx, f(newx)+rnorm(length(newx),mean=0,sd=.3) )
g4 = splinefun( newx, f(newx)+rnorm(length(newx),mean=2.3,sd=.3) )
slice_plot(g3(x) ~ x, domain(x=c(0,100)), color = "gray", size=3 ) 
G <- antiD(g3(x) ~ x)
G(100) - G(0)
```

```{r bsd1, echo=FALSE}
askMC(
  "The integral of the function from x=0 to x = 100 is roughly ...",
  -150, -70, -20, -10, -2, -1,  0, 1, 2, 10, 20, 70, "+150+",
  random_answer_order = FALSE
)
```

**Comments**

1. Integration is a painstaking elaboration on multiplication. To get the areas of each "box" framed by the coordinate lines, you multiply the horizontal and vertical extent of the box. In general, these quantities have dimension, for instance, "flow" (L^3 / T) and "time" (T). The "area" has corresponding  dimension multiplication, for instance "flow $\times$ time = volume" (L^3).

2. The "area" metaphor for integration breaks down and needs to be replaced with "signed area." You're counting the boxes between the function and the x-axis. When a box is below the axis it counts as negative area.
