---
title: "Counting squares 2"
author: "Daniel Kaplan"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)

knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)
```


At a glance, give a rough numerical value for the integral of the function shown in the plot.

```{r echo = FALSE}
xpts = c(10,21,33,45,68,82,93)
ypts = c(4,6,7,2,3.5,5.5,6.5) - 3
f = splinefun(xpts,ypts)
g1 = function(x){f(x) + .3*sin(2*pi*x/9)}
newx = seq(10,93,by=10)
set.seed(65)
g2 = approxfun(newx,f(newx))
g3 = splinefun( newx, f(newx)+rnorm(length(newx),mean=0,sd=.3) )
g4 = splinefun( newx, f(newx)+rnorm(length(newx),mean=2.3,sd=.3) )
slice_plot(g1(x) ~ x, domain(x=c(0,100)), color = "gray", size=3 ) 
G1 <- antiD(g1(x) ~ x)
G1(100) - G1(0)
```

```{r bsd2, echo=FALSE}
etude2::etudeQ(
  "The integral of the function from x=0 to x = 100 is roughly ...",
  -150, -70, -20, -10, -2, -1,  0, 1, 2, 10, 20, "70", "+150+",
  random_answer_order = FALSE
)
```

Comments:

1. This function has both high-frequency ups and downs on the scale of about 10 x-units, as well as low frequency components on a scale of about 50 x-units. The "area" doesn't depend very much on the high-frequency components because the quick ups and downs cancel each other out.

2. OK, a little bit of algebra ... This function is a linear combination of high-frequency stuff which we might think about as $\cos(\omega_{big} t)$ and low-frequency stuff like $\cos(\omega_{small} t)$. Each of these is one of our basic modeling functions, so you should feel good about the following statement:
$$\int \cos(\omega t) dt = \frac{1}{\omega}\sin(\omega t)$$
Notice that when $\omega$ is large, the quantity $1 / \omega$ is small. Which is pretty much the same thing as saying "high-frequency oscillations have an integral that is small compared to low-frequency oscillations."
